---
title: "Solutions: Why Math in Psychology?"
subtitle: "New Handbook of Mathematical Psychology Vol.3 - Preface"
---

## Conceptual Questions

**A1.** The core message is that mathematics in psychology is primarily about **conceptual clarity** and **avoiding conceptual confusions**, not about computation. Before we can compute anything, we need to explicate our hidden assumptions. Often, when we make these assumptions explicit, we find they are not compelling, or we discover alternative assumptions that lead to different conclusions. The value of mathematical thinking lies in forcing us to state our assumptions clearly.

**A2.** Hidden assumptions in the Dupin example:

-   D1: Motives A and B can be combined in a well-defined sense
-   D2: Each motive can be assigned a real number value (a and b)
-   D3: The combination of A and B can also be assigned a real number value (c)
-   D4: The combination is represented by the **sum** of values (a + b)
-   D5: We can empirically observe that c ≠ a + b

The key hidden assumption is D4. If we don't assume combination = sum, then c doesn't have to equal a + b. It could be ab, max(a,b), or something else entirely.

**A3.**

| Aspect | Luce | Rozeboom |
|------------------------|------------------|------------------------------|
| Dimensional constants | Not allowed | Allowed |
| Conclusion | Only power functions ψ(x) = bx\^β satisfy scale invariance | ANY function of x/x₀ satisfies scale invariance |
| Implication | Strong constraint on psychophysical laws | No constraint - Fechner's log law is valid |

The key difference: Rozeboom showed that Luce's "principle" contained a hidden assumption (no dimensional constants). Once dimensional constants like x₀ are allowed, the restriction to power functions disappears.

**A4.** b) Ordinal scale

**A5.** b) f(x) = k · g(x) for some positive constant k

**A6.** c) No dimensional constants are allowed in the formula

## Mathematical Exercises

**A7.** Starting with ψ(kx) = K(k) · ψ(x) and ψ(x) = x\^β:

Left side: ψ(kx) = (kx)\^β = k\^β · x\^β

Right side: K(k) · ψ(x) = K(k) · x\^β

Setting them equal: k\^β · x\^β = K(k) · x\^β

Dividing both sides by x\^β: k\^β = K(k)

Therefore: **K(k) = k\^β**

**A8.** For ψ(x) = log(x), we compute K(k) = ψ(kx)/ψ(x) = log(kx)/log(x):

-   For x = 2, k = 3: K(3) = log(6)/log(2) = 1.79/0.69 ≈ 2.58
-   For x = 4, k = 3: K(3) = log(12)/log(4) = 2.48/1.39 ≈ 1.79
-   For x = 8, k = 3: K(3) = log(24)/log(8) = 3.18/2.08 ≈ 1.53

K(k) gives **different values** for different x! This violates Luce's requirement that K(k) depend only on k, not on x. Therefore, the log function does NOT satisfy Luce's principle.

**A9.** With dimensional constants, we have s = s₀ · ψ(x/x₀).

When we rescale: x → kx and x₀ → kx₀ (same units, so both scale by k)

The ratio becomes: (kx)/(kx₀) = x/x₀

The k cancels out! The argument to ψ is unchanged.

Therefore s = s₀ · ψ(x/x₀) is automatically scale-invariant for ANY function ψ, because the ratio x/x₀ is dimensionless and does not change when we rescale units.

## Coding Exercises

**A10.**

```{r}
k <- 3
x_values <- c(2, 4, 8)

for (beta in c(0.5, 1, 2, 3)) {
  K_values <- sapply(x_values, function(x) (k * x)^beta / x^beta)
  cat("beta =", beta, ": K(k) =", unique(round(K_values, 6)), "\n")
}
```

Output shows K(k) = k\^β is constant across all x values for each β.

**A11.**

```{r}
k <- 3
x_values <- c(2, 4, 8)

K_values <- sapply(x_values, function(x) log(k * x) / log(x))

data.frame(
  x = x_values,
  K_k = round(K_values, 3)
)
```

Output shows K(k) = 2.585, 1.792, 1.528 - different for each x! Log fails Luce's principle.

**A12.**

```{r}
x <- 10
x0 <- 2
k <- 5

original_ratio <- x / x0
scaled_ratio <- (k * x) / (k * x0)

data.frame(
  original_ratio = original_ratio,
  scaled_ratio = scaled_ratio,
  equal = original_ratio == scaled_ratio
)
```

Output: Both ratios equal 5. The k cancels, proving scale invariance.

## Applied Problems

**A13.**

a)  Yes, ψ(x) = √x = x\^0.5 is a power function, so it satisfies Luce's principle.

b)  K(k) = k\^0.5 = √k

Verification: ψ(kx) = √(kx) = √k · √x = k\^0.5 · ψ(x) ✓

c)  Yes, s = s₀ · √(x/x₀) satisfies scale invariance because x/x₀ is dimensionless. When we rescale, (kx)/(kx₀) = x/x₀, so the ratio is unchanged.

**A14.**

a)  The hidden assumption is that "A presented first, B presented second" is represented simply as A ∧ B. This ignores the temporal/positional information.

b)  Positional labeling solves this by including position as part of the proposition:

-   A₁ = "A presented first"
-   B₂ = "B presented second"

Now A₁ ∧ B₂ and A₂ ∧ B₁ are **different propositions**. Commutativity is preserved: A₁ ∧ B₂ ≡ B₂ ∧ A₁.

c)  Examples from psychology:

-   **Priming effects**: Seeing word A before word B affects recognition differently than B before A
-   **Serial position effects in memory**: Items presented first (primacy) vs. last (recency) are remembered differently
-   **Anchoring effects**: The first number presented affects subsequent numerical judgments

In each case, the underlying mathematical structure (e.g., memory strength, response probability) is preserved when we properly encode temporal position as part of the stimulus description.
