---
title: "Chapter 1: Initial Visual Encoding (Part 2)"
subtitle: "Perceptual Inference: From Cones to Perception"
author: "Wandell & Brainard - New Handbook of Mathematical Psychology Vol.3"
format:
  revealjs:
    theme: simple
    slide-number: true
    smaller: true
---

# Section 1.5: Perceptual Inference

## The Problem: Ambiguity

**Part 1 showed:** Cone signals are ambiguous — many scenes produce the same response.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

# Many scenes -> same cone response
set.seed(42)
scenes <- c("Yellow\nbanana", "Red+Green\nmix", "Orange\nsunset")
response <- c(0.7, 0.5, 0.1)

# Scene 1
barplot(c(0.7, 0.5, 0.1), names.arg = c("L", "M", "S"), col = c("red", "green", "blue"),
        main = scenes[1], ylim = c(0, 1))

# Scene 2
barplot(c(0.7, 0.5, 0.1), names.arg = c("L", "M", "S"), col = c("red", "green", "blue"),
        main = scenes[2], ylim = c(0, 1))

# Scene 3
barplot(c(0.7, 0.5, 0.1), names.arg = c("L", "M", "S"), col = c("red", "green", "blue"),
        main = scenes[3], ylim = c(0, 1))
```

**Same L, M, S response!** Which scene is it?

. . .

> "The brain decodes the sensory representation to produce perceptions that are a likely guess about the state of the external world. Perception is an unconscious inference." — Helmholtz

## The Bayesian Solution

**x** = scene (what we want to know)

**y** = cone excitations (what we measure)

$$p(\mathbf{x}|\mathbf{y}) = K \cdot p(\mathbf{y}|\mathbf{x}) \cdot p(\mathbf{x})$$

. . .

| Term    | Name           | Meaning                                           |
|---------|----------------|---------------------------------------------------|
| p(y\|x) | **Likelihood** | "If scene is x, how likely is cone response y?"   |
| p(x)    | **Prior**      | "How common is scene x in the world?"             |
| p(x\|y) | **Posterior**  | "Given cones say y, what scene x is most likely?" |

. . .

**Key insight:** The prior resolves ambiguity — it tells us which of the many possible scenes is most plausible!

## Figure 1.9: Setup — A Tiny 2-Pixel World

**Simplify to understand:** Imagine a scene with only 2 pixels

-   **x₁** = intensity of pixel 1 (0 to 1)
-   **x₂** = intensity of pixel 2 (0 to 1)
-   **y** = what the cone measures (a weighted sum)

```{r}
#| echo: true
# Create a grid of all possible (x1, x2) combinations
x1 <- seq(0, 1, length.out = 60)  # 60 values from 0 to 1
x2 <- seq(0, 1, length.out = 60)  # 60 values from 0 to 1

# expand.grid creates all combinations: 60 × 60 = 3600 points
grid <- expand.grid(x1 = x1, x2 = x2)
```

Each row = one possible scene (x₁, x₂)

## Step 1: The Prior p(x) — What Scenes Are Likely?

**Prior belief:** Neighboring pixels tend to be similar (natural images are smooth)

We model this with a **2D Gaussian** with positive correlation:

```{r}
#| echo: true
library(mvtnorm)  # for multivariate normal

# Mean: expect both pixels around 0.5 (middle gray)
mu_prior <- c(0.5, 0.5)

# Covariance matrix: pixels are correlated!
sigma_prior <- matrix(c(0.025, 0.020,   # variance of x1, covariance
                        0.020, 0.025),  # covariance, variance of x2
                      nrow = 2, ncol = 2)
sigma_prior
```

**Positive covariance (0.020)** means: if x₁ is high, x₂ is probably high too!

::: callout-note
## We Choose These Parameters!

The prior parameters (μ, Σ) encode our **beliefs about natural scenes**:

-   Learned from datasets of natural images
-   Based on domain knowledge (e.g., "surfaces are smooth")
-   Can be refined with more data
:::

## Step 1a: Understanding the Multivariate Gaussian

**The formula:** For a 2D Gaussian, the probability density at point **x** = (x₁, x₂) is:

$$p(\mathbf{x}) = \frac{1}{2\pi|\Sigma|^{1/2}} \exp\left(-\frac{1}{2}(\mathbf{x}-\boldsymbol{\mu})^T \Sigma^{-1} (\mathbf{x}-\boldsymbol{\mu})\right)$$

. . .

**Our parameters:**

| Parameter                | Value      | Meaning                                |
|--------------------------|------------|----------------------------------------|
| **μ** (mean)             | (0.5, 0.5) | Most likely pixel values = middle gray |
| **σ₁²** (variance of x₁) | 0.025      | How much pixel 1 varies                |
| **σ₂²** (variance of x₂) | 0.025      | How much pixel 2 varies                |
| **σ₁₂** (covariance)     | 0.020      | How pixels vary *together*             |

. . .

**The covariance matrix Σ:**

$$\Sigma = \begin{pmatrix} \sigma_1^2 & \sigma_{12} \\ \sigma_{12} & \sigma_2^2 \end{pmatrix} = \begin{pmatrix} 0.025 & 0.020 \\ 0.020 & 0.025 \end{pmatrix}$$

**Positive σ₁₂** → pixels are **positively correlated** → the Gaussian is stretched along the diagonal!

## Step 1a (cont.): The `dmvnorm()` Function in R

**Package:** `mvtnorm` (multivariate normal)

``` r
dmvnorm(x, mean, sigma)
```

| Argument | Type             | Description                               |
|----------|------------------|-------------------------------------------|
| `x`      | vector or matrix | Point(s) to evaluate, e.g., `c(0.3, 0.7)` |
| `mean`   | vector           | Mean vector **μ**, e.g., `c(0.5, 0.5)`    |
| `sigma`  | matrix           | Covariance matrix **Σ** (2×2 for 2D)      |

. . .

**Example:**

```{r}
#| echo: true
library(mvtnorm)

# What is the probability density at point (0.5, 0.5)?
dmvnorm(x = c(0.5, 0.5), mean = c(0.5, 0.5),
        sigma = matrix(c(0.025, 0.020, 0.020, 0.025), 2, 2))

# What about at (0.9, 0.1)? (very different pixels - unlikely!)
dmvnorm(x = c(0.9, 0.1), mean = c(0.5, 0.5),
        sigma = matrix(c(0.025, 0.020, 0.020, 0.025), 2, 2))
```

The smooth scene (0.5, 0.5) is **much more likely** than the jagged scene (0.9, 0.1)!

## Step 1b: Calculate Prior for Each Scene

```{r}
#| echo: true
# dmvnorm() = density of multivariate normal
# For each (x1, x2) pair, how likely is that scene?

# Calculate prior probability for EACH of the 3600 grid points
prior_values <- numeric(nrow(grid))  # empty vector

for (i in 1:nrow(grid)) {
  point <- c(grid$x1[i], grid$x2[i])  # one (x1, x2) pair
  prior_values[i] <- dmvnorm(point, mean = mu_prior, sigma = sigma_prior)
}

# Reshape into 60×60 matrix for plotting
prior <- matrix(prior_values, nrow = length(x1), ncol = length(x2))
```

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4.5
blue_cols <- colorRampPalette(c("white", "lightblue", "steelblue", "navy"))(30)
image(x1, x2, prior, col = blue_cols, xlab = "Pixel 1 (x1)", ylab = "Pixel 2 (x2)", main = "Prior p(x): Smooth scenes are likely")
contour(x1, x2, prior, add = TRUE, col = "navy", lwd = 1.5, drawlabels = FALSE, nlevels = 8)
abline(a = 0, b = 1, lty = 2, lwd = 2, col = "black")
text(0.75, 0.85, "x1 ≈ x2\n(smooth)", cex = 0.9)
```

## Step 2: The Likelihood p(y\|x) — What Did the Cone See?

**The cone measures a weighted sum** of the two pixels:

$$y = 0.5 \cdot x_1 + 0.5 \cdot x_2 + \text{noise}$$

Suppose the cone measured **y = 0.25**. Which scenes could produce this?

```{r}
#| echo: true
# Any (x1, x2) where 0.5*x1 + 0.5*x2 ≈ 0.25 is consistent!
# That means: x1 + x2 ≈ 0.5

# Likelihood = probability of observing y=0.25 given scene (x1, x2)
y_observed <- 0.25
noise_sd <- 0.025  # measurement noise

likelihood_values <- numeric(nrow(grid))

for (i in 1:nrow(grid)) {
  # What would cone measure for this scene?
  y_predicted <- 0.5 * grid$x1[i] + 0.5 * grid$x2[i]

  # How likely is y_observed given y_predicted?
  likelihood_values[i] <- dnorm(y_observed, mean = y_predicted, sd = noise_sd)
}

likelihood <- matrix(likelihood_values, nrow = length(x1), ncol = length(x2))
```

## Step 2b: Visualize the Likelihood

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 5
orange_cols <- colorRampPalette(c("white", "lightyellow", "orange", "red3"))(30)
image(x1, x2, likelihood, col = orange_cols, xlab = "Pixel 1 (x1)", ylab = "Pixel 2 (x2)",
      main = "Likelihood p(y|x): Which scenes explain y=0.25?")
contour(x1, x2, likelihood, add = TRUE, col = "red4", lwd = 1.5, drawlabels = FALSE, nlevels = 8)
abline(a = 0.5, b = -1, lty = 2, lwd = 2, col = "black")
text(0.15, 0.25, "x1 + x2 = 0.5", cex = 1.1, srt = -45)

# Show some example points on the ridge x1 + x2 = 0.5
points(0, 0.5, pch = 19, cex = 1.5, col = "darkred")
points(0.25, 0.25, pch = 19, cex = 1.5, col = "darkred")
points(0.5, 0, pch = 19, cex = 1.5, col = "darkred")
```

**The ridge:** All scenes where x₁ + x₂ = 0.5 are equally likely given the cone response!

-   (0, 0.5), (0.25, 0.25), (0.5, 0) all produce y = 0.25

## Step 3: Posterior = Prior × Likelihood

**Bayes' rule:** Multiply prior and likelihood, then normalize

```{r}
#| echo: true
# Element-wise multiplication!
posterior <- prior * likelihood

# Normalize so it sums to 1 (optional, for proper probability)
posterior <- posterior / sum(posterior)
```

**What happens geometrically?**

-   Prior says: "x₁ ≈ x₂" (diagonal blob centered at 0.5, 0.5)
-   Likelihood says: "x₁ + x₂ = 0.5" (ridge)
-   **Intersection:** Where both are satisfied → (0.25, 0.25)!

## Step 3b: The Full Picture

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4

# Recalculate for clean visualization
prior <- matrix(dmvnorm(grid, mu_prior, sigma_prior), length(x1), length(x2))
likelihood <- matrix(dnorm(0.5*grid$x1 + 0.5*grid$x2, mean = 0.25, sd = 0.025), length(x1), length(x2))
posterior <- prior * likelihood

blue_cols <- colorRampPalette(c("white", "lightblue", "steelblue", "navy"))(30)
orange_cols <- colorRampPalette(c("white", "lightyellow", "orange", "red3"))(30)
green_cols <- colorRampPalette(c("white", "lightgreen", "green3", "darkgreen"))(30)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

# Prior
image(x1, x2, prior, col = blue_cols, xlab = "Pixel 1", ylab = "Pixel 2", main = "Prior p(x)")
contour(x1, x2, prior, add = TRUE, col = "navy", lwd = 1.5, drawlabels = FALSE, nlevels = 8)

# Likelihood
image(x1, x2, likelihood, col = orange_cols, xlab = "Pixel 1", ylab = "Pixel 2", main = "Likelihood p(y|x)")
contour(x1, x2, likelihood, add = TRUE, col = "red4", lwd = 1.5, drawlabels = FALSE, nlevels = 8)
abline(a = 0.5, b = -1, lty = 2, lwd = 2, col = "black")

# Posterior
image(x1, x2, posterior, col = green_cols, xlab = "Pixel 1", ylab = "Pixel 2", main = "Posterior p(x|y)")
contour(x1, x2, posterior, add = TRUE, col = "darkgreen", lwd = 1.5, drawlabels = FALSE, nlevels = 8)
points(0.25, 0.25, pch = 4, cex = 2, lwd = 3, col = "red")
```

|   | Prior | Likelihood | Posterior |
|----|----|----|----|
| **Shape** | Diagonal blob | Ridge (x₁+x₂=0.5) | Narrow peak |
| **Says** | "Pixels are similar" | "Many scenes fit y" | "Best guess: (0.25, 0.25)" |

## What Does the Prior Encode?

The prior p(x) captures **statistical regularities of natural scenes**.

But where do these come from?

. . .

**Two sources:**

1.  **Low-level statistics** — measurable patterns in natural images
    -   Spatial correlations (nearby pixels similar)
    -   Spectral correlations (smooth reflectances)
2.  **High-level knowledge** — learned from experience
    -   Objects have typical colors
    -   Light usually comes from above
    -   Surfaces are usually smooth

Let's look at each...

## Prior Knowledge: Spatial Correlations

**Observation:** In natural images, nearby pixels tend to have similar intensities.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Spatial correlation
set.seed(111)
x <- 1:100

# Smooth natural image (random walk)
natural <- cumsum(rnorm(100, 0, 0.1))
natural <- (natural - min(natural)) / (max(natural) - min(natural))

# Random noise
random <- runif(100)

# Natural image
plot(x, natural, type = "l", lwd = 3, col = "darkgreen",
     main = "Natural image (1D slice)", xlab = "Position (pixels)", ylab = "Intensity",
     ylim = c(0, 1))

# Random noise
plot(x, random, type = "l", lwd = 2, col = "gray40",
     main = "Random noise", xlab = "Position (pixels)", ylab = "Intensity",
     ylim = c(0, 1))
```

**Left:** Smooth transitions — this is what real scenes look like!

**Right:** Random jumps — extremely unlikely in nature.

→ The prior assigns **high probability to smooth images**, low probability to noisy ones.

## Prior Knowledge: Spectral Correlations

**Observation:** Natural surface reflectances are spectrally smooth.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

wavelengths <- seq(400, 700, by = 5)

# Smooth natural reflectance (like a leaf or skin)
smooth_spectrum <- 0.3 + 0.4 * exp(-0.5 * ((wavelengths - 550) / 80)^2)

# Random reflectance
set.seed(456)
noisy_spectrum <- runif(length(wavelengths), 0.1, 0.9)

plot(wavelengths, smooth_spectrum, type = "l", lwd = 3, col = "orange",
     main = "Natural reflectance (e.g., leaf)", xlab = "Wavelength (nm)", ylab = "Reflectance",
     ylim = c(0, 1))

plot(wavelengths, noisy_spectrum, type = "l", lwd = 2, col = "gray40",
     main = "Random reflectance", xlab = "Wavelength (nm)", ylab = "Reflectance",
     ylim = c(0, 1))
```

**Left:** Smooth curve — typical of natural materials (leaves, skin, wood).

**Right:** Jagged spectrum — almost never occurs in nature.

→ The prior says: **"Reflectances that jump wildly between wavelengths are unlikely."**

## Prior Knowledge: High-Level Beliefs

Beyond statistics, we have **learned knowledge** about the world:

| Prior belief | Why it helps |
|----|----|
| Bananas are usually yellow | Resolve color ambiguity under different lighting |
| Light comes from above | Interpret shading as 3D shape |
| Surfaces are smooth, not spiky | Fill in missing texture information |
| Objects are solid, continuous | Group fragments into whole objects |

. . .

::: callout-tip
## The Brain as a Bayesian Machine

The brain learns these priors through **experience** — millions of visual scenes over a lifetime.

When cone signals are ambiguous, the prior "votes" for the most plausible interpretation!
:::

## Ideal Observer Theory: The Key Question

We've built a forward model: Scene → Optics → Cones → Signals

Now the big question:

. . .

> **How well COULD a visual system perform, given only the information in cone signals?**

. . .

This is important because:

-   It sets an **upper bound** on performance
-   It tells us what's **physically possible**
-   It helps us understand **where limitations come from**

## What is an Ideal Observer?

An **ideal observer** is a hypothetical system that:

1.  **Knows the forward model perfectly** — exactly how scenes become cone signals
2.  **Knows the prior perfectly** — the true statistics of natural scenes
3.  **Uses optimal Bayesian inference** — combines likelihood and prior perfectly
4.  **Has no additional limitations** — no neural noise, no memory limits, no attention lapses

. . .

In other words: **the best possible performance given the information available.**

::: callout-note
## Think of it as...

The ideal observer is like a "perfect brain" attached to real eyes. It extracts ALL the information that the cones provide — nothing more, nothing less.
:::

## Why Compare Humans to Ideal Observers?

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3.5

par(mar = c(1, 1, 2, 1))
plot.new()

# Draw comparison diagram
rect(0.05, 0.5, 0.35, 0.9, col = "lightblue", border = "navy")
text(0.2, 0.7, "Human\nPerformance", cex = 1.2, font = 2)

rect(0.65, 0.5, 0.95, 0.9, col = "lightgreen", border = "darkgreen")
text(0.8, 0.7, "Ideal Observer\nPerformance", cex = 1.2, font = 2)

arrows(0.37, 0.7, 0.63, 0.7, lwd = 3, code = 3)
text(0.5, 0.78, "Compare", cex = 1.1, font = 2)

# Outcomes
text(0.5, 0.35, "What the comparison tells us:", cex = 1.1, font = 2)
text(0.08, 0.18, "• Human = Ideal → Limited by early encoding (optics, cones)", adj = 0, cex = 1)
text(0.08, 0.05, "• Human < Ideal → Additional factors limit performance", adj = 0, cex = 1)
```

**The gap between human and ideal tells us where to look for limitations!**

## Example: Contrast Detection Task

**Task:** Detect a faint striped pattern (grating) on a gray background.

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.5

par(mfrow = c(1, 3), mar = c(2, 2, 3, 1))

# Create grayscale palette (black to white)
gray_pal <- gray(seq(0, 1, length.out = 256))

# High contrast (50%) - easy to see
x <- seq(0, 6*pi, length.out = 100)
y <- seq(0, 6*pi, length.out = 100)
z_high <- outer(x, y, function(x, y) 0.5 + 0.5 * sin(x))
image(z_high, col = gray_pal, axes = FALSE, main = "High contrast: 50%", zlim = c(0, 1))
box()

# Medium contrast (10%) - harder to see
z_med <- outer(x, y, function(x, y) 0.5 + 0.10 * sin(x))
image(z_med, col = gray_pal, axes = FALSE, main = "Medium contrast: 10%", zlim = c(0, 1))
box()

# Low contrast (2%) - at threshold
z_low <- outer(x, y, function(x, y) 0.5 + 0.02 * sin(x))
image(z_low, col = gray_pal, axes = FALSE, main = "Low contrast: 2% (threshold)", zlim = c(0, 1))
box()
```

Can you see the stripes in all three? The rightmost is near human detection threshold!

**Threshold** = minimum contrast to discriminate grating from uniform field (e.g., 80% correct)

**Sensitivity** = 1 / threshold (higher = better)

## What is Spatial Frequency?

**Spatial frequency** = how many stripe cycles fit in one degree of visual angle.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3

par(mfrow = c(1, 4), mar = c(2, 1, 3, 1))

gray_pal <- gray(seq(0, 1, length.out = 256))

# Different spatial frequencies
for (sf in c(1, 4, 8, 16)) {
  x <- seq(0, 2*pi*sf, length.out = 200)
  y <- seq(0, 2*pi, length.out = 50)
  z <- outer(x, y, function(x, y) 0.5 + 0.4 * sin(x))
  image(z, col = gray_pal, axes = FALSE, zlim = c(0, 1),
        main = paste0(sf, " cycle/degree"))
  box()
}
```

| Spatial frequency  | Stripe width   | What it represents            |
|--------------------|----------------|-------------------------------|
| Low (1-2 c/deg)    | Wide stripes   | Coarse details, large objects |
| Medium (4-8 c/deg) | Medium stripes | Typical object features       |
| High (16-32 c/deg) | Narrow stripes | Fine details, textures        |

**Question:** Are we equally sensitive to all spatial frequencies?

## The Contrast Sensitivity Function (CSF)

We measure **threshold** (minimum detectable contrast) at each spatial frequency.

Then compute **sensitivity = 1 / threshold**.

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 4

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Human CSF
sf <- c(0.5, 1, 2, 4, 8, 16, 32)
human_csf <- c(20, 80, 200, 300, 150, 30, 5)

plot(sf, human_csf, type = "b", pch = 19, lwd = 3, col = "black", log = "xy",
     xlab = "Spatial frequency (cycles/degree)", ylab = "Contrast sensitivity",
     main = "Human CSF", ylim = c(1, 500))
abline(v = 4, lty = 2, col = "red")
text(5, 400, "Peak ~4 c/deg", col = "red", cex = 0.9)

# Interpret the shape
plot.new()
text(0.5, 0.9, "Why this shape?", cex = 1.4, font = 2)
text(0.05, 0.72, "• Low frequencies: Limited by lateral inhibition", adj = 0, cex = 1.1)
text(0.05, 0.55, "• Peak at ~4 c/deg: Optimal for typical viewing", adj = 0, cex = 1.1)
text(0.05, 0.38, "• High frequencies: Limited by optics + cone spacing", adj = 0, cex = 1.1)
text(0.05, 0.15, "The CSF describes what we CAN and CANNOT see!", adj = 0, cex = 1.1, font = 2)
```

## Human vs Ideal Observer: The Gap

Now compare human CSF to the **ideal observer** (optimal Bayesian with perfect knowledge):

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Human CSF
sf <- c(0.5, 1, 2, 4, 8, 16, 32)
human_csf <- c(20, 80, 200, 300, 150, 30, 5)
ideal_csf <- c(800, 900, 950, 900, 600, 200, 50)

plot(sf, human_csf, type = "b", pch = 19, lwd = 2, col = "black", log = "xy",
     xlab = "Spatial frequency (cycles/degree)", ylab = "Contrast sensitivity",
     main = "Human vs Ideal Observer", ylim = c(1, 1000))
lines(sf, ideal_csf, type = "b", pch = 17, lwd = 2, col = "steelblue")
legend("topright", c("Human", "Ideal observer"), pch = c(19, 17),
       col = c("black", "steelblue"), lwd = 2, bg = "white")

# Shade the gap
polygon(c(sf, rev(sf)), c(human_csf, rev(ideal_csf)),
        col = rgb(1, 0, 0, 0.15), border = NA)

# Show the gap ratio
plot(sf, ideal_csf / human_csf, type = "b", pch = 15, lwd = 2, col = "red3", log = "x",
     xlab = "Spatial frequency (cycles/degree)", ylab = "Ideal / Human ratio",
     main = "The Gap: How much better is ideal?", ylim = c(0, 50))
abline(h = 1, lty = 2, col = "gray")
text(2, 5, "If ratio = 1, human is optimal", cex = 0.9, col = "gray30")
```

**Key finding:** Ideal observer is **much more sensitive** than humans!

→ Something beyond optics and cones is limiting human performance.

::: callout-note
## Note

Values shown are schematic/illustrative. See Banks, Geisler, & Bennett (1987) and Figure 1.10B in the chapter for actual data.
:::

## Computational Observers: The Problem

The ideal observer is **too good** — much more sensitive than humans.

**Why the gap?**

The ideal observer has "perfect knowledge":

-   Knows exactly how stimuli map to cone excitations
-   Knows the exact noise statistics
-   Uses the mathematically optimal decision rule

. . .

**But real humans:**

-   Don't have perfect templates
-   Must learn the decision rule from experience
-   Have additional sources of noise and variability
-   Eyes move during stimulus presentation

→ We need a more **realistic** model!

## What is a Computational Observer?

A **computational observer** relaxes the ideal observer assumptions one at a time:

| Factor | Ideal observer | Computational observer |
|----|----|----|
| Optics | Simplified PSF | Realistic wavefront aberrations |
| Decision rule | Optimal (known templates) | Learned from noisy examples |
| Eye position | Fixed | Fixational drift during stimulus |
| Cone response | Excitations only | Full photocurrent model with gain control |

. . .

**Key idea:** Each added realism brings the model **closer to human performance**.

If we can match human CSF with realistic factors → we understand what limits vision!

## Computational Observers: Closing the Gap

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.5

par(mar = c(4, 4, 3, 1))

sf <- c(5, 10, 20, 30, 40, 50)
human <- c(100, 50, 20, 10, 5, 2)
ideal <- c(800, 500, 200, 80, 30, 10)
realistic_optics <- c(600, 350, 130, 50, 18, 6)
learned_rule <- c(400, 220, 80, 30, 12, 4)
eye_movements <- c(250, 140, 55, 22, 9, 3)
photocurrent <- c(120, 60, 25, 12, 6, 2.5)

plot(sf, human, type = "b", pch = 19, lwd = 3, col = "black", log = "y",
     xlab = "Spatial frequency (c/deg)", ylab = "Contrast sensitivity",
     main = "Adding realism closes the gap (schematic)", ylim = c(1, 1000))
lines(sf, ideal, type = "b", pch = 17, lwd = 2, col = "gray60")
lines(sf, realistic_optics, type = "b", pch = 15, lwd = 2, col = "red3")
lines(sf, learned_rule, type = "b", pch = 16, lwd = 2, col = "blue")
lines(sf, eye_movements, type = "b", pch = 18, lwd = 2, col = "green4")
lines(sf, photocurrent, type = "b", pch = 8, lwd = 2, col = "purple")

legend("topright",
       c("Human", "Ideal", "+ Realistic optics", "+ Learned decision rule",
         "+ Fixational eye movements", "+ Photocurrent model"),
       pch = c(19, 17, 15, 16, 18, 8),
       col = c("black", "gray60", "red3", "blue", "green4", "purple"),
       lwd = 2, cex = 0.75, bg = "white")
```

## What Each Factor Does

| Factor added | What it models | Effect on sensitivity |
|----|----|----|
| **Realistic optics** | Wavefront aberrations (Thibos et al., 2002) | Small reduction |
| **Learned decision rule** | Template matching trained on noisy examples | Moderate reduction |
| **Fixational eye movements** | Drift during stimulus — adds blur | Further reduction |
| **Photocurrent model** | Gain control + additional noise in cones | Brings close to human |

. . .

::: callout-tip
By the time we account for realistic optics, learned decisions, eye movements, and photocurrent dynamics, the computational observer **matches human performance** at high spatial frequencies!

This tells us these factors — not mysterious "neural inefficiency" — explain the gap.
:::

## Image Reconstruction: The Setup

Beyond threshold tasks, can we use Bayesian inference to **reconstruct what we're looking at**?

**What we have:** Cone excitations **y** (noisy measurements)

**What we want:** The original image **x** (pixel values)

. . .

**The forward model:**

$$\bar{\mathbf{y}} = \mathbf{R}\mathbf{x}$$

| Symbol | Meaning                                     |
|--------|---------------------------------------------|
| **x**  | Image pixel values (what we want to find)   |
| **R**  | Matrix that maps pixels → cone excitations  |
| **ȳ**  | Mean cone excitations                       |
| **y**  | Actual cone excitations (ȳ + Poisson noise) |

The matrix **R** encodes optics, cone positions, and spectral sensitivities.

## Image Reconstruction: The Method

**Given:** Noisy cone excitations **y**

**Find:** Most likely image **x**

. . .

**Use Bayes' rule:**

$$p(\mathbf{x}|\mathbf{y}) \propto p(\mathbf{y}|\mathbf{x}) \cdot p(\mathbf{x})$$

| Component              | What we use                                      |
|------------------------|--------------------------------------------------|
| **Likelihood** p(y\|x) | Poisson noise model: y \~ Poisson(Rx)            |
| **Prior** p(x)         | Gaussian prior encoding natural image statistics |
| **Estimate** x̂         | Find x that maximizes the posterior              |

. . .

**Key insight:** The prior "fills in" information lost by sparse cone sampling!

## Image Reconstruction: Results at Different Eccentricities

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4

par(mfrow = c(1, 4), mar = c(2, 1, 3, 1))

# Simulate reconstruction quality at different eccentricities
set.seed(42)

# Original image (simple pattern)
n <- 30
original <- matrix(0, n, n)
# Add some structure
for (i in 1:n) {
  for (j in 1:n) {
    original[i, j] <- 0.5 + 0.3 * sin(i/3) * cos(j/3) + 0.1 * sin(i/2 + j/2)
  }
}

image(original, col = gray.colors(50), axes = FALSE, main = "Original image x")
box()

# 1 degree - good reconstruction
recon_1deg <- original + matrix(rnorm(n*n, 0, 0.05), n, n)
image(recon_1deg, col = gray.colors(50), axes = FALSE, main = "1° (fovea)")
box()

# 10 degrees - some blur
recon_10deg <- original
for (i in 2:(n-1)) {
  for (j in 2:(n-1)) {
    recon_10deg[i,j] <- mean(original[(i-1):(i+1), (j-1):(j+1)])
  }
}
recon_10deg <- recon_10deg + matrix(rnorm(n*n, 0, 0.1), n, n)
image(recon_10deg, col = gray.colors(50), axes = FALSE, main = "10° (mid-periphery)")
box()

# 18 degrees - very blurry
recon_18deg <- original
for (k in 1:3) {
  for (i in 2:(n-1)) {
    for (j in 2:(n-1)) {
      recon_18deg[i,j] <- mean(recon_18deg[(i-1):(i+1), (j-1):(j+1)])
    }
  }
}
recon_18deg <- recon_18deg + matrix(rnorm(n*n, 0, 0.15), n, n)
image(recon_18deg, col = gray.colors(50), axes = FALSE, main = "18° (far periphery)")
box()
```

| Eccentricity | Cone density | PSF blur | Reconstruction quality           |
|--------------|--------------|----------|----------------------------------|
| 1° (fovea)   | High         | Sharp    | Excellent                        |
| 10°          | Medium       | Medium   | Good (prior helps)               |
| 18°          | Low          | Blurry   | Recognizable (prior helps more!) |

## Why Does Reconstruction Work in the Periphery?

**The problem:** Fewer cones + blurrier optics → less information

**The solution:** The prior compensates!

. . .

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3

par(mar = c(1, 1, 2, 1))
plot.new()

# Likelihood column
rect(0.02, 0.3, 0.30, 0.9, col = "lightyellow", border = "orange3")
text(0.16, 0.75, "Likelihood\n(from cones)", cex = 1.1, font = 2)
text(0.16, 0.5, "Ambiguous:\nmany images\nfit the data", cex = 0.95)

# Plus sign
text(0.35, 0.6, "+", cex = 3, font = 2)

# Prior column
rect(0.40, 0.3, 0.68, 0.9, col = "lightblue", border = "navy")
text(0.54, 0.75, "Prior\n(natural images)", cex = 1.1, font = 2)
text(0.54, 0.5, "Constrains:\nsmooth, correlated\npixels likely", cex = 0.95)

# Equals sign
text(0.73, 0.6, "=", cex = 3, font = 2)

# Posterior column
rect(0.78, 0.3, 1.0, 0.9, col = "lightgreen", border = "darkgreen")
text(0.89, 0.75, "Posterior\n(reconstruction)", cex = 1.1, font = 2)
text(0.89, 0.5, "Best guess:\nplausible image\ngiven data", cex = 0.95)
```

**Even with massive information loss, the brain can "guess" reasonable images** because unlikely scenes (random noise, jagged edges) are downweighted by the prior!

## Why Reconstruction Matters

**The prior compensates for information loss:**

| Eccentricity | Cone density | PSF    | Reconstruction                   |
|--------------|--------------|--------|----------------------------------|
| 1° (fovea)   | High         | Sharp  | Excellent                        |
| 10°          | Medium       | Medium | Good (prior helps)               |
| 18°          | Low          | Blurry | Recognizable (prior helps more!) |

. . .

**Key insight:** Even with massive information loss in periphery, the brain can still "guess" reasonable images because:

1.  Natural images have structure (prior)
2.  Unlikely scenes are downweighted
3.  Ambiguity is resolved by experience

# Section 1.6: Summary

## The Forward Calculation (Part 1)

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3

par(mar = c(1, 1, 2, 1))
plot.new()

# Pipeline
rect(0.0, 0.4, 0.18, 0.8, col = "lightblue", border = "navy")
text(0.09, 0.6, "Scene\n(7D light field)", cex = 0.9, font = 2)

arrows(0.20, 0.6, 0.24, 0.6, lwd = 2)

rect(0.26, 0.4, 0.44, 0.8, col = "lightyellow", border = "orange3")
text(0.35, 0.6, "Optics\n(blur, PSF)", cex = 0.9, font = 2)

arrows(0.46, 0.6, 0.50, 0.6, lwd = 2)

rect(0.52, 0.4, 0.70, 0.8, col = "lightpink", border = "red3")
text(0.61, 0.6, "Cone Mosaic\n(3 types, noise)", cex = 0.9, font = 2)

arrows(0.72, 0.6, 0.76, 0.6, lwd = 2)

rect(0.78, 0.4, 0.98, 0.8, col = "lightgray", border = "gray30")
text(0.88, 0.6, "Cone Signals\n(L, M, S + noise)", cex = 0.9, font = 2)

text(0.5, 0.15, "Linear algebra + Poisson noise + Computational models (ISETBio)", cex = 1, font = 3)
```

**Key math:** n = C · e (matrix multiplication), Poisson noise, convolution

**Key insight:** Massive information loss at every stage!

## The Inverse Problem (Part 2)

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3

par(mar = c(1, 1, 2, 1))
plot.new()

# Inverse pipeline
rect(0.0, 0.4, 0.22, 0.8, col = "lightgray", border = "gray30")
text(0.11, 0.6, "Cone Signals\n(ambiguous!)", cex = 0.9, font = 2)

arrows(0.24, 0.6, 0.28, 0.6, lwd = 2)

rect(0.30, 0.4, 0.52, 0.8, col = "lightyellow", border = "orange3")
text(0.41, 0.65, "Likelihood", cex = 0.9, font = 2)
text(0.41, 0.5, "p(y|x)", cex = 1.1, font = 3)

text(0.56, 0.6, "×", cex = 2, font = 2)

rect(0.60, 0.4, 0.78, 0.8, col = "lightblue", border = "navy")
text(0.69, 0.65, "Prior", cex = 0.9, font = 2)
text(0.69, 0.5, "p(x)", cex = 1.1, font = 3)

arrows(0.80, 0.6, 0.84, 0.6, lwd = 2)

rect(0.86, 0.4, 1.0, 0.8, col = "lightgreen", border = "darkgreen")
text(0.93, 0.6, "Percept\np(x|y)", cex = 0.9, font = 2)

text(0.5, 0.15, "Bayesian inference = Helmholtz's Unconscious Inference", cex = 1, font = 3)
```

**Key math:** p(x\|y) ∝ p(y\|x) · p(x)

**Key insight:** Prior knowledge resolves ambiguity!

## Key Takeaways

**1. Vision is a forward + inverse problem:**

-   Forward: Scene → Optics → Cones → Signals (physics, linear algebra)
-   Inverse: Signals → Perception (Bayesian inference)

. . .

**2. Information is lost at every stage:**

-   ∞ wavelengths → 3 cone types (metamerism)
-   Continuous → discrete (sampling)
-   Exact → noisy (Poisson)

. . .

**3. The brain solves ambiguity using priors:**

-   Prior = learned statistics of the natural world
-   Perception = "best guess" given evidence + experience

. . .

**4. Computational models are essential:**

-   Math principles guide, but reality is complex
-   ISETBio: optics + mosaic + noise + eccentricity

## Final Thought

> "When you come to a fork in the road, take it." — Yogi Berra

. . .

**Multiple approaches to vision science:**

| Approach        | Focus                          |
|-----------------|--------------------------------|
| Mathematical    | Forward models, linear algebra |
| Behavioral      | Thresholds, matching, scaling  |
| Neurobiological | Anatomy, physiology            |
| Engineering     | Build systems that see         |

**All valid. All complementary. The field is large enough for everyone!**
