---
title: "Chapter 1: Initial Visual Encoding (Part 1)"
subtitle: "Forward Calculation: Scene to Cone Excitations"
author: "Wandell & Brainard - New Handbook of Mathematical Psychology Vol.3"
format:
  revealjs:
    theme: simple
    slide-number: true
    smaller: true
---

# Section 1.1: Introduction

## Vision Science and Mathematics

Vision research spans: physics, psychology, neuroscience, computation, engineering.

**Mathematics is the common language.**

. . .

| Core Method         | Purpose                  |
|---------------------|--------------------------|
| Linear algebra      | Describe transformations |
| Noise distributions | Quantify uncertainty     |
| Bayesian inference  | Probabilistic reasoning  |

## Chapter Roadmap

**Part 1: Forward Calculation (§1.2-1.4)**

Scene → Optics → Retinal image → Cone excitations

. . .

**Part 2: Perceptual Inference (§1.5)**

How does the brain interpret cone signals to see objects, depth, color?

. . .

**Helmholtz's Unconscious Inference (1867):** Perception is not passive recording — the brain *infers* the most likely cause of sensory signals, using prior experience, without conscious awareness.

## The Compression Problem

Light has **infinite wavelengths** (400-700nm continuous spectrum).

But we have only **3 cone types** → compressed to 3 numbers!

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 2.5
library(ggplot2)
wavelengths <- seq(400, 700, by = 5)
L_cone <- exp(-0.5 * ((wavelengths - 565) / 50)^2)
M_cone <- exp(-0.5 * ((wavelengths - 535) / 45)^2)
S_cone <- exp(-0.5 * ((wavelengths - 445) / 35)^2)

cone_data <- data.frame(
  wavelength = rep(wavelengths, 3),
  sensitivity = c(L_cone, M_cone, S_cone),
  cone = rep(c("L (long)", "M (medium)", "S (short)"), each = length(wavelengths))
)

ggplot(cone_data, aes(x = wavelength, y = sensitivity, color = cone)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("L (long)" = "red", "M (medium)" = "green", "S (short)" = "blue")) +
  labs(x = "Wavelength (nm)", y = "Sensitivity", color = NULL) +
  theme_minimal() + theme(legend.position = "top")
```

**∞ dimensions → 3 dimensions = information loss = ambiguity**

## Metamerism: Ambiguity in Action

**Different spectra → same cone response!**

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 2.2
e_yellow <- exp(-0.5 * ((wavelengths - 580) / 20)^2)
e_red <- exp(-0.5 * ((wavelengths - 620) / 25)^2)
e_green <- exp(-0.5 * ((wavelengths - 540) / 25)^2)
e_mixture <- 0.6 * e_red + 0.5 * e_green

metamer_data <- data.frame(
  wavelength = rep(wavelengths, 2),
  intensity = c(e_yellow / max(e_yellow), e_mixture / max(e_mixture)),
  spectrum = rep(c("Yellow light", "Red+Green mix"), each = length(wavelengths))
)

ggplot(metamer_data, aes(x = wavelength, y = intensity, color = spectrum)) +
  geom_line(linewidth = 1) +
  labs(x = "Wavelength (nm)", y = "Intensity", color = NULL) +
  theme_minimal() + theme(legend.position = "top")
```

| Possible Cause    | Cone Response       |
|-------------------|---------------------|
| Yellow banana     | L=0.7, M=0.5, S=0.1 |
| Red+Green mixture | L=0.7, M=0.5, S=0.1 |
| Orange sunset     | L=0.7, M=0.5, S=0.1 |

**The brain must guess:** Which scene is most likely?

## Bayesian Inference: The Brain's Solution

**Combine sensory evidence with prior knowledge:**

$$P(\text{scene} | \text{cones}) \propto \underbrace{P(\text{cones} | \text{scene})}_{\text{likelihood}} \times \underbrace{P(\text{scene})}_{\text{prior}}$$

. . .

| Term       | Question it Answers                                 |
|------------|-----------------------------------------------------|
| Likelihood | "How well does this scene explain my cone signals?" |
| Prior      | "How common is this scene in my experience?"        |
| Posterior  | "What scene should I perceive?"                     |

. . .

**Result:** Brain picks the scene that best explains the data AND is plausible given past experience.

# Section 1.2: Scene to Retinal Image

## The Light Field

**Leonardo da Vinci (1509):** A pinhole camera forms an image *wherever* you place it.

→ Light rays must fill all of space, traveling in all directions!

. . .

**Gershun (1939)** formalized this as the **light field:**

$$LF(x, y, z, \alpha, \beta, \lambda, \rho)$$

| Parameter | Meaning           |
|-----------|-------------------|
| x, y, z   | Position in space |
| α, β      | Direction of ray  |
| λ         | Wavelength        |
| ρ         | Polarization      |

## Visualizing the Light Field

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
library(ggplot2)

wall_y <- 10

ggplot() +
  # Draw the wall
  geom_segment(aes(x = -5, xend = 9, y = wall_y, yend = wall_y),
               color = "gray50", linewidth = 1) +
  # Draw colored objects on the wall
  geom_segment(aes(x = -3, xend = -1, y = wall_y, yend = wall_y),
               color = "red", linewidth = 6) +
  geom_segment(aes(x = 1, xend = 3, y = wall_y, yend = wall_y),
               color = "green3", linewidth = 6) +
  geom_segment(aes(x = 5, xend = 7, y = wall_y, yend = wall_y),
               color = "blue", linewidth = 6) +
  # Draw the observer (eye)
  geom_point(aes(x = 0, y = 0), size = 5, color = "black") +
  annotate("text", x = 0, y = -1, label = "Eye", size = 4) +
  # Draw rays from eye to objects
  geom_segment(aes(x = 0, xend = -2, y = 0, yend = wall_y),
               color = "red", linetype = "dashed", alpha = 0.7,
               arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 0, xend = 2, y = 0, yend = wall_y),
               color = "green3", linetype = "dashed", alpha = 0.7,
               arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 0, xend = 6, y = 0, yend = wall_y),
               color = "blue", linetype = "dashed", alpha = 0.7,
               arrow = arrow(length = unit(0.2, "cm"))) +
  # Labels
  annotate("text", x = -2, y = wall_y + 0.8, label = "α = -11°", color = "red", size = 3.5) +
  annotate("text", x = 2, y = wall_y + 0.8, label = "α = +11°", color = "green4", size = 3.5) +
  annotate("text", x = 6, y = wall_y + 0.8, label = "α = +31°", color = "blue", size = 3.5) +
  coord_fixed(ratio = 1) +
  xlim(-6, 10) + ylim(-2, 12) +
  labs(title = "Light Field: Rays from scene to eye",
       subtitle = "Each direction α sees a different object",
       x = "Position", y = "") +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

**Light field query:** "At position (x,y), looking in direction α, what wavelength λ do I see?"

## From Light Field to Retinal Image

The eye captures only rays entering the pupil → **Incident Light Field**

$$ILF(u, v, \alpha, \beta, \lambda, t)$$

. . .

The lens focuses rays onto retina → **Retinal Spectral Irradiance**

$$E(r_x, r_y, \lambda, t)$$

. . .

```{r}
#| echo: false
dimensions <- data.frame(
  Stage = c("Light Field", "Incident Light Field", "Retinal Image", "Cone Response"),
  Dimensions = c("7D", "7D", "4D", "3 numbers!"),
  Parameters = c("x,y,z,α,β,λ,ρ", "u,v,α,β,λ,ρ,t", "rx,ry,λ,t", "L, M, S")
)
knitr::kable(dimensions, align = "lcc")
```

**Massive dimensionality reduction!**

## The Transformation Chain

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3
library(ggplot2)

# Create boxes for each stage
boxes <- data.frame(
  x = c(1, 3, 5, 7),
  label = c("Scene", "Optics", "Cone\nMosaic", "Cone\nExcitations"),
  desc = c("Light sources\n+ objects", "Lens focuses\n+ blurs", "L, M, S\nsampling", "3 numbers\nper location")
)

ggplot(boxes) +
  # Boxes
  geom_tile(aes(x = x, y = 0), width = 1.6, height = 1.8, fill = "lightblue", color = "navy") +
  geom_text(aes(x = x, y = 0.3, label = label), fontface = "bold", size = 4) +
  geom_text(aes(x = x, y = -0.4, label = desc), size = 2.8, color = "gray30") +
  # Arrows
  annotate("segment", x = 1.85, xend = 2.15, y = 0, yend = 0,
           arrow = arrow(length = unit(0.3, "cm")), linewidth = 1) +
  annotate("segment", x = 3.85, xend = 4.15, y = 0, yend = 0,
           arrow = arrow(length = unit(0.3, "cm")), linewidth = 1) +
  annotate("segment", x = 5.85, xend = 6.15, y = 0, yend = 0,
           arrow = arrow(length = unit(0.3, "cm")), linewidth = 1) +
  xlim(0, 8) + ylim(-1.5, 1.5) +
  theme_void() +
  labs(title = "The Initial Visual Encoding Pipeline")
```

. . .

| Stage                | What happens          | Information change |
|----------------------|-----------------------|--------------------|
| Scene → Optics       | Light enters pupil    | Spatial filtering  |
| Optics → Mosaic      | Retinal image formed  | Blur + absorption  |
| Mosaic → Excitations | Cones integrate light | λ → 3 values       |

## What Gets Lost? (1) Optical Blur

The lens + diffraction blur the image before it reaches the retina:

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 3

# Create a simple "scene" - sharp stripes
x <- seq(-2, 2, length.out = 200)
sharp <- sin(10 * x)^2  # Sharp stripes

# Blur by convolution with Gaussian kernel
blur_kernel <- dnorm(seq(-1, 1, length.out = 50), sd = 0.15)
blur_kernel <- blur_kernel / sum(blur_kernel)
blurred <- stats::filter(sharp, blur_kernel, sides = 2)
blurred[is.na(blurred)] <- sharp[is.na(blurred)]

par(mfrow = c(1, 2), mar = c(3, 2, 3, 1))
image(matrix(sharp, nrow = 1), col = gray.colors(100), axes = FALSE, main = "Scene (sharp)")
image(matrix(blurred, nrow = 1), col = gray.colors(100), axes = FALSE, main = "Retinal image (blurred)")
```

Fine details are lost — the eye has limited spatial resolution!

## What Gets Lost? (2) Cone Sampling

Continuous retinal image → discrete cone mosaic samples:

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 3

set.seed(42)
# Create continuous "retinal image"
x_cont <- seq(0, 10, length.out = 200)
image_cont <- 0.5 + 0.3*sin(2*x_cont) + 0.2*sin(5*x_cont)

# Sample at discrete cone positions (irregularly spaced like real mosaic)
n_cones <- 25
cone_positions <- sort(runif(n_cones, 0.5, 9.5))
cone_samples <- approx(x_cont, image_cont, cone_positions)$y

par(mfrow = c(1, 2), mar = c(3, 3, 3, 1))
plot(x_cont, image_cont, type = "l", lwd = 2, col = "gray30",
     main = "Continuous retinal image", xlab = "", ylab = "Intensity")
plot(cone_positions, cone_samples, pch = 19, col = "darkred", cex = 1.2,
     main = "Discrete cone samples", xlab = "Position", ylab = "Intensity",
     xlim = c(0, 10), ylim = range(image_cont))
segments(cone_positions, 0, cone_positions, cone_samples, col = "darkred", lwd = 1)
```

Spatial information between cones is lost forever!

## What Gets Lost? (3) Metamers

**Metamers** = physically different spectra that produce **identical cone responses**.

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 3

wavelengths <- seq(400, 700, by = 5)

# Cone sensitivities
L <- exp(-0.5 * ((wavelengths - 565) / 50)^2)
M <- exp(-0.5 * ((wavelengths - 535) / 45)^2)
S <- exp(-0.5 * ((wavelengths - 445) / 35)^2)

# Spectrum A: Monochromatic yellow (580nm)
spectrum1 <- exp(-0.5 * ((wavelengths - 580) / 15)^2)

# Spectrum B: Red + Green mixture - SOLVE for exact match!
red_peak <- exp(-0.5 * ((wavelengths - 630) / 20)^2)
green_peak <- exp(-0.5 * ((wavelengths - 530) / 20)^2)

# Target cone responses from yellow
target_L <- sum(spectrum1 * L)
target_M <- sum(spectrum1 * M)

# Solve: w_r * red_L + w_g * green_L = target_L
#        w_r * red_M + w_g * green_M = target_M
red_L <- sum(red_peak * L); red_M <- sum(red_peak * M)
green_L <- sum(green_peak * L); green_M <- sum(green_peak * M)

# Matrix solve: [red_L green_L; red_M green_M] * [w_r; w_g] = [target_L; target_M]
A <- matrix(c(red_L, red_M, green_L, green_M), nrow = 2)
b <- c(target_L, target_M)
weights <- solve(A, b)

spectrum2 <- weights[1] * red_peak + weights[2] * green_peak

# Calculate cone responses (should match!)
resp1 <- c(sum(spectrum1 * L), sum(spectrum1 * M), sum(spectrum1 * S))
resp2 <- c(sum(spectrum2 * L), sum(spectrum2 * M), sum(spectrum2 * S))

# Normalize for display
max_resp <- max(c(resp1, resp2))
resp1 <- resp1 / max_resp
resp2 <- resp2 / max_resp

par(mfrow = c(1, 3), mar = c(4, 3, 3, 1))
plot(wavelengths, spectrum1/max(spectrum1), type = "l", lwd = 3, col = "goldenrod",
     main = "Spectrum A: Yellow light", xlab = "λ (nm)", ylab = "", ylim = c(0, 1))
plot(wavelengths, spectrum2/max(spectrum2), type = "l", lwd = 3, col = "darkgreen",
     main = "Spectrum B: Red+Green mix", xlab = "λ (nm)", ylab = "", ylim = c(0, 1))
barplot(rbind(resp1, resp2), beside = TRUE, col = c("goldenrod", "darkgreen"),
        names.arg = c("L", "M", "S"), main = "Same cone response!",
        legend = c("Yellow", "R+G"), args.legend = list(x = "topright", cex = 0.8))
```

**Can we tell them apart? NO!** Same L, M, S → identical perception.

→ This is why the brain must use **Bayesian inference** to guess the scene!

**Different physics → identical perception!** This is why TV/phone screens work with just RGB pixels.

# Section 1.3: Mathematical Principles

## Linear Systems: The Key Tool

A system $L$ is **linear** if it follows the **superposition rule**:

$$L(x + y) = L(x) + L(y)$$

. . .

This implies **homogeneity**: $L(mx) = mL(x)$

. . .

**Why is this powerful?**

If we know how a system responds to **basis inputs**, we can predict its response to **any input** (as a weighted sum)!

## How Cones Compute Their Response

$$N = \int C(\lambda) \cdot E(\lambda) \, d\lambda = \text{weighted sum of light}$$

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4

wavelengths <- seq(400, 700, by = 5)

# Green light spectrum
E <- exp(-0.5 * ((wavelengths - 530) / 25)^2)

# L-cone sensitivity
C <- exp(-0.5 * ((wavelengths - 565) / 50)^2)

# Product
product <- C * E

# Response = sum (area under product)
N <- sum(product) * 5  # multiply by delta_lambda

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# Step 1: Spectrum
plot(wavelengths, E, type = "l", lwd = 3, col = "green4",
     main = "Step 1: Light spectrum E(λ)", xlab = "λ (nm)", ylab = "Intensity")
polygon(c(wavelengths, rev(wavelengths)), c(E, rep(0, length(E))),
        col = rgb(0, 0.5, 0, 0.3), border = NA)

# Step 2: Cone sensitivity
plot(wavelengths, C, type = "l", lwd = 3, col = "red",
     main = "Step 2: L-cone sensitivity C(λ)", xlab = "λ (nm)", ylab = "Sensitivity")
polygon(c(wavelengths, rev(wavelengths)), c(C, rep(0, length(C))),
        col = rgb(1, 0, 0, 0.3), border = NA)

# Step 3: Product
plot(wavelengths, product, type = "l", lwd = 3, col = "purple",
     main = "Step 3: Multiply C(λ) × E(λ)", xlab = "λ (nm)", ylab = "Product")
polygon(c(wavelengths, rev(wavelengths)), c(product, rep(0, length(product))),
        col = rgb(0.5, 0, 0.5, 0.3), border = NA)

# Step 4: Sum = N
barplot(N, col = "purple", main = "Step 4: Sum all = N",
        ylab = "Cone response", ylim = c(0, N * 1.2))
text(0.7, N * 0.5, paste0("N = ", round(N, 1)), cex = 1.5, col = "white", font = 2)
```

**Result:** One number N = how strongly this cone responds to this light!

## Matrix Form: n = C · e

**Simplified example** with 5 wavelengths (450, 500, 550, 600, 650 nm):

$$\underbrace{\begin{bmatrix} n_L \\ n_M \\ n_S \end{bmatrix}}_{\mathbf{n}\,(3 \times 1)} = \underbrace{\begin{bmatrix} 0.1 & 0.3 & 0.7 & 1.0 & 0.8 \\ 0.2 & 0.6 & 1.0 & 0.6 & 0.2 \\ 1.0 & 0.3 & 0.0 & 0.0 & 0.0 \end{bmatrix}}_{\mathbf{C}\,(3 \times 5)} \cdot \underbrace{\begin{bmatrix} 0.0 \\ 0.2 \\ 0.8 \\ 1.0 \\ 0.4 \end{bmatrix}}_{\mathbf{e}\,(5 \times 1)}$$

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3

# Define matrices
C <- matrix(c(0.1, 0.2, 1.0,
              0.3, 0.6, 0.3,
              0.7, 1.0, 0.0,
              1.0, 0.6, 0.0,
              0.8, 0.2, 0.0), nrow = 3, ncol = 5)
rownames(C) <- c("L", "M", "S")
colnames(C) <- c("450", "500", "550", "600", "650")

e <- c(0.0, 0.2, 0.8, 1.0, 0.4)
n <- C %*% e

layout(matrix(c(1, 2, 3, 4, 5), nrow = 1), widths = c(3, 1, 2, 1, 2))
par(mar = c(3, 3, 3, 1))

# C matrix as heatmap
image(t(C[3:1, ]), col = heat.colors(20), axes = FALSE, main = "C matrix (3×5)")
axis(1, at = seq(0, 1, length.out = 5), labels = colnames(C), cex.axis = 0.8)
axis(2, at = c(0, 0.5, 1), labels = c("S", "M", "L"), las = 2, cex.axis = 0.9)
mtext("wavelength (nm)", side = 1, line = 2, cex = 0.7)

# Multiply sign
par(mar = c(0, 0, 0, 0))
plot.new()
text(0.5, 0.5, "×", cex = 4, font = 2)

# e vector
par(mar = c(3, 3, 3, 1))
barplot(e, names.arg = c("450", "500", "550", "600", "650"),
        col = "orange", main = "e vector (5×1)", cex.names = 0.7)
mtext("spectrum", side = 1, line = 2, cex = 0.7)

# Equals sign
par(mar = c(0, 0, 0, 0))
plot.new()
text(0.5, 0.5, "=", cex = 4, font = 2)

# n result
par(mar = c(3, 3, 3, 1))
barplot(as.vector(n), names.arg = c("L", "M", "S"),
        col = c("red", "green", "blue"), main = "n (3×1)")
```

| Matrix | Meaning                                         |
|--------|-------------------------------------------------|
| **C**  | Cone sensitivities (brighter = more sensitive)  |
| **e**  | Spectrum (yellow-ish light: peaks at 550-600nm) |
| **n**  | Result: L responds most, S responds least       |

## Matrix Multiplication: Step by Step

**Row × Column = one number:**

$$n_L = [0.1, 0.3, 0.7, 1.0, 0.8] \cdot \begin{bmatrix} 0.0 \\ 0.2 \\ 0.8 \\ 1.0 \\ 0.4 \end{bmatrix}$$

. . .

$$n_L = (0.1 \times 0.0) + (0.3 \times 0.2) + (0.7 \times 0.8) + (1.0 \times 1.0) + (0.8 \times 0.4)$$

$$n_L = 0 + 0.06 + 0.56 + 1.0 + 0.32 = \mathbf{1.94}$$

. . .

Same for M and S → final result: $\mathbf{n} = [1.94, 1.24, 0.06]^T$

## Why Matrix Form is Powerful

**Measure C once → predict response to ANY spectrum!**

. . .

| You have                      | You can predict        |
|-------------------------------|------------------------|
| C matrix (cone sensitivities) | Response to sunlight   |
| \+ any spectrum e             | Response to LED screen |
|                               | Response to laser      |
|                               | Response to ANY light! |

. . .

This is why **color science works**: measure 3 cone curves, predict perception of infinite possible spectra!

## Principle of Univariance

> "The output of a receptor depends upon its quantum catch, but not upon what quanta are caught." — Rushton (1972)

**Translation:** A cone reports HOW MANY photons, not WHAT COLOR they were.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3

par(mfrow = c(1, 3), mar = c(2, 2, 4, 1))

# Scenario 1: Red photons
barplot(c(100), col = "red", ylim = c(0, 120), main = "100 red photons absorbed")
text(0.7, 110, "Cone says:\n'100'", cex = 1.2, font = 2)

# Scenario 2: Green photons
barplot(c(100), col = "green", ylim = c(0, 120), main = "100 green photons absorbed")
text(0.7, 110, "Cone says:\n'100'", cex = 1.2, font = 2)

# Scenario 3: Mixed
barplot(c(50, 50), col = c("red", "green"), ylim = c(0, 120), main = "50 red + 50 green absorbed")
text(0.7, 110, "Cone says:\n'100'", cex = 1.2, font = 2)
```

**Same response!** Once absorbed, wavelength information is **lost forever**.

## Why Univariance Matters

A single cone is **color-blind**. To see color, compare L, M, S:

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3

par(mfrow = c(1, 3), mar = c(3, 3, 3, 1))

# Red light
barplot(c(L = 100, M = 30, S = 5), col = c("red", "green", "blue"),
        main = "Red light → L wins", ylim = c(0, 120))

# Green light
barplot(c(L = 50, M = 90, S = 10), col = c("red", "green", "blue"),
        main = "Green light → M wins", ylim = c(0, 120))

# Blue light
barplot(c(L = 10, M = 20, S = 80), col = c("red", "green", "blue"),
        main = "Blue light → S wins", ylim = c(0, 120))
```

**Color = pattern of L, M, S responses**, not any single cone's output!

## Why Are Cone Responses Noisy?

Photons arrive **randomly** — like raindrops falling on a bucket.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3

set.seed(42)
par(mfrow = c(1, 3), mar = c(3, 3, 3, 1))

# Three trials of counting photons
trial1 <- rpois(1, 100)
trial2 <- rpois(1, 100)
trial3 <- rpois(1, 100)

barplot(trial1, col = "steelblue", ylim = c(0, 130), main = "Trial 1")
text(0.7, trial1 + 10, trial1, cex = 1.5, font = 2)

barplot(trial2, col = "steelblue", ylim = c(0, 130), main = "Trial 2")
text(0.7, trial2 + 10, trial2, cex = 1.5, font = 2)

barplot(trial3, col = "steelblue", ylim = c(0, 130), main = "Trial 3")
text(0.7, trial3 + 10, trial3, cex = 1.5, font = 2)
```

**Same light source (mean = 100)**, but each measurement is different!

This randomness follows a **Poisson distribution**.

## Poisson Distribution: Shape Depends on Mean

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3.5

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Poisson with mean = 10
x_dim <- 0:25
prob_dim <- dpois(x_dim, lambda = 10)
barplot(prob_dim, names.arg = x_dim, col = "steelblue",
        main = "Poisson (μ = 10)", xlab = "Photon count", ylab = "Probability")
abline(v = 10 * 1.2 + 0.5, col = "red", lwd = 2, lty = 2)
text(15, max(prob_dim) * 0.9, "mean = 10\nSD = 3.2", col = "red", cex = 1)

# Poisson with mean = 100
x_bright <- 70:130
prob_bright <- dpois(x_bright, lambda = 100)
barplot(prob_bright, names.arg = x_bright, col = "gold",
        main = "Poisson (μ = 100)", xlab = "Photon count", ylab = "Probability",
        cex.names = 0.6)
text(40, max(prob_bright) * 0.9, "mean = 100\nSD = 10", col = "red", cex = 1)
```

**Key insight: Variance = Mean**

| Mean (μ) | Variance (σ²) | SD (σ)    | Spread |
|----------|---------------|-----------|--------|
| 10       | 10            | √10 ≈ 3.2 | Narrow |
| 100      | 100           | √100 = 10 | Wider  |

More photons → wider distribution → but **relative** noise (SD/mean) gets smaller!

## Poisson vs Gaussian

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Small mean: Poisson ≠ Gaussian
x_small <- 0:20
pois_small <- dpois(x_small, 5)
gaus_small <- dnorm(x_small, 5, sqrt(5))

plot(x_small, pois_small, type = "h", lwd = 3, col = "steelblue",
     main = "μ = 5: Poisson ≠ Gaussian", xlab = "Count", ylab = "Probability",
     ylim = c(0, max(pois_small) * 1.1))
lines(x_small, gaus_small, col = "red", lwd = 2)
legend("topright", c("Poisson", "Gaussian"), col = c("steelblue", "red"), lwd = c(3, 2), cex = 0.8)

# Large mean: Poisson ≈ Gaussian
x_large <- 70:130
pois_large <- dpois(x_large, 100)
gaus_large <- dnorm(x_large, 100, sqrt(100))

plot(x_large, pois_large, type = "h", lwd = 3, col = "steelblue",
     main = "μ = 100: Poisson ≈ Gaussian", xlab = "Count", ylab = "Probability",
     ylim = c(0, max(pois_large) * 1.1))
lines(x_large, gaus_large, col = "red", lwd = 2)
legend("topright", c("Poisson", "Gaussian"), col = c("steelblue", "red"), lwd = c(3, 2), cex = 0.8)
```

|                   | Poisson               | Gaussian                |
|-------------------|-----------------------|-------------------------|
| **Parameters**    | μ only (variance = μ) | μ and σ² (independent)  |
| **Values**        | Integers ≥ 0          | Any real number         |
| **Use for**       | Counting (photons!)   | Continuous measurements |
| **When similar?** | When μ \> \~30        | —                       |

## What is Shift-Invariance?

**Question:** If I move a dot across my visual field, does it stay the same shape?

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3

par(mfrow = c(1, 3), mar = c(2, 2, 3, 1))

# Create a simple "retinal image" with a dot at different positions
x <- seq(-5, 5, length.out = 100)

# Dot in the center
dot_center <- exp(-x^2 / 0.2)
plot(x, dot_center, type = "l", lwd = 3, col = "blue",
     main = "Dot at center", xlab = "", ylab = "", ylim = c(0, 1.2))
polygon(c(x, rev(x)), c(dot_center, rep(0, length(x))), col = rgb(0, 0, 1, 0.3), border = NA)

# Dot shifted left
dot_left <- exp(-(x + 3)^2 / 0.2)
plot(x, dot_left, type = "l", lwd = 3, col = "blue",
     main = "Dot shifted left", xlab = "", ylab = "", ylim = c(0, 1.2))
polygon(c(x, rev(x)), c(dot_left, rep(0, length(x))), col = rgb(0, 0, 1, 0.3), border = NA)

# Dot shifted right
dot_right <- exp(-(x - 3)^2 / 0.2)
plot(x, dot_right, type = "l", lwd = 3, col = "blue",
     main = "Dot shifted right", xlab = "", ylab = "", ylim = c(0, 1.2))
polygon(c(x, rev(x)), c(dot_right, rep(0, length(x))), col = rgb(0, 0, 1, 0.3), border = NA)
```

**Shift-invariant:** Same blur happens everywhere → shift input = shift output (same shape).

## Point Spread Function (PSF)

**The PSF tells us: "What does a single point of light look like after passing through the eye?"**

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3.5

par(mfrow = c(1, 3), mar = c(3, 3, 3, 1))

# Perfect point (scene)
x <- seq(-3, 3, length.out = 100)
perfect <- ifelse(abs(x) < 0.1, 1, 0)
plot(x, perfect, type = "h", lwd = 3, col = "black",
     main = "Scene: Perfect point", xlab = "Position", ylab = "Intensity", ylim = c(0, 1.2))

# Arrow
plot.new()
arrows(0.1, 0.5, 0.9, 0.5, lwd = 4, col = "gray40")
text(0.5, 0.7, "Through\nthe eye", cex = 1.2, font = 2)

# Blurred PSF (retinal image)
psf <- exp(-x^2 / 0.5)
plot(x, psf, type = "l", lwd = 3, col = "red",
     main = "Retinal image: Blurred blob", xlab = "Position", ylab = "Intensity", ylim = c(0, 1.2))
polygon(c(x, rev(x)), c(psf, rep(0, length(x))), col = rgb(1, 0, 0, 0.3), border = NA)
```

**PSF = the "blur kernel"** — the lens + diffraction spreads a perfect point into a blob.

## What is Convolution?

**Convolution = "slide the PSF across every point in the image and add up"**

Think of it as **stamping** the PSF at every bright location:

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4

par(mfrow = c(2, 2), mar = c(3, 3, 3, 1))

x <- seq(-5, 5, length.out = 200)

# Scene: Two points
scene <- ifelse(abs(x + 2) < 0.1 | abs(x - 2) < 0.1, 1, 0)
plot(x, scene, type = "h", lwd = 3, col = "black",
     main = "Step 1: Scene (two points)", xlab = "x", ylab = "")

# PSF
psf_x <- seq(-2, 2, length.out = 50)
psf_y <- exp(-psf_x^2 / 0.3)
psf_y <- psf_y / max(psf_y)
plot(psf_x, psf_y, type = "l", lwd = 3, col = "red",
     main = "Step 2: PSF (blur kernel)", xlab = "x", ylab = "")
polygon(c(psf_x, rev(psf_x)), c(psf_y, rep(0, length(psf_x))), col = rgb(1, 0, 0, 0.3), border = NA)

# Stamp PSF at each point
blob1 <- exp(-(x + 2)^2 / 0.3)
blob2 <- exp(-(x - 2)^2 / 0.3)
plot(x, blob1, type = "l", lwd = 2, col = "blue", ylim = c(0, 1.2),
     main = "Step 3: Stamp PSF at each point", xlab = "x", ylab = "")
lines(x, blob2, lwd = 2, col = "green4")
legend("topright", c("Left point", "Right point"), col = c("blue", "green4"), lwd = 2, cex = 0.8)

# Sum = convolution result
result <- blob1 + blob2
plot(x, result, type = "l", lwd = 3, col = "purple",
     main = "Step 4: Add up = Output", xlab = "x", ylab = "")
polygon(c(x, rev(x)), c(result, rep(0, length(x))), col = rgb(0.5, 0, 0.5, 0.2), border = NA)
```

**Convolution = sum of all the "stamps"**

## Convolution in Action: Edge Blur

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3

# 1D demonstration: sharp edge blurred by PSF
x <- 1:100
sharp_edge <- c(rep(0, 50), rep(1, 50))

# PSF: Gaussian blur
psf <- dnorm(seq(-3, 3, length.out = 21), sd = 1)
psf <- psf / sum(psf)

# Convolve
blurred <- stats::filter(sharp_edge, psf, sides = 2)
blurred[is.na(blurred)] <- sharp_edge[is.na(blurred)]

par(mfrow = c(1, 3), mar = c(4, 3, 3, 1))
plot(x, sharp_edge, type = "l", lwd = 3, main = "Sharp edge (scene)", xlab = "x", ylab = "")
plot(seq(-10, 10, length.out = 21), psf, type = "l", lwd = 3, col = "red",
     main = "PSF (blur kernel)", xlab = "x", ylab = "")
polygon(c(seq(-10, 10, length.out = 21), rev(seq(-10, 10, length.out = 21))),
        c(psf, rep(0, 21)), col = rgb(1, 0, 0, 0.3), border = NA)
plot(x, blurred, type = "l", lwd = 3, col = "purple",
     main = "Blurred edge (retina)", xlab = "x", ylab = "")
```

**Sharp edge + blur kernel = soft gradient** — this is why we can't see infinitely fine detail!

## The Convolution Formula

$$\text{Retinal image}(x) = \int \text{Scene}(u) \cdot \text{PSF}(x - u) \, du$$

. . .

**In words:**

| Symbol     | Meaning                            |
|------------|------------------------------------|
| Scene(u)   | Intensity at each scene position u |
| PSF(x - u) | The blur kernel centered at u      |
| ∫ ... du   | Sum over all positions             |
| Result     | Blurred image at position x        |

. . .

**Why is this powerful?** Measure PSF once → predict the retinal image of ANY scene!

## Reality Check: The Eye is NOT Perfectly Shift-Invariant

**Shift-invariance assumes:** Same PSF everywhere. But the real eye varies!

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3

par(mfrow = c(1, 3), mar = c(3, 3, 3, 1))

x <- seq(-3, 3, length.out = 100)

# Center of retina (fovea): sharp PSF
psf_center <- exp(-x^2 / 0.1)
plot(x, psf_center, type = "l", lwd = 3, col = "darkgreen",
     main = "Fovea (center): Sharp", xlab = "Position", ylab = "PSF", ylim = c(0, 1.1))
polygon(c(x, rev(x)), c(psf_center, rep(0, length(x))), col = rgb(0, 0.5, 0, 0.3), border = NA)
text(0, 0.5, "High\nresolution", cex = 1, font = 2)

# Mid-periphery: medium PSF
psf_mid <- exp(-x^2 / 0.4)
plot(x, psf_mid, type = "l", lwd = 3, col = "orange",
     main = "Mid-periphery: Wider", xlab = "Position", ylab = "PSF", ylim = c(0, 1.1))
polygon(c(x, rev(x)), c(psf_mid, rep(0, length(x))), col = rgb(1, 0.5, 0, 0.3), border = NA)
text(0, 0.5, "Medium\nresolution", cex = 1, font = 2)

# Far periphery: blurry PSF
psf_periph <- exp(-x^2 / 1.0)
plot(x, psf_periph, type = "l", lwd = 3, col = "red",
     main = "Periphery: Blurry", xlab = "Position", ylab = "PSF", ylim = c(0, 1.1))
polygon(c(x, rev(x)), c(psf_periph, rep(0, length(x))), col = rgb(1, 0, 0, 0.3), border = NA)
text(0, 0.5, "Low\nresolution", cex = 1, font = 2)
```

. . .

| Where it varies | Why                                                  |
|-----------------|------------------------------------------------------|
| Eccentricity    | Fewer cones, more aberrations in periphery           |
| Wavelength      | Lens bends blue more than red (chromatic aberration) |
| Pupil size      | Larger pupil → more blur from aberrations            |

**Consequence:** Simple convolution works locally, but computational models (ISETBio) needed for full accuracy.

# Section 1.4: Computational Model

## Why We Need Computational Models

**Math formulas work for one location.** But the real eye **varies everywhere!**

| Parameter     | Fovea (0°)    | Periphery (10°+) |
|---------------|---------------|------------------|
| PSF (blur)    | Small, sharp  | Large, blurry    |
| Cone density  | \~200,000/mm² | \~10,000/mm²     |
| Cone aperture | Small         | Large            |
| S-cones       | **None!**     | Present          |

. . .

> "There is little value in expressing the full complexity of these calculations in pure mathematical form." — Wandell & Brainard

**Solution:** Computational modeling with **ISETBio** (free, open-source)

## Figure 1.5: PSF Varies with Eccentricity and Wavelength

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4

par(mfrow = c(2, 3), mar = c(3, 3, 3, 1))

# Simulate PSF at different eccentricities for 550nm (top row)
x <- seq(-5, 5, length.out = 50)
y <- seq(-5, 5, length.out = 50)

# 2 deg, 550nm - sharp
psf_2deg <- outer(x, y, function(x, y) exp(-(x^2 + y^2) / 1))
image(x, y, psf_2deg, col = heat.colors(20), main = "2° eccentricity (550nm)",
      xlab = "μm", ylab = "μm", axes = FALSE)
box()

# 8 deg, 550nm - medium
psf_8deg <- outer(x, y, function(x, y) exp(-(x^2 + y^2) / 3))
image(x, y, psf_8deg, col = heat.colors(20), main = "8° eccentricity (550nm)",
      xlab = "μm", ylab = "μm", axes = FALSE)
box()

# 20 deg, 550nm - blurry
psf_20deg <- outer(x, y, function(x, y) exp(-(x^2 + y^2) / 8))
image(x, y, psf_20deg, col = heat.colors(20), main = "20° eccentricity (550nm)",
      xlab = "μm", ylab = "μm", axes = FALSE)
box()

# Same for 450nm (bottom row) - even blurrier due to chromatic aberration
psf_2deg_450 <- outer(x, y, function(x, y) exp(-(x^2 + y^2) / 2))
image(x, y, psf_2deg_450, col = heat.colors(20), main = "2° eccentricity (450nm)",
      xlab = "μm", ylab = "μm", axes = FALSE)
box()

psf_8deg_450 <- outer(x, y, function(x, y) exp(-(x^2 + y^2) / 5))
image(x, y, psf_8deg_450, col = heat.colors(20), main = "8° eccentricity (450nm)",
      xlab = "μm", ylab = "μm", axes = FALSE)
box()

psf_20deg_450 <- outer(x, y, function(x, y) exp(-(x^2 + y^2) / 12))
image(x, y, psf_20deg_450, col = heat.colors(20), main = "20° eccentricity (450nm)",
      xlab = "μm", ylab = "μm", axes = FALSE)
box()
```

**Top row (550nm):** PSF increases with eccentricity.

**Bottom row (450nm):** Even blurrier! Eye can't focus both wavelengths — **chromatic aberration** is the largest aberration.

## Figure 1.6: Cone Mosaic Changes Across Retina

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3

par(mfrow = c(1, 4), mar = c(2, 2, 3, 1))

set.seed(42)

# 0 degrees - very dense, small cones, NO S-cones
n0 <- 400
x0 <- runif(n0, 0, 10); y0 <- runif(n0, 0, 10)
# Only L and M cones in central fovea!
cols0 <- sample(c("red", "green3"), n0, replace = TRUE, prob = c(0.5, 0.5))
plot(x0, y0, pch = 19, cex = 0.6, col = cols0, xlim = c(0, 10), ylim = c(0, 10),
     main = "0° (fovea)", xlab = "", ylab = "", axes = FALSE, asp = 1)
box()
text(5, -0.5, "No S-cones!", cex = 0.9, font = 3, col = "blue")

# 4 degrees - still dense, some S-cones appear
n4 <- 250
x4 <- runif(n4, 0, 10); y4 <- runif(n4, 0, 10)
cols4 <- sample(c("red", "green3", "blue"), n4, replace = TRUE, prob = c(0.4, 0.4, 0.2))
plot(x4, y4, pch = 19, cex = 0.9, col = cols4, xlim = c(0, 10), ylim = c(0, 10),
     main = "4°", xlab = "", ylab = "", axes = FALSE, asp = 1)
box()

# 15 degrees - sparse cones, add rods (gray)
n15_cones <- 80
x15 <- runif(n15_cones, 0, 10); y15 <- runif(n15_cones, 0, 10)
cols15 <- sample(c("red", "green3", "blue"), n15_cones, replace = TRUE, prob = c(0.35, 0.35, 0.3))
plot(x15, y15, pch = 19, cex = 1.5, col = cols15, xlim = c(0, 10), ylim = c(0, 10),
     main = "15°", xlab = "", ylab = "", axes = FALSE, asp = 1)
# Add rods
n_rods <- 200
xr <- runif(n_rods, 0, 10); yr <- runif(n_rods, 0, 10)
points(xr, yr, pch = 19, cex = 0.4, col = "gray50")
box()

# 25 degrees - very sparse large cones, many rods
n25_cones <- 40
x25 <- runif(n25_cones, 0, 10); y25 <- runif(n25_cones, 0, 10)
cols25 <- sample(c("red", "green3", "blue"), n25_cones, replace = TRUE, prob = c(0.35, 0.35, 0.3))
plot(x25, y25, pch = 19, cex = 2.2, col = cols25, xlim = c(0, 10), ylim = c(0, 10),
     main = "25°", xlab = "", ylab = "", axes = FALSE, asp = 1)
n_rods2 <- 300
xr2 <- runif(n_rods2, 0, 10); yr2 <- runif(n_rods2, 0, 10)
points(xr2, yr2, pch = 19, cex = 0.3, col = "gray50")
box()
```

**Fovea → Periphery:** Cone density ↓, aperture size ↑, rods appear (gray dots).

**Surprise:** Central fovea has **no S-cones** — it's dichromatic! (Optimizes spatial resolution by omitting blurry short-wavelength channel)

## Figure 1.7: Fovea vs Periphery — The Trade-off

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4

par(mfrow = c(2, 2), mar = c(3, 3, 3, 1))

# Simulate grating responses
x <- seq(0, 2*pi, length.out = 100)

# 6 c/deg grating - both can resolve
grating_6 <- sin(6 * x)

# Fovea sampling (dense)
fov_pos <- seq(0, 2*pi, length.out = 60)
fov_samples <- sin(6 * fov_pos) + rnorm(60, 0, 0.1)

plot(x, grating_6, type = "l", lwd = 2, col = "gray70", main = "Fovea: 6 c/deg grating",
     xlab = "Position", ylab = "Response", ylim = c(-1.5, 1.5))
points(fov_pos, fov_samples, pch = 19, cex = 0.8, col = "darkred")
text(pi, 1.3, "Pattern preserved!", cex = 1, font = 2)

# Periphery sampling (sparse) for 6 c/deg
per_pos <- seq(0, 2*pi, length.out = 15)
per_samples <- sin(6 * per_pos) + rnorm(15, 0, 0.15)

plot(x, grating_6, type = "l", lwd = 2, col = "gray70", main = "10° Periphery: 6 c/deg",
     xlab = "Position", ylab = "Response", ylim = c(-1.5, 1.5))
points(per_pos, per_samples, pch = 19, cex = 1.5, col = "darkblue")
text(pi, 1.3, "Still OK", cex = 1, font = 2)

# 12 c/deg grating - fovea can resolve, periphery can't
grating_12 <- sin(12 * x)

fov_samples_12 <- sin(12 * fov_pos) + rnorm(60, 0, 0.1)
plot(x, grating_12, type = "l", lwd = 2, col = "gray70", main = "Fovea: 12 c/deg grating",
     xlab = "Position", ylab = "Response", ylim = c(-1.5, 1.5))
points(fov_pos, fov_samples_12, pch = 19, cex = 0.8, col = "darkred")
text(pi, 1.3, "Pattern preserved!", cex = 1, font = 2)

# Periphery can't resolve 12 c/deg - aliasing!
per_samples_12 <- sin(12 * per_pos) + rnorm(15, 0, 0.2)
plot(x, grating_12, type = "l", lwd = 2, col = "gray70", main = "10° Periphery: 12 c/deg",
     xlab = "Position", ylab = "Response", ylim = c(-1.5, 1.5))
points(per_pos, per_samples_12, pch = 19, cex = 1.5, col = "darkblue")
text(pi, 1.3, "Wobbly! Pattern lost", cex = 1, font = 2, col = "red")
```

**Fovea:** Dense sampling → fine patterns preserved.

**Periphery:** Sparse sampling → high-frequency patterns look "wobbly" (aliased).

## The Fovea-Periphery Trade-off

|   | Fovea | Periphery |
|----|----|----|
| **Cone density** | High (\~200k/mm²) | Low (\~10k/mm²) |
| **Cone aperture** | Small | Large |
| **Excitations/cone** | Fewer | **More** (larger apertures!) |
| **Spatial resolution** | High | Low |
| **Color vision** | Dichromatic (L, M only) | Trichromatic (L, M, S) |

. . .

**Why?**

-   **Fovea:** Optimized for fine detail, reading, recognition
-   **Periphery:** Optimized for detecting motion, threats (sensitivity over resolution)

## Spatial Derivatives: What Neurons Compute

**Neurons don't report absolute light — they compute local changes (derivatives):**

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

# Image intensity (smooth gradient with edge)
x <- seq(0, 10, length.out = 100)
intensity <- 0.3 + 0.5 * pnorm(x, 5, 0.8)

plot(x, intensity, type = "l", lwd = 3, col = "gray30",
     main = "Light intensity", xlab = "Position", ylab = "Intensity", ylim = c(0, 1))

# Spatial derivative → edge detection
deriv <- c(0, diff(intensity)) * 15
plot(x, deriv, type = "l", lwd = 3, col = "purple",
     main = "Spatial derivative → Edge", xlab = "Position", ylab = "Change")
abline(h = 0, lty = 2, col = "gray")
polygon(c(x, rev(x)), c(pmax(deriv, 0), rep(0, length(x))), col = rgb(0.5, 0, 0.5, 0.3), border = NA)

# Summary as visual
plot.new()
text(0.5, 0.95, "Neurons detect:", cex = 1.3, font = 2)
text(0.1, 0.75, "Spatial Δ", adj = 0, cex = 1.1)
text(0.65, 0.75, "→ Edges", adj = 0, col = "purple", cex = 1.1, font = 2)
text(0.1, 0.55, "Wavelength Δ", adj = 0, cex = 1.1)
text(0.65, 0.55, "→ Color", adj = 0, col = "orange", cex = 1.1, font = 2)
text(0.1, 0.35, "Time Δ", adj = 0, cex = 1.1)
text(0.65, 0.35, "→ Motion", adj = 0, col = "blue", cex = 1.1, font = 2)
text(0.1, 0.15, "Left-Right Δ", adj = 0, cex = 1.1)
text(0.65, 0.15, "→ Depth", adj = 0, col = "darkgreen", cex = 1.1, font = 2)
```

**Why derivatives?** Local contrast matters more than absolute brightness. Also highly compressible (most derivatives ≈ 0 in natural scenes).

## Part 1 Summary: Forward Calculation Complete

**Scene → Optics → Cone Mosaic → Cone Excitations**

| Stage             | What happens          | Key math               |
|-------------------|-----------------------|------------------------|
| **Optics**        | Light blurred by PSF  | Convolution (locally)  |
| **Cone sampling** | Continuous → discrete | Matrix multiplication  |
| **Cone response** | Photons counted       | Linear + Poisson noise |
| **Neural coding** | Local differences     | Derivatives            |

. . .

**Key insight:** Massive information loss at every stage!

-   ∞ wavelengths → 3 cone types
-   Continuous image → discrete samples
-   Exact photon count → noisy estimate

**The brain receives ambiguous, noisy signals. How does it perceive the world?**

→ **Part 2: Perceptual Inference** (Bayesian approach)
