---
title: "Exercises: Why Math in Psychology?"
subtitle: "New Handbook of Mathematical Psychology Vol.3 - Preface"
categories: [preface, measurement, psychophysics]
---

## Conceptual Questions

### Open-ended

**Q1.** In your own words, explain the core message of the preface. Why do the authors argue that mathematics in psychology is "not about computing"?

**Q2.** In the Dupin example from Edgar Allan Poe, what hidden assumptions lead to the false conclusion that "mathematics does not apply to motives"? List at least three.

**Q3.** Compare and contrast Luce's approach (without dimensional constants) and Rozeboom's critique (with dimensional constants). What is the key difference in their conclusions about psychophysical laws?

### Multiple Choice

**Q4.** Which measurement scale allows only monotonic transformations?

a)  Nominal scale
b)  Ordinal scale
c)  Interval scale
d)  Ratio scale

**Q5.** For a ratio scale, if f and g are two valid measurement functions, what is their relationship?

a)  f(x) = g(x) + c for some constant c
b)  f(x) = k · g(x) for some positive constant k
c)  f(x) = a · g(x) + b for constants a and b
d)  f(x) and g(x) are unrelated

**Q6.** What was Luce's hidden assumption that Rozeboom identified?

a)  Stimulus and sensation must be on ratio scales
b)  The psychophysical function must be continuous
c)  No dimensional constants are allowed in the formula
d)  K(k) must be a power function

## Mathematical Exercises

**Q7.** Starting from Luce's functional equation ψ(kx) = K(k) · ψ(x), derive K(k) for the power function ψ(x) = x\^β.

*Hint: Substitute ψ(x) = x\^β into both sides and simplify.*

**Q8.** Show that the logarithmic function ψ(x) = log(x) does NOT satisfy Luce's principle.

*Hint: Try to find K(k) = ψ(kx)/ψ(x) for different values of x. Is K(k) constant across different x values?*

**Q9.** Prove that when dimensional constants are allowed, ANY function ψ in the form s = s₀ · ψ(x/x₀) is scale-invariant.

*Hint: Show that when you rescale x → kx, you must also rescale x₀ → kx₀, and the ratio x/x₀ remains unchanged.*

## Coding Exercises

**Q10.** Write R code to verify that ψ(x) = x\^β satisfies Luce's principle for β = 0.5, 1, 2, and 3.

For each β, show that K(k) = ψ(kx)/ψ(x) gives the same value for different x values.

```{r}
# Starter code
k <- 3
x_values <- c(2, 4, 8)


# Your code here: compute K(k) for different beta values
beta_values <- c(0.5, 1.2, 3)


K <- function(x, beta, k) (k*x)^beta / x^beta
answer_matrix <- matrix(nrow = length(x_values), ncol = length(beta_values))
for (i in 1:length(x_values)) {
  for (j in 1:length(beta_values)) {
    answer_matrix[i,j] <- K(x_values[i], beta_values[j], k)
  }
}

answer_matrix # Each row (x) in the same columnn (beta) is the same
```

**Q11.** Write R code to demonstrate that ψ(x) = log(x) fails Luce's principle.

Compute K(k) = log(kx)/log(x) for k = 3 and x = 2, 4, 8. Show that K(k) depends on x.

```{r}
# Starter code
k <- 3
x_values <- c(2, 4, 8)

# Your code here
for (x in x_values) {
  K_value <-log(k * x) / log(x)
  cat("x =", x, ": K(k) =", K_value, "\n")
}
```

**Q12.** Write R code to demonstrate that dimensional constants make the ratio x/x₀ scale-invariant.

```{r}
# Starter code
x <- 10
x0 <- 2
k <- 5

# Your code here: show (k*x)/(k*x0) == x/x0
(k*x) / (k*x0) == x / x0
```

## Applied Problems

**Q13.** A researcher proposes the psychophysical function ψ(x) = √x (square root).

a)  Does this function satisfy Luce's principle (without dimensional constants)?
b)  What is K(k) for this function?
c)  If we rewrite it with a dimensional constant as s = s₀ · √(x/x₀), does it satisfy scale invariance? Why?

**Q14.** Consider the order effects example from the preface. Classical logic says A ∧ B ≡ B ∧ A, but empirically the order of presentation matters.

a)  What is the hidden assumption that leads to the apparent "failure" of logic?
b)  How does positional labeling (A₁, A₂, B₁, B₂) resolve this?
c)  Give another example from psychology where order might matter but the underlying mathematical structure is preserved.
