---
title: "Ideal Observer Model Example: Crowding Superiority"
subtitle: "Implementing Cicchini, Chopin, & Burr (2022) — Equations 1-19"
date: 2026-02-09
format: html
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

Visual crowding — the inability to identify peripheral targets surrounded by nearby items — is traditionally viewed as a processing bottleneck. Cicchini, Chopin, & Burr (2022, *Nature Communications*) proposed a radical alternative: crowding is **optimal Bayesian integration**.

::: {.callout-important title="The Core Claim"}
The visual system doesn't blindly mix nearby items. It computes a **weighted average** of target and flanker information, where the weight depends on:

1. **Reliability** of each source (noisy target → lean more on flankers)
2. **Similarity** between items (different flankers → ignore them)

When flankers match the target, this averaging **reduces noise** — crowding actually *helps*.
:::

The paper presents two models and two fitting functions:

| Component | Equations | What it does |
|-----------|----------|--------------|
| **Ideal Observer (IO)** | Eq 1–11 | Always integrates, weight decreases smoothly with difference |
| **Causal Inference (CI)** | Eq 12–17 | Adds a gate: integrate only if items likely share a common cause |
| **Fitting functions** | Eq 18–19 | DoG for bias, Gaussian for scatter — used to fit empirical data |

::: {.callout-tip title="What We'll Build"}
By the end of this demo, you'll have working R functions for every equation in the paper, and you'll see how they produce the three key predictions: **DoG bias**, **U-shaped scatter**, and **RMSE minimum at d = 0**.
:::

```{r}
#| label: setup
library(tidyverse)
library(patchwork)

theme_set(
  theme_minimal(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(color = "grey40")
    )
)
```

------------------------------------------------------------------------

# Part I: Ideal Observer Model (Eq 1–11)

## The Model Setup

Consider a display with three items arranged radially in the periphery:

- **Target (T):** The item the observer must report (e.g., its orientation)
- **Flanker 1 (F₁) and Flanker 2 (F₂):** Neighboring items that may influence the report

Each item's representation is noisy:

- Target: perceived as $T + \epsilon_T$, where $\epsilon_T \sim \mathcal{N}(0, \sigma_T^2)$
- Flankers: perceived as $F_i + \epsilon_{F_i}$, where $\epsilon_{F_i} \sim \mathcal{N}(0, \sigma_F^2)$

::: {.callout-note title="What do σ_T and σ_F represent?"}
In the experiment, **rounded ovals** are hard to judge ($\sigma \approx 13.2°$) and **elongated ovals** are easy ($\sigma \approx 10.0°$). These are measured empirically from single-target trials, not assumed.
:::

## Equation 1: RMSE Decomposition

Total RMS error ($E$) can be decomposed into bias ($B$) and precision (scatter standard deviation: $S$), whose squares sum to give total squared error:

$$E = \sqrt{B^2 + S^2} \tag{1}$$

::: {.callout-important title="Key Equation"}
This decomposition is the foundation of the paper's analysis. Crowding affects bias and scatter differently — you can't see the full picture by measuring only one.
:::

**Concrete example:** If bias = 3° and scatter = 4°, then RMSE = $\sqrt{9 + 16}$ = 5°.

| Bias (B) | Scatter (S) | RMSE (E) |
|----------|-------------|----------|
| 0° | 10° | 10.0° |
| 3° | 4° | 5.0° |
| 5° | 5° | 7.1° |
| 10° | 0° | 10.0° |

## Equation 2: General Weighted Response

The ideal response ($R$) is a weighted combination of target and flankers:

$$R = w_1 F_1 + w_2 F_2 + (1 - w_1 - w_2) T \tag{2}$$

The weights sum to 1, ensuring the response is a proper weighted average.

## Equation 3: Equal-Weight Simplification

Since the two flankers had the same aspect ratio, they should be weighted equally ($w_1 = w_2 = w$):

$$R = w F_1 + w F_2 + (1 - 2w) T \tag{3}$$

::: {.callout-note title="Why equal weights?"}
Both flankers are at the same distance from the target and have the same shape (same σ). There's no reason to treat one differently from the other.
:::

**Parameter table:**

| Symbol | Meaning | Range |
|--------|---------|-------|
| $w$ | Weight per flanker | 0 to ~0.33 |
| $1 - 2w$ | Weight on target | ~0.33 to 1 |
| $T$ | Target orientation | Degrees |
| $F_1, F_2$ | Flanker orientations | Degrees |

## Equations 4–7: Deriving the Bias

**Equation 4** — Mean response:

$$\mu_R = w(\mu_1 + \mu_2) + (1 - 2w)\mu_T \tag{4}$$

**Equation 5** — Bias = mean response minus true target:

$$B = \mu_R - x_T = w(\mu_1 + \mu_2 - 2\mu_T) \tag{5}$$

**Equation 6** — Define $d$ as the average flanker–target difference:

$$d = \frac{\mu_1 + \mu_2}{2} - \mu_T \tag{6}$$

**Equation 7** — Bias in terms of $d$:

$$B = 2wd \tag{7}$$

::: {.callout-note title="Why is bias = 2wd?"}
Each flanker pulls the response by $w \cdot d$ (weight times difference). Two flankers → $2wd$. More weight → more pull. Larger difference → more pull. But since $w$ decreases with $d$ (Eq 10), the product $2wd$ traces a **derivative-of-Gaussian** shape.
:::

**Concrete example:** If $w = 0.25$ and $d = 10°$, then bias = $2 \times 0.25 \times 10 = 5°$. The response is pulled 5° toward the flankers.

```{r}
#| label: bias-function
compute_bias <- function(w, d) {
  2 * w * d
}
```

## Equation 8: Scatter

Since $R$ is a weighted sum of independent random variables, its variance follows $\text{Var}(aX + bY) = a^2\text{Var}(X) + b^2\text{Var}(Y)$:

$$S^2 = 2w^2 \sigma_F^2 + (1 - 2w)^2 \sigma_T^2 \tag{8}$$

::: {.callout-note title="Why are the coefficients squared?"}
Variance measures **squared** deviations. When you multiply a random variable by a constant $a$, the variance is multiplied by $a^2$, not $a$. This is why $w$ becomes $w^2$ and $(1-2w)$ becomes $(1-2w)^2$.
:::

**Concrete example:** If $w = 0.25$, $\sigma_T = 13°$, $\sigma_F = 10°$:

- Flanker contribution: $2 \times 0.25^2 \times 100 = 12.5$
- Target contribution: $(1 - 0.5)^2 \times 169 = 42.25$
- Total: $S = \sqrt{12.5 + 42.25} = \sqrt{54.75} = 7.4°$

Compare to no flankers ($w = 0$): $S = \sigma_T = 13°$. The flankers **reduced scatter from 13° to 7.4°**!

```{r}
#| label: scatter-function
compute_scatter <- function(w, sigma_T, sigma_F) {
  sqrt(2 * w^2 * sigma_F^2 + (1 - 2 * w)^2 * sigma_T^2)
}
```

## Equations 9–10: Optimal Weight

**Equation 9** — RMSE² expanded:

$$E^2 = 4w^2 d^2 + 2w^2 \sigma_F^2 + (1 - 2w)^2 \sigma_T^2 \tag{9}$$

```{r}
#| label: total-error-sq
total_error_sq <- function(w, d, sigma_T, sigma_F) {
  4 * w^2 * d^2 + 2 * w^2 * sigma_F^2 + (1 - 4*w + 4*w^2) * sigma_T^2
}
```

**Equation 10** — Minimizing $E^2$ by setting $dE^2/dw = 0$:

$$w_{opt} = \frac{\sigma_T^2}{2\sigma_T^2 + \sigma_F^2 + 2d^2} \tag{10}$$

::: {.callout-important title="Key Equation: Optimal Weight"}
This is the heart of the model. The weight balances three factors:

| Factor | Effect on $w$ | Intuition |
|--------|--------------|-----------|
| Large $\sigma_T$ (unreliable target) | $w$ increases | Lean more on flankers |
| Large $\sigma_F$ (unreliable flankers) | $w$ decreases | Flankers aren't helpful |
| Large $d$ (different flankers) | $w$ decreases | Flankers carry irrelevant info |
:::

**Concrete examples:**

```{r}
#| label: optimal-weight
optimal_weight <- function(sigma_T, sigma_F, d) {
  sigma_T^2 / (2 * sigma_T^2 + sigma_F^2 + 2 * d^2)
}
```

```{r}
#| label: weight-examples
examples <- tribble(
  ~sigma_T, ~sigma_F, ~d,
  13.2,     10.0,     0,
  13.2,     10.0,     15,
  13.2,     10.0,     45,
  10.0,     13.2,     0,
  10.0,     10.0,     0
)

examples <- examples %>%
  mutate(
    w = optimal_weight(sigma_T, sigma_F, d),
    target_weight = 1 - 2 * w,
    description = case_when(
      sigma_T == 13.2 & d == 0  ~ "Unreliable T, identical flankers",
      sigma_T == 13.2 & d == 15 ~ "Unreliable T, moderate d",
      sigma_T == 13.2 & d == 45 ~ "Unreliable T, very different",
      sigma_T == 10.0 & sigma_F == 13.2 ~ "Reliable T, unreliable F",
      sigma_T == 10.0 & sigma_F == 10.0 ~ "Equal reliability"
    )
  )

knitr::kable(
  examples %>% select(description, sigma_T, sigma_F, d, w, target_weight),
  digits = 3,
  col.names = c("Condition", "σ_T", "σ_F", "d", "w (per flanker)", "1-2w (target)")
)
```

## Equation 11: Regression to the Mean

The paper also found regression to the mean (responses pulled toward 45°). The full response equation adds a regression factor $\rho$ and a sub-optimality scaling $\alpha$:

$$R = \alpha \rho (w F_1 + w F_2 + (1 - 2w) T) + (1 - \rho) \cdot 45 \tag{11}$$

::: {.callout-warning title="When to use Eq 11 vs Eq 3"}
Eq 11 is for **fitting real data** (with regression to the mean and sub-optimal behaviour). In our simulations below, we use the **pure model** (Eq 3 + 10) to show the core predictions without these corrections.
:::

## RMSE Function

```{r}
#| label: rmse-function
compute_rmse <- function(bias, scatter) {
  sqrt(bias^2 + scatter^2)
}
```

------------------------------------------------------------------------

## Simulation 1: Bias (Derivative-of-Gaussian)

The bias should follow a **DoG pattern**: zero at $d = 0$, peaking at intermediate $d$, returning toward zero at large $d$.

```{r}
#| label: sim1-bias
#| fig-width: 8
#| fig-height: 5

sigma_T <- 15
sigma_F <- 10
d_range <- seq(-90, 90, by = 1)

sim1 <- tibble(
  d = d_range,
  w = optimal_weight(sigma_T, sigma_F, d_range),
  bias = compute_bias(w, d_range)
)

ggplot(sim1, aes(x = d, y = bias)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(linewidth = 1.2, color = "#DC143C") +
  labs(
    title = "Simulation 1: Bias (Derivative-of-Gaussian Pattern)",
    subtitle = bquote(sigma[T] == .(sigma_T) * "," ~ sigma[F] == .(sigma_F)),
    x = "Target-Flanker Difference (d)",
    y = "Bias (degrees)"
  ) +
  scale_x_continuous(breaks = seq(-90, 90, 30))
```

The S-shaped curve shows **assimilative bias**. At intermediate differences (~20–30°), responses are maximally pulled toward the flankers. At very large differences, the flankers are too dissimilar to integrate, so bias drops back toward zero.

## Simulation 2: Scatter (U-Shape)

Scatter should show a **U-shape**: minimum at $d = 0$ (identical flankers provide the most useful additional information).

```{r}
#| label: sim2-scatter
#| fig-width: 8
#| fig-height: 5

sim2 <- tibble(
  d = d_range,
  w = optimal_weight(sigma_T, sigma_F, d_range),
  scatter = compute_scatter(w, sigma_T, sigma_F)
)

ggplot(sim2, aes(x = d, y = scatter)) +
  geom_hline(yintercept = sigma_T, linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(linewidth = 1.2, color = "#2E8B57") +
  labs(
    title = "Simulation 2: Scatter (U-Shaped Precision Improvement)",
    subtitle = bquote(sigma[T] == .(sigma_T) * "," ~ sigma[F] == .(sigma_F) ~
                      "| Dotted line = no-crowding baseline"),
    x = "Target-Flanker Difference (d)",
    y = "Scatter (degrees)"
  ) +
  scale_x_continuous(breaks = seq(-90, 90, 30)) +
  annotate("text", x = 0, y = min(sim2$scatter) - 0.3,
           label = paste0("Min scatter: ", round(min(sim2$scatter), 1), "\u00b0"),
           hjust = 0.5, vjust = 1, size = 3.5, color = "#2E8B57", fontface = "bold")
```

::: {.callout-tip title="The 'Crowding Superiority' Prediction"}
When flankers are identical to the target ($d = 0$), scatter drops **below** the no-crowding baseline ($\sigma_T$). The visual system effectively uses the flankers as additional, independent measurements of the same thing.
:::

## Simulation 3: RMSE

RMSE combines bias and scatter. Despite bias increasing at intermediate differences, the scatter reduction at $d = 0$ produces a minimum in total error.

```{r}
#| label: sim3-rmse
#| fig-width: 8
#| fig-height: 5

sim3 <- tibble(
  d = d_range,
  w = optimal_weight(sigma_T, sigma_F, d_range),
  bias = compute_bias(w, d_range),
  scatter = compute_scatter(w, sigma_T, sigma_F),
  rmse = compute_rmse(bias, scatter)
)

ggplot(sim3, aes(x = d, y = rmse)) +
  geom_hline(yintercept = sigma_T, linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(linewidth = 1.2, color = "#4169E1") +
  labs(
    title = "Simulation 3: Total Error (RMSE)",
    subtitle = bquote(sigma[T] == .(sigma_T) * "," ~ sigma[F] == .(sigma_F) ~
                      "| Dotted line = no-crowding baseline"),
    x = "Target-Flanker Difference (d)",
    y = "RMSE (degrees)"
  ) +
  scale_x_continuous(breaks = seq(-90, 90, 30)) +
  annotate("text", x = 0, y = min(sim3$rmse) - 0.3,
           label = paste0("Min RMSE: ", round(min(sim3$rmse), 1), "\u00b0"),
           hjust = 0.5, vjust = 1, size = 3.5, color = "#4169E1", fontface = "bold")
```

Total error is lowest when $d = 0$. This is the **crowding superiority effect** — having identical flankers actually helps perception.

## Simulation 4: Effect of Target Reliability

A key prediction: the integration benefit should be **stronger** for unreliable targets. With a noisy target representation, there is more to gain from incorporating flanker information.

```{r}
#| label: sim4-reliability
#| fig-width: 10
#| fig-height: 8

sigma_F_fixed <- 10

reliability_sim <- expand_grid(
  d = d_range,
  sigma_T = c(8, 15, 25),
  sigma_F = sigma_F_fixed
) %>%
  mutate(
    reliability_desc = factor(
      case_when(
        sigma_T == 8  ~ "High reliability (easy)",
        sigma_T == 15 ~ "Medium reliability",
        sigma_T == 25 ~ "Low reliability (hard)"
      ),
      levels = c("High reliability (easy)",
                 "Medium reliability",
                 "Low reliability (hard)")
    ),
    w = optimal_weight(sigma_T, sigma_F, d),
    bias = compute_bias(w, d),
    scatter = compute_scatter(w, sigma_T, sigma_F),
    rmse = compute_rmse(bias, scatter)
  )

plot_data <- reliability_sim %>%
  pivot_longer(
    cols = c(bias, scatter, rmse),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(metric,
      levels = c("bias", "scatter", "rmse"),
      labels = c("Bias", "Scatter", "RMSE")
    )
  )

ggplot(plot_data, aes(x = d, y = value, color = reliability_desc)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
  geom_line(linewidth = 1) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  scale_color_manual(values = c("#2E8B57", "#4169E1", "#DC143C")) +
  labs(
    title = "Simulation 4: Effect of Target Reliability on Crowding",
    subtitle = bquote(sigma[F] == .(sigma_F_fixed) ~ "(fixed) | Three levels of target noise"),
    x = "Target-Flanker Difference (d)",
    y = "Value (degrees)",
    color = "Target Reliability"
  ) +
  scale_x_continuous(breaks = seq(-90, 90, 30)) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 13)
  )
```

- **Low reliability** (red, $\sigma_T = 25$): Largest bias, deepest scatter dip, biggest RMSE benefit
- **High reliability** (green, $\sigma_T = 8$): Minimal bias, minimal scatter change, little RMSE benefit

This matches the Bayesian interpretation: you lean on priors (flankers) more when your data (target) is noisy.

## Weight Function

```{r}
#| label: weight-surface
#| fig-width: 8
#| fig-height: 5

weight_data <- expand_grid(
  d = seq(0, 90, by = 1),
  sigma_T = c(8, 12, 15, 20, 25)
) %>%
  mutate(
    w = optimal_weight(sigma_T, sigma_F = 10, d)
  )

ggplot(weight_data, aes(x = d, y = w, color = factor(sigma_T))) +
  geom_line(linewidth = 1.2) +
  scale_color_viridis_d(option = "C", end = 0.85, direction = -1) +
  labs(
    title = "Optimal Integration Weight (Eq. 10)",
    subtitle = "How much should the observer rely on each flanker?",
    x = "Target-Flanker Difference |d| (degrees)",
    y = "Weight per flanker (w)",
    color = expression(sigma[T])
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  theme(legend.position = "right")
```

- Higher $\sigma_T$ (noisier target) → more weight on flankers across all differences
- Weight drops off with difference — dissimilar flankers are down-weighted
- At $d = 0$: weight is purely determined by the noise ratio $\sigma_T^2 / (2\sigma_T^2 + \sigma_F^2)$

## Combined Summary: Three Key Predictions

```{r}
#| label: summary-plot
#| fig-width: 10
#| fig-height: 4

summary_data <- sim3 %>%
  select(d, bias, scatter, rmse) %>%
  pivot_longer(-d, names_to = "metric", values_to = "value") %>%
  mutate(
    metric = factor(metric,
      levels = c("bias", "scatter", "rmse"),
      labels = c("Bias (DoG)", "Scatter (U-shape)", "RMSE (minimum at d=0)")
    )
  )

colors <- c("Bias (DoG)" = "#DC143C",
            "Scatter (U-shape)" = "#2E8B57",
            "RMSE (minimum at d=0)" = "#4169E1")

ggplot(summary_data, aes(x = d, y = value, color = metric)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(linewidth = 1.2) +
  facet_wrap(~metric, scales = "free_y") +
  scale_color_manual(values = colors) +
  labs(
    title = "Ideal Observer Model: Three Key Predictions",
    subtitle = bquote(sigma[T] == .(sigma_T) * "," ~ sigma[F] == .(sigma_F)),
    x = "Target-Flanker Difference (d)",
    y = "Degrees"
  ) +
  scale_x_continuous(breaks = seq(-90, 90, 45)) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 11))
```

::: {.callout-tip title="Part I Summary"}
| Equation | Function | What it computes |
|----------|----------|-----------------|
| Eq 7 | `compute_bias(w, d)` | Bias = 2wd |
| Eq 8 | `compute_scatter(w, σ_T, σ_F)` | Scatter from weighted variance |
| Eq 1 | `compute_rmse(bias, scatter)` | Total error = √(B² + S²) |
| Eq 10 | `optimal_weight(σ_T, σ_F, d)` | Weight that minimizes RMSE |
:::

------------------------------------------------------------------------

# Part II: Causal Inference Model (Eq 12–17)

The basic ideal observer **always** integrates — weight $w$ decreases smoothly with $d$, but never reaches zero. There is no mechanism to decide *whether* to integrate. The Causal Inference (CI) model adds a **gating step**: integrate only when target and flankers likely share a common cause.

## Equation 12: Maximum Likelihood Weight

When two sources are known to come from the same cause, the optimal weight for source $A$ is:

$$w^{max}_A = \frac{\sigma_B^2}{\sigma_A^2 + \sigma_B^2} \tag{12}$$

::: {.callout-note title="How does this differ from Eq 10?"}
Eq 10 weights one flanker among three items (target + 2 flankers) and includes a $d^2$ term. Eq 12 is the classic **two-source** MLE weight — no $d^2$ because the common cause assumption guarantees $d = 0$.
:::

```{r}
#| label: mle-weight
mle_weight <- function(sigma_A, sigma_B) {
  sigma_B^2 / (sigma_A^2 + sigma_B^2)
}
```

**Concrete example:** If $\sigma_T = 13°$ and $\sigma_F = 10°$, then $w_{max} = 100 / (169 + 100) = 0.372$. The more reliable flanker gets 37% weight.

## Equations 13–14: Common Cause Probability

**Equation 13** — Full Bayesian with prior $P(S) = \mathcal{N}(\mu_P, \sigma_P^2)$:

$$P(A, B \mid C = 1) \propto \exp\!\left(-\frac{1}{2} \cdot \frac{(\mu_A - \mu_B)^2 \sigma_P^2 + (\mu_A - \mu_P)^2 \sigma_B^2 + (\mu_B - \mu_P)^2 \sigma_A^2}{\sigma_A^2 \sigma_B^2 + \sigma_A^2 \sigma_P^2 + \sigma_B^2 \sigma_P^2}\right) \tag{13}$$

The three numerator terms penalize different mismatches:

| Term | Meaning |
|------|---------|
| $(\mu_A - \mu_B)^2 \sigma_P^2$ | A and B differ from **each other** |
| $(\mu_A - \mu_P)^2 \sigma_B^2$ | A differs from the **prior** |
| $(\mu_B - \mu_P)^2 \sigma_A^2$ | B differs from the **prior** |

**Equation 14** — Simplified with flat prior ($\sigma_P^2 \to \infty$):

$$P(A, B \mid C = 1) \propto \exp\!\left(-\frac{(\mu_A - \mu_B)^2}{2(\sigma_A^2 + \sigma_B^2)}\right) \tag{14}$$

::: {.callout-note title="Why does a flat prior simplify things?"}
When $\sigma_P^2 \to \infty$, it dominates in both numerator and denominator. The first term $(\mu_A - \mu_B)^2 \sigma_P^2$ dominates the numerator; $\sigma_P^2(\sigma_A^2 + \sigma_B^2)$ dominates the denominator. They cancel, leaving only: how different are A and B from each other, relative to their combined noise?
:::

```{r}
#| label: common-cause-prob
prob_common_cause <- function(d, sigma_A, sigma_B) {
  exp(-d^2 / (2 * (sigma_A^2 + sigma_B^2)))
}
```

**Concrete example:** If $\sigma_T = 13.2°$, $\sigma_F = 10°$, and $d = 0$: P = $\exp(0) = 1$. If $d = 20°$: P = $\exp(-400 / (2 \times 274.24)) = \exp(-0.73) = 0.48$. At $d = 45°$: P = $\exp(-2025/548.48) = \exp(-3.69) = 0.025$.

### Visualization: The Causal Inference Gate

```{r}
#| label: ci-gate
#| fig-width: 8
#| fig-height: 5

d_ci <- seq(-90, 90, by = 1)
sigma_T_ci <- 13.2
sigma_F_ci <- 10.0

ci_gate <- tibble(
  d = d_ci,
  p_cc = prob_common_cause(d_ci, sigma_T_ci, sigma_F_ci)
)

ggplot(ci_gate, aes(x = d, y = p_cc)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  geom_hline(yintercept = 1, linetype = "dotted", color = "grey50") +
  geom_line(linewidth = 1.3, color = "#E65100") +
  labs(
    title = "P(common cause | d) \u2014 The Causal Inference Gate (Eq. 14)",
    subtitle = bquote(sigma[T] == .(sigma_T_ci) * "," ~ sigma[F] == .(sigma_F_ci)),
    x = "Target-Flanker Difference (d)",
    y = "P(common cause)"
  ) +
  scale_x_continuous(breaks = seq(-90, 90, 30)) +
  scale_y_continuous(limits = c(-0.05, 1.15), breaks = seq(0, 1, 0.25))
```

When $d = 0$ (identical items): $P \approx 1$ → integrate fully. When $d$ is large: $P \approx 0$ → do not integrate.

## Equation 15: CI Model Bias

In Cicchini et al.'s experiment, the two sources are the **flanker compound** ($\sigma_C^2 = \sigma_F^2/2$) and the **target**. Combining Eqs 12 and 14:

$$B = \frac{\sigma_T^2}{\sigma_C^2 + \sigma_T^2} \cdot \exp\!\left(-\frac{d^2}{2(\sigma_C^2 + \sigma_T^2)}\right) \cdot d \tag{15}$$

```{r}
#| label: ci-bias-function
ci_bias <- function(d, sigma_T, sigma_F) {
  sigma_C2 <- sigma_F^2 / 2
  w_max <- sigma_T^2 / (sigma_C2 + sigma_T^2)
  p_cc <- exp(-d^2 / (2 * (sigma_C2 + sigma_T^2)))
  w_max * p_cc * d
}
```

## Equation 16: CI Bias with Free Parameters

For fitting real data, two free parameters ($\beta$ for amplitude, $\gamma$ for breadth) plus regression factor $\rho = 0.67$:

$$B = \beta \rho \cdot \frac{\sigma_T^2}{\sigma_C^2 + \sigma_T^2} \cdot \exp\!\left(-\frac{d^2}{2\gamma^2(\sigma_C^2 + \sigma_T^2)}\right) \cdot d \tag{16}$$

```{r}
#| label: ci-bias-full
ci_bias_full <- function(d, sigma_T, sigma_F, beta, rho, gamma) {
  sigma_C2 <- sigma_F^2 / 2
  w_ratio <- sigma_T^2 / (sigma_C2 + sigma_T^2)
  p_cc <- exp(-d^2 / (2 * gamma^2 * (sigma_C2 + sigma_T^2)))
  beta * rho * w_ratio * p_cc * d
}
```

## Equation 17: Product of Two Gaussians

The product of two Gaussians with different means has a peak at:

$$\frac{1}{2\pi \sigma_A \sigma_B} \exp\!\left(-\frac{(\mu_A - \mu_B)^2}{2(\sigma_A^2 + \sigma_B^2)}\right) \tag{17}$$

This embeds the same Gaussian dependency on $|\mu_A - \mu_B|$ as Eq 14 — the same gating behaviour emerges whether you compute the common-cause probability (Bayesian) or the distribution overlap (product).

## IO vs CI Model Comparison

```{r}
#| label: io-vs-ci
#| fig-width: 12
#| fig-height: 5

d_comp <- seq(-45, 45, by = 0.5)

comparison <- tibble(
  d = rep(d_comp, 2),
  model = rep(c("Ideal Observer (Eq. 10)", "Causal Inference (Eq. 15)"), each = length(d_comp)),
  bias = c(
    compute_bias(optimal_weight(sigma_T_ci, sigma_F_ci, d_comp), d_comp),
    ci_bias(d_comp, sigma_T_ci, sigma_F_ci)
  ),
  weight = c(
    optimal_weight(sigma_T_ci, sigma_F_ci, d_comp),
    {
      sigma_C2 <- sigma_F_ci^2 / 2
      w_max <- sigma_T_ci^2 / (sigma_C2 + sigma_T_ci^2)
      p_cc <- exp(-d_comp^2 / (2 * (sigma_C2 + sigma_T_ci^2)))
      w_max * p_cc
    }
  )
)

p_bias <- ggplot(comparison, aes(x = d, y = bias, color = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Ideal Observer (Eq. 10)" = "#1565C0",
                                "Causal Inference (Eq. 15)" = "#E65100")) +
  labs(title = "Bias Comparison",
       x = "Target-Flanker Difference (d)", y = "Bias") +
  theme(legend.position = "none")

p_weight <- ggplot(comparison, aes(x = d, y = weight, color = model)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Ideal Observer (Eq. 10)" = "#1565C0",
                                "Causal Inference (Eq. 15)" = "#E65100")) +
  labs(title = "Weight Comparison",
       x = "Target-Flanker Difference (d)", y = "Weight (w)") +
  theme(legend.position = "bottom", legend.title = element_blank())

p_bias + p_weight +
  plot_annotation(
    title = "Ideal Observer vs Causal Inference Model",
    subtitle = bquote(sigma[T] == .(sigma_T_ci) * "," ~ sigma[F] == .(sigma_F_ci))
  )
```

**Key differences:**

- **IO model** (blue): Weight is always $> 0$ — it decays smoothly but never reaches zero
- **CI model** (orange): Weight is gated to zero at large $d$ via the Gaussian gate
- **At $d = 0$**: Both models predict similar behaviour — maximum integration weight and zero bias

::: {.callout-tip title="Part II Summary"}
| Equation | Function | What it computes |
|----------|----------|-----------------|
| Eq 12 | `mle_weight(σ_A, σ_B)` | MLE weight for common-cause fusion |
| Eq 14 | `prob_common_cause(d, σ_A, σ_B)` | Gate: should we integrate? |
| Eq 15 | `ci_bias(d, σ_T, σ_F)` | CI model bias (weight × gate × d) |
| Eq 16 | `ci_bias_full(...)` | CI bias with free parameters for fitting |
:::

------------------------------------------------------------------------

# Part III: Model Fitting Functions (Eq 18–19)

Cicchini et al. fit their empirical data using a derivative-of-Gaussian for bias and a Gaussian for scatter.

## Equation 18: Derivative-of-Gaussian Fit for Bias

Bias was fitted by a DoG — a Gaussian of width $s$ multiplied by a straight line of slope $a$:

$$B = a \cdot (\theta - m) \cdot \exp\!\left(-\frac{(\theta - m)^2}{s^2}\right) + b \tag{18}$$

| Parameter | Meaning |
|-----------|---------|
| $a$ | Weight/amplitude (slope of linear component) |
| $m$ | Centre of the DoG |
| $s$ | Width of the Gaussian window |
| $b$ | Vertical offset |

```{r}
#| label: dog-fit-function
dog_fit <- function(theta, a, m, s, b) {
  a * (theta - m) * exp(-(theta - m)^2 / s^2) + b
}
```

## Equation 19: Gaussian Fit for Scatter

$$S = a \cdot \exp\!\left(-\frac{(\theta - m)^2}{s^2}\right) + b \tag{19}$$

| Parameter | Meaning |
|-----------|---------|
| $a$ | Amplitude of the Gaussian dip (negative for U-shape) |
| $m$ | Centre |
| $s$ | Width (**yoked** to Eq 18) |
| $b$ | Baseline scatter at large differences |

::: {.callout-note title="Why yoke the width parameter?"}
Since bias and scatter likely originate from the same integration process, the width $s$ is shared between Eq 18 and 19. This reduces free parameters and ensures both curves reflect the same integration window.
:::

```{r}
#| label: gaussian-scatter-function
gaussian_scatter_fit <- function(theta, a, m, s, b) {
  a * exp(-(theta - m)^2 / s^2) + b
}
```

## Fitting Demonstration

We generate synthetic data from the IO model, add noise, and show how the DoG and Gaussian functions capture the underlying patterns.

```{r}
#| label: fitting-demo
#| fig-width: 12
#| fig-height: 5

set.seed(42)
d_fit <- seq(-45, 45, by = 5)
sigma_T_fit <- 13.2
sigma_F_fit <- 10.0

w_fit <- optimal_weight(sigma_T_fit, sigma_F_fit, d_fit)
true_bias <- compute_bias(w_fit, d_fit)
true_scatter <- compute_scatter(w_fit, sigma_T_fit, sigma_F_fit)

observed_bias <- true_bias + rnorm(length(d_fit), 0, 0.5)
observed_scatter <- true_scatter + rnorm(length(d_fit), 0, 0.3)

df_obs <- data.frame(d = d_fit, bias = observed_bias, scatter = observed_scatter)

nls_bias <- nls(bias ~ a * (d - m) * exp(-(d - m)^2 / s^2) + b,
                data = df_obs, start = list(a = 0.3, m = 0, s = 20, b = 0))

nls_scatter <- nls(scatter ~ a * exp(-(d - m)^2 / s^2) + b,
                   data = df_obs, start = list(a = -4, m = 0, s = 20, b = 13))

theta_smooth <- seq(-45, 45, by = 0.5)
coef_b <- coef(nls_bias)
coef_s <- coef(nls_scatter)
fit_bias <- dog_fit(theta_smooth, coef_b["a"], coef_b["m"], coef_b["s"], coef_b["b"])
fit_scatter <- gaussian_scatter_fit(theta_smooth, coef_s["a"], coef_s["m"],
                                    coef_s["s"], coef_s["b"])

p_fit_bias <- ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(aes(x = d_fit, y = observed_bias), size = 2.5, color = "#DC143C") +
  geom_line(aes(x = theta_smooth, y = fit_bias), linewidth = 1.2,
            color = "#DC143C", alpha = 0.6) +
  labs(title = "Bias: DoG Fit (Eq. 18)",
       x = "Target-Flanker Difference (d)", y = "Bias (degrees)") +
  scale_x_continuous(breaks = seq(-45, 45, 15))

p_fit_scatter <- ggplot() +
  geom_hline(yintercept = sigma_T_fit, linetype = "dotted", color = "grey50") +
  geom_point(aes(x = d_fit, y = observed_scatter), size = 2.5, color = "#2E8B57") +
  geom_line(aes(x = theta_smooth, y = fit_scatter), linewidth = 1.2,
            color = "#2E8B57", alpha = 0.6) +
  labs(title = "Scatter: Gaussian Fit (Eq. 19)",
       x = "Target-Flanker Difference (d)", y = "Scatter (degrees)") +
  scale_x_continuous(breaks = seq(-45, 45, 15))

p_fit_bias + p_fit_scatter +
  plot_annotation(
    title = "Model Fitting Functions (Eq. 18\u201319)",
    subtitle = "Synthetic data from IO model + noise, fitted with DoG and Gaussian"
  )
```

The DoG captures the S-shaped bias pattern, while the inverted Gaussian captures the U-shaped scatter dip.

::: {.callout-tip title="Part III Summary"}
| Equation | Function | What it fits |
|----------|----------|-------------|
| Eq 18 | `dog_fit(θ, a, m, s, b)` | Bias data (DoG) |
| Eq 19 | `gaussian_scatter_fit(θ, a, m, s, b)` | Scatter data (inverted Gaussian) |
:::

------------------------------------------------------------------------

# Summary

## All Functions at a Glance

| Equation | R Function | Purpose |
|----------|-----------|---------|
| Eq 1 | `compute_rmse(bias, scatter)` | Total error decomposition |
| Eq 7 | `compute_bias(w, d)` | Predicted bias |
| Eq 8 | `compute_scatter(w, σ_T, σ_F)` | Predicted scatter |
| Eq 9 | `total_error_sq(w, d, σ_T, σ_F)` | RMSE² (for optimization) |
| Eq 10 | `optimal_weight(σ_T, σ_F, d)` | Optimal flanker weight |
| Eq 12 | `mle_weight(σ_A, σ_B)` | MLE weight (common cause) |
| Eq 14 | `prob_common_cause(d, σ_A, σ_B)` | Causal inference gate |
| Eq 15 | `ci_bias(d, σ_T, σ_F)` | CI model bias |
| Eq 16 | `ci_bias_full(...)` | CI bias with free parameters |
| Eq 18 | `dog_fit(θ, a, m, s, b)` | DoG fitting function |
| Eq 19 | `gaussian_scatter_fit(θ, a, m, s, b)` | Gaussian fitting function |

## Key Parameter Effects

| Parameter | Increase → | Bias effect | Scatter effect |
|-----------|-----------|-------------|----------------|
| $\sigma_T$ | Noisier target | Larger DoG amplitude | Deeper U-shape dip |
| $\sigma_F$ | Noisier flankers | Smaller DoG | Shallower dip |
| $d$ | More different flankers | Rise then fall (DoG) | Rise (U-shape) |

::: {.callout-important title="The Three Predictions"}
1. **Bias** follows a derivative-of-Gaussian: zero at d = 0, peaks at intermediate d, returns to zero
2. **Scatter** follows a U-shape: minimum at d = 0 (below unflanked baseline!)
3. **RMSE** has minimum at d = 0: identical flankers improve total accuracy
:::
