---
title: "Mathematical Psychology Foundations"
subtitle: "New Handbook of Mathematical Psychology Vol.3 - Preface"
date: 2026-01-21
format: html
execute:
  echo: true
---

```{r setup}
#| echo: false
#| message: false
library(ggplot2)
```

Mathematics in psychology is **not** about computing (adding, multiplying, etc.). It's about **conceptual clarity** and avoiding conceptual confusions.

Before we can compute, we need to explicate our hidden assumptions. Often when we do this, we find they are not that compelling.

## Measurement Scales

Entities are measured on specific scales: ordinal, interval, ratio.

For **ratio scales**, any two valid measurement functions f and g are related by:

$$f(x) = k \cdot g(x)$$

for some positive constant k (like meters vs centimeters).

```{r ratio-scale}
# Example: measuring the same stimulus in different units
stimulus <- c(1, 2, 3, 4, 5)

g_meters <- stimulus
f_centimeters <- 100 * stimulus

# Key property: RATIOS are preserved
data.frame(
  stimulus = stimulus,
  k_constant = 100,
  meters = g_meters,
  centimeters = f_centimeters,
  ratio_m = g_meters / g_meters[1],
  ratio_cm = f_centimeters / f_centimeters[1]
)
```

## Luce's Principle of Theory Construction

Luce (1959) used measurement scales to constrain psychophysical functions.

**Setup:**

-   x = stimulus magnitude (ratio scale)
-   s = sensation magnitude (ratio scale)
-   ψ = psychophysical function: s = ψ(x)

**Key idea:** If we rescale the stimulus by factor k (x → kx), the function should still work with some corresponding rescaling of sensation (s → cs).

**Concrete example with ψ(x) = x²**

Let's use a known psychophysical function and trace through the derivation.

```{r luce-simulation}
# Define an example psychophysical function: psi(x) = x^2
psi <- function(x) x^2

# Original measurements
x <- c(1, 2, 3, 4, 5)
s <- psi(x)

data.frame(x = x, s = s)
```

```{r plot-psi}
#| fig-width: 4
#| fig-height: 2.5

ggplot(data.frame(x = x, s = s), aes(x = x, y = s)) +
  geom_line() +
  geom_point() +
  labs(title = "ψ(x) = x²", x = "x (stimulus)", y = "s (sensation)") +
  theme_minimal()
```

Now suppose we change our measurement unit for x by factor k = 3 (like switching units):

```{r rescale-stimulus}
k <- 3
x_new <- k * x  # rescaled stimulus measurements

# What is s when we plug in the new x?
s_from_new_x <- psi(x_new)

data.frame(
  x_original = x,
  x_rescaled = x_new,
  s_original = s,
  s_from_rescaled = s_from_new_x
)
```

Luce's principle requires: ψ(kx) = K(k) · ψ(x)

We need to find K(k). We know ψ(kx) and ψ(x), so we can solve for K(k):

$$K(k) = \frac{\psi(kx)}{\psi(x)} = c$$which c is the constant for the s

```{r find-K}
# For x = 1:
s_from_new_x[1] / s[1]  # psi(3*1) / psi(1) = 9 / 1 = 9

# For x = 2:
s_from_new_x[2] / s[2]  # psi(3*2) / psi(2) = 36 / 4 = 9

# For x = 3:
s_from_new_x[3] / s[3]  # psi(3*3) / psi(3) = 81 / 9 = 9
```

K(k) = 9 for all x values! This is crucial - K(k) must depend only on k, not on x.

And notice: K(k) = 9 = 3² = k². So for ψ(x) = x², we get K(k) = k².

**The general pattern:** For ψ(x) = x\^β, we get K(k) = k\^β

```{r verify-pattern}
# Test with k = 3, x = 4, different beta values
k <- 3
x <- 4

# beta = 0.5
(k * x)^0.5  # psi(kx)
k^0.5 * x^0.5  # K(k) * psi(x)

# beta = 1
(k * x)^1
k^1 * x^1

# beta = 2
(k * x)^2
k^2 * x^2

# beta = 3
(k * x)^3
k^3 * x^3
```

All pairs match. The pattern holds for any β.

**Why logarithm fails:**

Let's try the same approach with ψ(x) = log(x):

```{r log-fails}
# psi(x) = log(x)
k <- 3

# Try to find K(k) = psi(kx) / psi(x) for different x values:

# For x = 2:
log(k * 2) / log(2)  

# For x = 4:
log(k * 4) / log(4)  

# For x = 8:
log(k * 8) / log(8)  
```

The ratio ψ(kx)/ψ(x) gives **different values** for different x!

For power function: K(k) = 9 for all x (constant) For log function: K(k) = 2.58, 1.79, 1.53... (depends on x)

Luce's principle requires K(k) to depend **only on k**, not on x. Log fails this requirement.

## The Hidden Assumption: Fechner's Counterexample

But Fechner's logarithmic law exists:

$$s = s_0 \cdot \log\left(\frac{x}{x_0}\right)$$

How can this be? Rozeboom (1962) found Luce's **hidden assumption**: no dimensional constants.

**What is a dimensional constant?**

A dimensional constant is a fixed value measured in the same units as your variable. For example:

-   x₀ = threshold stimulus (measured in same units as x)
-   s₀ = baseline sensation (measured in same units as s)

The key: when you rescale x by factor k, you must also rescale x₀ by k (same units!). So the ratio x/x₀ is unchanged:

```{r dimensional-constants}
x <- 4
x0 <- 1
k <- 3

# Original ratio
original <- x / x0

# After rescaling BOTH x and x0 by k
scaled <- (k * x) / (k * x0)

data.frame(
  original_ratio = original,
  scaled_ratio = scaled,
  unchanged = original == scaled
)
```

The k cancels out! The ratio x/x₀ doesn't change when you rescale.

**So what does this mean?**

Any function that uses the ratio x/x₀ is automatically scale-invariant:

$$s = s_0 \cdot \psi\left(\frac{x}{x_0}\right)$$

Here ψ can be ANY increasing function - log, power, square root, anything!

**Summary:**

-   Luce (without dimensional constants): only power functions satisfy scale invariance
-   Rozeboom (with dimensional constants): ANY function of x/x₀ satisfies scale invariance

Luce's hidden assumption was "no dimensional constants." Once you allow them, his restriction to power functions disappears.

Luce accepted this criticism and withdrew his "principle of theory construction" (Luce, 1962).

## Hidden Assumptions

This story illustrates the core message of the preface:

**Mathematics is not about computing** - it's about conceptual clarity and avoiding confusion.

Luce's derivation was mathematically correct. The algebra was flawless. But it contained a hidden assumption (no dimensional constants) that wasn't explicitly stated. When Rozeboom made this assumption explicit, the apparent restriction to power functions disappeared.

This pattern repeats throughout science:

1.  Someone makes a mathematical argument
2.  The conclusion seems surprising or restrictive
3.  Upon closer inspection, there's a hidden assumption
4.  Making it explicit resolves the puzzle

The value of mathematical thinking is not in the calculations themselves, but in forcing us to **state our assumptions clearly**. Often, once we do this, we find they are not that compelling - or we discover alternative assumptions that lead to different conclusions.
