---
title: "Initial Visual Encoding"
subtitle: "Handbook Ch.1 - Wandell & Brainard"
date: 2026-01-23
format: html
execute:
  echo: true
---

```{r setup}
#| echo: false
#| message: false
library(ggplot2)
library(tidyr)
```

## 1.1 Introduction

Vision science uses mathematics to communicate across disciplines. The initial visual encoding describes how light from a scene becomes neural signals - from physics (light) through optics (eye) to biology (photoreceptors).

::: {.callout-note title="Why Mathematics in Vision Science?"}
Vision research serves many purposes:

| Investigator        | Goal                                 |
|---------------------|--------------------------------------|
| **Medical**         | Diagnose and repair visual disorders |
| **Psychologists**   | Quantify rules of perception         |
| **Neuroscientists** | Relate experience to neural signals  |
| **Engineers**       | Design artificial vision systems     |

Mathematics provides a **common language** across all these disciplines.
:::

### Core Mathematical Methods

This chapter covers three fundamental approaches:

```{r intro-methods}
#| fig-cap: "The three mathematical pillars of vision science"
#| fig-width: 10
#| fig-height: 3

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

# 1. Linear Algebra - show a matrix transformation
set.seed(42)
x <- rnorm(50, 0, 1)
y <- rnorm(50, 0, 1)

# Apply a linear transformation (rotation + scaling)
theta <- pi/6  # 30 degree rotation
A <- matrix(c(cos(theta), sin(theta), -sin(theta), cos(theta)), 2, 2) * 1.5
transformed <- t(A %*% rbind(x, y))

plot(x, y, pch = 19, col = "blue", xlim = c(-4, 4), ylim = c(-4, 4),
     xlab = "x", ylab = "y", main = "1. Linear Algebra\n(Transformations)")
points(transformed[,1], transformed[,2], pch = 19, col = "red")
arrows(x, y, transformed[,1], transformed[,2], length = 0.05, col = "gray")
legend("topright", c("Original", "Transformed"), col = c("blue", "red"), pch = 19, cex = 0.8)

# 2. Noise Distributions - Poisson
lambda <- 100  # mean photon count
samples <- rpois(1000, lambda)

hist(samples, breaks = 30, col = "lightblue", border = "white",
     main = "2. Noise Distributions\n(Poisson photon noise)",
     xlab = "Photon count", prob = TRUE)
curve(dnorm(x, lambda, sqrt(lambda)), add = TRUE, col = "red", lwd = 2)
abline(v = lambda, lty = 2, col = "blue", lwd = 2)
text(lambda + 5, 0.035, sprintf("mean = %d", lambda), col = "blue")

# 3. Bayesian Inference - prior × likelihood = posterior
x_vals <- seq(-3, 5, length.out = 200)
prior <- dnorm(x_vals, 0, 1)
likelihood <- dnorm(x_vals, 2, 0.8)
posterior <- prior * likelihood
posterior <- posterior / sum(posterior) * length(x_vals) / 8  # normalize for plotting

plot(x_vals, prior, type = "l", col = "blue", lwd = 2,
     xlab = "Parameter value", ylab = "Density",
     main = "3. Bayesian Inference\n(Prior × Likelihood = Posterior)",
     ylim = c(0, max(c(prior, likelihood, posterior)) * 1.1))
lines(x_vals, likelihood, col = "green", lwd = 2)
lines(x_vals, posterior, col = "red", lwd = 2)
legend("topright", c("Prior", "Likelihood", "Posterior"),
       col = c("blue", "green", "red"), lwd = 2, cex = 0.8)
```

### Chapter Structure: The Forward-Inverse Framework

```{r intro-chapter-structure}
#| fig-cap: "Chapter structure: Forward calculation then inverse inference"
#| fig-width: 10
#| fig-height: 4

par(mar = c(1, 1, 2, 1))
plot(NULL, xlim = c(0, 10), ylim = c(0, 6), axes = FALSE, xlab = "", ylab = "",
     main = "Chapter Roadmap")

# Boxes
rect(0.5, 4, 2.5, 5.5, col = "lightyellow", border = "orange", lwd = 2)
rect(3.5, 4, 5.5, 5.5, col = "lightblue", border = "blue", lwd = 2)
rect(6.5, 4, 8.5, 5.5, col = "lightgreen", border = "darkgreen", lwd = 2)

# Labels
text(1.5, 4.75, "SCENE\n(Light Field)", cex = 0.9)
text(4.5, 4.75, "OPTICS\n(Eye)", cex = 0.9)
text(7.5, 4.75, "CONES\n(Excitations)", cex = 0.9)

# Forward arrows
arrows(2.5, 4.75, 3.5, 4.75, lwd = 2, col = "black")
arrows(5.5, 4.75, 6.5, 4.75, lwd = 2, col = "black")

# Section labels
text(3, 5.8, "§1.2-1.3", col = "gray40", cex = 0.8)
text(6, 5.8, "§1.4", col = "gray40", cex = 0.8)

# Part labels
text(4.5, 6, "PART 1: Forward Calculation", col = "blue", font = 2)

# Inference arrow (curved, going back)
arrows(7.5, 3.8, 1.5, 3.8, lwd = 2, col = "red", code = 1)
text(4.5, 3.5, "PART 2: Inverse Problem (§1.5)", col = "red", font = 2)
text(4.5, 3, "Given cone excitations, infer scene properties", col = "red", cex = 0.8)

# Bottom: perception
rect(3.5, 1, 5.5, 2.5, col = "lavender", border = "purple", lwd = 2)
text(4.5, 1.75, "PERCEPTION\n(Inference)", cex = 0.9)
arrows(4.5, 3.8, 4.5, 2.5, lwd = 2, col = "purple")

cat("Chapter structure:\n")
cat("  Part 1 (§1.2-1.4): Forward calculation - how scene becomes cone signals\n")
cat("  Part 2 (§1.5): Inverse problem - how brain infers scene from signals\n")
```

### Helmholtz's Key Insight (1866)

::: {.callout-important title="Unconscious Inference"}
> "The general rule determining the ideas of vision... is that such objects are always imagined as being present in the field of vision as **would have to be there in order to produce the same impression on the nervous mechanism**." — Hermann von Helmholtz (1866)

**Translation:** We perceive what would most likely have caused our sensory signals!

This is the foundation for Bayesian models of perception.
:::

### Principles vs. Computation

Mathematical principles are essential but not sufficient:

-   **Analogy:** Newton's gravity law is simple, but predicting tides at a specific harbor requires detailed computation accounting for local geography.
-   **In vision:** "Color vision is 3D" is a principle, but predicting color matches at different retinal locations requires accounting for macular pigment, chromatic aberration, and cone density variations.

::: {.callout-tip title="Chapter Goals"}
1.  **Value both** principles (equations) AND computations (software)
2.  **Build knowledge**: develop principles → test against experiments → combine with computation
3.  **Produce models** that account for visual performance and enable engineering advances
:::

------------------------------------------------------------------------

## 1.2 Scene to Retinal Image

### Light Field

Leonardo da Vinci noticed that a pinhole camera forms an image wherever you place it - meaning light rays must be present everywhere, traveling in all directions. Gershun (1939) formalized this as the **light field**:

$$LF(x, y, z, \alpha, \beta, \lambda, \rho)$$

Each ray is described by:

-   Position: $(x, y, z)$
-   Direction: $(\alpha, \beta)$
-   Wavelength: $\lambda$
-   Polarization: $\rho$

This is a 7-dimensional function!

### Incident Light Field

The eye only captures rays entering the pupil - the **incident light field**:

$$ILF(u, v, \alpha, \beta, \lambda, \rho, t)$$

where $(u, v)$ is position at the pupil aperture and $t$ is time.

### Plenoptic Function

Adelson & Bergen (1991) simplified further for human vision. The eye focuses rays onto the retina, so direction is determined by retinal position $(r_x, r_y)$. Ignoring polarization (unimportant for perception):

$$E(r_x, r_y, \lambda, t)$$

This is the **retinal spectral irradiance** - power per unit area as a function of wavelength at each retinal location.

```{r light-field-reduction}
# The eye dramatically reduces information dimensionality
dimensions <- data.frame(
  representation = c("Light Field", "Incident Light Field", "Retinal Image"),
  dimensions = c(7, 7, 4),
  parameters = c("x,y,z,α,β,λ,ρ", "u,v,α,β,λ,ρ,t", "rx,ry,λ,t")
)
dimensions
```

The key insight: many different light fields produce the **same** retinal image. Information is inevitably lost.

### Simulating a Light Field (Toy Model)

> This is a simplified **toy model** for learning - not from a published source.

Let's make the abstract light field concrete by building a simple 2D scene and querying it. We'll work in 2D (dropping z) with one direction angle α (dropping β) to keep things visualizable.

**Our scene**: Three colored rectangles at different positions along a wall.

```{r define-scene}
# Define a simple scene: colored rectangles on a wall at y = 10
# Each object has: x_min, x_max, and a spectrum (simplified to peak wavelength)
scene <- list(
  list(x_min = -3, x_max = -1, peak_wavelength = 620),  # red object
  list(x_min = 1, x_max = 3, peak_wavelength = 530),    # green object
  list(x_min = 5, x_max = 7, peak_wavelength = 460)     # blue object
)
wall_y <- 10
```

```{r plot-scene}
#| fig-width: 7
#| fig-height: 5

# Visualize the scene from bird's eye view
scene_plot <- ggplot() +
  # Draw the wall
  geom_segment(aes(x = -5, xend = 9, y = wall_y, yend = wall_y),
               color = "gray50", linewidth = 1) +
  # Draw the colored objects on the wall
  geom_segment(aes(x = -3, xend = -1, y = wall_y, yend = wall_y),
               color = "red", linewidth = 4) +
  geom_segment(aes(x = 1, xend = 3, y = wall_y, yend = wall_y),
               color = "green3", linewidth = 4) +
  geom_segment(aes(x = 5, xend = 7, y = wall_y, yend = wall_y),
               color = "blue", linewidth = 4) +
  # Draw the observer
  geom_point(aes(x = 0, y = 0), size = 4, color = "black") +
  annotate("text", x = 0, y = -0.8, label = "Observer (0, 0)", size = 3.5) +
  # Draw rays from observer to center of each object
  geom_segment(aes(x = 0, xend = -2, y = 0, yend = wall_y),
               color = "red", linetype = "dashed", alpha = 0.6) +
  geom_segment(aes(x = 0, xend = 2, y = 0, yend = wall_y),
               color = "green3", linetype = "dashed", alpha = 0.6) +
  geom_segment(aes(x = 0, xend = 6, y = 0, yend = wall_y),
               color = "blue", linetype = "dashed", alpha = 0.6) +
  # Draw the "straight ahead" reference line
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = wall_y),
               color = "gray30", linetype = "dotted") +
  annotate("text", x = 0.3, y = 5, label = "α = 0°", size = 3, angle = 90) +
  # Label the angles
  annotate("text", x = -2, y = wall_y + 0.6, label = "α ≈ -11°",
           size = 3, color = "red") +
  annotate("text", x = 2, y = wall_y + 0.6, label = "α ≈ +11°",
           size = 3, color = "green4") +
  annotate("text", x = 6, y = wall_y + 0.6, label = "α ≈ +31°",
           size = 3, color = "blue") +
  # Formatting
  coord_fixed(ratio = 1) +
  xlim(-5, 9) + ylim(-2, 12) +
  labs(title = "Bird's Eye View of the Scene",
       subtitle = "Observer at origin, wall at y = 10, three colored objects",
       x = "x position", y = "y position") +
  theme_minimal()

scene_plot
```

**Understanding the angles:**

The angle α tells you which direction you're looking:

-   **α = 0** means looking straight ahead (along the y-axis)
-   **α \< 0** (negative) means looking LEFT
-   **α \> 0** (positive) means looking RIGHT

The angle to each object depends on geometry. From (0, 0), to hit the center of the green object at (2, 10):

$$\alpha = \arctan\left(\frac{x_{object}}{y_{wall}}\right) = \arctan\left(\frac{2}{10}\right) \approx 11.3°$$

```{r angle-calculation}
# Calculate the viewing angle to each object's center
object_centers <- c(-2, 2, 6)  # x-coordinates of object centers
angles_to_objects <- atan(object_centers / wall_y) * 180 / pi  # convert to degrees

data.frame(
  object = c("Red", "Green", "Blue"),
  center_x = object_centers,
  angle_degrees = round(angles_to_objects, 1)
)
```

**The light field function:**

```{r light-field-function}
light_field <- function(x, y, alpha, lambda) {
  if (y >= wall_y) return(0)

  # Ray trace: where does ray hit the wall?
  hit_x <- x + (wall_y - y) * tan(alpha)

  # Check if we hit an object
  for (obj in scene) {
    if (hit_x >= obj$x_min && hit_x <= obj$x_max) {
      # Return spectrum (Gaussian centered at object's peak wavelength)
      return(exp(-0.5 * ((lambda - obj$peak_wavelength) / 30)^2))
    }
  }
  return(0.05)  # empty wall
}
```

> **Reality vs Toy Model:** Real radiance varies due to light source distance, surface angle, texture, and shadows. Our toy model only uses spectrum match (Gaussian).

**Scan across directions** from (0,0):

```{r query-directions}
#| fig-width: 6
#| fig-height: 3

angles <- seq(-0.6, 0.6, length.out = 200)
results <- data.frame(
  angle = rep(angles * 180/pi, 3),
  radiance = c(
    sapply(angles, function(a) light_field(0, 0, a, 620)),
    sapply(angles, function(a) light_field(0, 0, a, 530)),
    sapply(angles, function(a) light_field(0, 0, a, 460))
  ),
  color = rep(c("red", "green", "blue"), each = 200)
)

ggplot(results, aes(angle, radiance, color = color)) +
  geom_line(linewidth = 1) +
  scale_color_identity() +
  labs(title = "Light field: scanning left to right",
       x = "Angle (degrees)", y = "Radiance") +
  theme_minimal()
```

Each peak = one object. The function answers: "Looking in direction α, how much light of wavelength λ do I see?"

### From Light Field to Cone Response

The eye compresses the light field in two steps:

1.  **Direction → Retinal position**: The lens maps viewing angle to a spot on the retina
2.  **Many wavelengths → 3 cone values**: Each cone type sums up light weighted by its sensitivity

```{r compression-step1}
#| fig-width: 5
#| fig-height: 3

directions <- seq(-0.4, 0.7, length.out = 30)
wavelengths <- seq(400, 700, by = 10)

# Pick a direction (change this to see different objects!)
alpha <- -0.1

# What spectrum arrives at this direction?
spectrum_at_alpha <- sapply(wavelengths, function(w) light_field(0, 0, alpha, w))

# Determine color based on which object we're looking at
hit_x <- (wall_y - 0) * tan(alpha)
plot_color <- if (hit_x >= -3 && hit_x <= -1) "red" else
              if (hit_x >= 1 && hit_x <= 3) "green3" else
              if (hit_x >= 5 && hit_x <= 7) "blue" else "gray50"
object_name <- if (hit_x >= -3 && hit_x <= -1) "red" else
               if (hit_x >= 1 && hit_x <= 3) "green" else
               if (hit_x >= 5 && hit_x <= 7) "blue" else "wall"

ggplot(data.frame(w = wavelengths, r = spectrum_at_alpha), aes(w, r)) +
  geom_line(color = plot_color, linewidth = 1) +
  geom_area(fill = plot_color, alpha = 0.3) +
  labs(title = paste0("Spectrum at α = ", round(alpha * 180/pi, 1), "°"),
       subtitle = paste0("This spot sees the ", object_name, " object"),
       x = "Wavelength (nm)", y = "Radiance") +
  theme_minimal()
```

```{r compression-step2}
#| fig-width: 5
#| fig-height: 3

# Cone sensitivities (different widths reflect real biology!)
# S-cone: narrow (SD=35), M-cone: medium (SD=45), L-cone: wide (SD=50)
L_sens <- exp(-0.5 * ((wavelengths - 565) / 50)^2)
M_sens <- exp(-0.5 * ((wavelengths - 535) / 45)^2)
S_sens <- exp(-0.5 * ((wavelengths - 445) / 35)^2)

# Cone response = sum(spectrum × sensitivity)
L_resp <- sum(spectrum_at_alpha * L_sens)
M_resp <- sum(spectrum_at_alpha * M_sens)
S_resp <- sum(spectrum_at_alpha * S_sens)

ggplot(data.frame(cone = c("L", "M", "S"), response = c(L_resp, M_resp, S_resp)),
       aes(cone, response, fill = cone)) +
  geom_col() +
  scale_fill_manual(values = c("L" = "red", "M" = "green", "S" = "blue")) +
  labs(title = "Cone responses at this position",
       subtitle = "31 wavelengths → 3 numbers",
       x = "", y = "Response") +
  theme_minimal() + theme(legend.position = "none")
```

M-cone responds most because green object matches M-cone sensitivity!

> **Note:** The different bandwidths (SD = 35, 45, 50) approximate real cone biology. S-cones are narrowest (most selective), L-cones are widest (respond to broader range). This comes from the molecular structure of each cone's photopigment (opsin).

```{r compression-full}
#| fig-width: 6
#| fig-height: 3

# Do this for ALL retinal positions
L_responses <- c()
M_responses <- c()
S_responses <- c()

for (a in directions) {
  # Step 1: Get spectrum at this direction
  spectrum <- c()
  for (w in wavelengths) {
    spectrum <- c(spectrum, light_field(0, 0, a, w))
  }

  # Step 2: Compute cone responses (weighted sum)
  L_responses <- c(L_responses, sum(spectrum * L_sens))
  M_responses <- c(M_responses, sum(spectrum * M_sens))
  S_responses <- c(S_responses, sum(spectrum * S_sens))
}

# Combine into data frame for plotting
cone_responses <- data.frame(
  position = rep(directions * 180/pi, 3),
  response = c(L_responses, M_responses, S_responses),
  cone = rep(c("L", "M", "S"), each = length(directions))
)

ggplot(cone_responses, aes(position, response, color = cone)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("L" = "red", "M" = "green", "S" = "blue")) +
  labs(title = "Cone responses across retina",
       x = "Retinal position (degrees)", y = "Response") +
  theme_minimal()
```

**The compression:** Light field (position × direction × wavelength) → just 3 numbers (L, M, S) per retinal spot. This is the "initial visual encoding."

## 1.3 Mathematical Principles

### Linear Systems

A system $L$ is **linear** if it follows the superposition rule:

$$L(x + y) = L(x) + L(y)$$

This implies **homogeneity**: $L(mx) = mL(x)$ for any scalar $m$.

```{r linear-system}
# Example: a linear transformation (like cone absorption)
L <- function(input) {
  0.6 * input[1] + 0.3 * input[2] + 0.1 * input[3]
}

x <- c(1, 2, 3)
y <- c(0.5, 1.5, 2.5)

# Test superposition: L(x + y) should equal L(x) + L(y)
L_sum <- L(x + y)
sum_L <- L(x) + L(y)

data.frame(
  test = c("L(x + y)", "L(x) + L(y)"),
  value = c(L_sum, sum_L)
)
```

Why is linearity so powerful? If we know how a system responds to a set of **basis inputs**, we can predict its response to **any** input (as a weighted sum of those basis responses).

### Cone Excitations

The transformation from spectral irradiance to cone excitations is linear. For a cone with fundamental $C(\lambda)$:

$$N(r_x, r_y) = \int C(\lambda) \cdot E(r_x, r_y, \lambda) \, d\lambda$$

In discrete form (how we compute it):

$$N(r_x, r_y) = \sum_{\lambda_i} C(\lambda_i) \cdot E(r_x, r_y, \lambda_i) \cdot \Delta\lambda$$

This is just a weighted sum - the inner product of the cone fundamental with the spectrum.

```{r cone-fundamentals}
# Simplified cone fundamentals (Gaussian approximations)
wavelengths <- seq(400, 700, by = 5)

# L-cone peaks ~565nm, M-cone ~535nm, S-cone ~445nm
L_cone <- exp(-0.5 * ((wavelengths - 565) / 50)^2)
M_cone <- exp(-0.5 * ((wavelengths - 535) / 45)^2)
S_cone <- exp(-0.5 * ((wavelengths - 445) / 35)^2)

cone_data <- data.frame(
  wavelength = rep(wavelengths, 3),
  sensitivity = c(L_cone, M_cone, S_cone),
  cone = rep(c("L (long)", "M (medium)", "S (short)"), each = length(wavelengths))
)

ggplot(cone_data, aes(x = wavelength, y = sensitivity, color = cone)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("L (long)" = "red", "M (medium)" = "green", "S (short)" = "blue")) +
  labs(title = "Human Cone Fundamentals (simplified)",
       x = "Wavelength (nm)", y = "Relative Sensitivity") +
  theme_minimal()
```

### Matrix Formulation

With three cone types (L, M, S), we can write cone excitations as matrix multiplication:

$$\mathbf{n}(r_x, r_y) = \mathbf{C} \cdot \mathbf{e}(r_x, r_y)$$

where:

-   $\mathbf{C}$ is a $3 \times w$ matrix (3 cone types, $w$ wavelengths)
-   $\mathbf{e}$ is a $w \times 1$ vector (spectral irradiance at one location)
-   $\mathbf{n}$ is a $3 \times 1$ vector (L, M, S cone excitations)

```{r matrix-formulation}
# Create cone matrix C (3 x w)
C <- rbind(L_cone, M_cone, S_cone)

# Example spectrum: "white" light (equal energy)
e_white <- rep(1, length(wavelengths))

# Compute cone excitations: n = C * e
n_white <- C %*% e_white

data.frame(
  cone = c("L", "M", "S"),
  excitation = round(as.vector(n_white), 2)
)
```

### Color Matching and Metamers

A profound consequence: **different spectra can produce identical cone excitations**. These pairs are called **metamers**.

For two spectra $\mathbf{e}_1$ and $\mathbf{e}_2$ to be metamers:

$$\mathbf{C} \cdot \mathbf{e}_1 = \mathbf{C} \cdot \mathbf{e}_2$$

This means their difference is in the **null space** of $\mathbf{C}$:

$$\mathbf{C} \cdot (\mathbf{e}_2 - \mathbf{e}_1) = \mathbf{0}$$

Wyszecki called vectors in this null space "**metameric blacks**" - adding them to any spectrum creates a metamer.

```{r metamers}
# Create two different spectra that produce the same cone response

# Spectrum 1: narrow yellow (580nm)
e_yellow <- exp(-0.5 * ((wavelengths - 580) / 15)^2)

# Spectrum 2: mixture of red + green
e_red <- exp(-0.5 * ((wavelengths - 620) / 15)^2)
e_green <- exp(-0.5 * ((wavelengths - 540) / 15)^2)

# Find weights to match yellow's cone excitations
n_yellow <- C %*% e_yellow

# We want: w_r * C*e_red + w_g * C*e_green = C*e_yellow
# Solve for weights using least squares
A <- cbind(C %*% e_red, C %*% e_green)
weights <- solve(t(A) %*% A) %*% t(A) %*% n_yellow

e_mixture <- as.numeric(weights[1]) * e_red + as.numeric(weights[2]) * e_green
n_mixture <- C %*% e_mixture

# Compare
data.frame(
  cone = c("L", "M", "S"),
  yellow = round(as.vector(n_yellow), 3),
  red_green_mix = round(as.vector(n_mixture), 3),
  difference = round(as.vector(n_yellow - n_mixture), 6)
)
```

The spectra are physically different, but perceptually identical!

```{r plot-metamers}
#| fig-width: 6
#| fig-height: 3

metamer_data <- data.frame(
  wavelength = rep(wavelengths, 2),
  intensity = c(e_yellow / max(e_yellow), e_mixture / max(e_mixture)),
  spectrum = rep(c("Yellow (narrow)", "Red+Green mix"), each = length(wavelengths))
)

ggplot(metamer_data, aes(x = wavelength, y = intensity, color = spectrum)) +
  geom_line(linewidth = 1) +
  labs(title = "Metameric Spectra",
       subtitle = "Different physics, same perception",
       x = "Wavelength (nm)", y = "Relative Intensity") +
  theme_minimal()
```

This is why displays work! They don't reproduce the original spectrum - they produce a **metamer** using just RGB primaries.

### Noise in Sensory Measurements

Cone excitations are inherently noisy. Two distributions matter:

**Gaussian**: $N(x; \mu, \sigma) = \frac{1}{\sigma\sqrt{2\pi}} \exp\left(-\frac{(x-\mu)^2}{2\sigma^2}\right)$

-   Two parameters: mean $\mu$, variance $\sigma^2$
-   Often used for "signal-independent" noise (same variance everywhere)

**Poisson**: $P(M; \mu) = e^{-\mu} \frac{\mu^M}{M!}$

-   One parameter: $\mu$ (mean = variance)
-   "Signal-dependent" noise: variance scales with signal

Cone excitations follow **Poisson** statistics - this is fundamental.

```{r noise-comparison}
#| fig-width: 7
#| fig-height: 3

# Compare Poisson and Gaussian for different means
mu_values <- c(5, 20, 100)

noise_data <- data.frame()

for (mu in mu_values) {
  x_range <- seq(max(0, mu - 4*sqrt(mu)), mu + 4*sqrt(mu), by = 1)

  noise_data <- rbind(noise_data,
    data.frame(x = x_range,
               prob = dpois(x_range, mu),
               dist = "Poisson",
               mu = paste0("mean = ", mu)),
    data.frame(x = x_range,
               prob = dnorm(x_range, mu, sqrt(mu)),
               dist = "Gaussian",
               mu = paste0("mean = ", mu))
  )
}

ggplot(noise_data, aes(x = x, y = prob, color = dist)) +
  geom_line(linewidth = 0.8) +
  geom_point(data = subset(noise_data, dist == "Poisson"), size = 1) +
  facet_wrap(~mu, scales = "free") +
  labs(title = "Gaussian approximation to Poisson",
       subtitle = "Works well when mean > 10",
       x = "Count", y = "Probability") +
  theme_minimal()
```

The critical difference: for **signal-independent** noise, variance is constant. For **Poisson** (signal-dependent), variance scales with the mean. This matters for dark vs bright regions!

### Image Formation and Shift-Invariance

For a planar image, the retinal image $E$ is formed from the displayed image $I$ through the optics:

$$E(r_x, r_y, \lambda) = \alpha E_1(r_x, r_y, \lambda) + \beta E_2(r_x, r_y, \lambda)$$

This is linear! But what about **shift-invariance**? A system is shift-invariant if shifting the input just shifts the output (without changing its form).

For shift-invariant systems, we can use **convolution**:

$$E(r_x, r_y) = \int \int I(u, v) \cdot P(r_x - u, r_y - v) \, du \, dv$$

where $P(r_x, r_y)$ is the **point spread function** - the retinal image of a single point.

```{r convolution}
#| fig-width: 6
#| fig-height: 3

# Simple 1D demonstration: sharp edge blurred by PSF
x <- 1:100

# Sharp edge (step function)
edge <- ifelse(x > 50, 1, 0)

# Point spread function (Gaussian blur)
psf_width <- 5
psf <- dnorm(seq(-15, 15), 0, psf_width)
psf <- psf / sum(psf)

# Convolve (using R's built-in filter)
blurred <- stats::filter(edge, psf, sides = 2)
blurred[is.na(blurred)] <- edge[is.na(blurred)]

conv_data <- data.frame(
  position = rep(x, 2),
  intensity = c(edge, blurred),
  image = rep(c("Original (display)", "Blurred (retina)"), each = length(x))
)

ggplot(conv_data, aes(x = position, y = intensity, color = image)) +
  geom_line(linewidth = 1) +
  labs(title = "Image formation by convolution",
       subtitle = "E(r) = I(x) * P(r-x)",
       x = "Position", y = "Intensity") +
  theme_minimal()
```

**Important caveat**: The human eye is NOT perfectly shift-invariant:

-   PSF varies with retinal eccentricity
-   PSF varies with wavelength (chromatic aberration)
-   Cone density varies across the retina

This is why computational models (like ISETBio) are essential for accurate predictions.

## 1.4 Computational Model of the Initial Encoding

### Why Computational Modeling?

The mathematical principles (linearity, convolution) are powerful guides, but the real eye **deviates significantly** from idealized assumptions:

-   Point spread function (PSF) varies with retinal eccentricity
-   PSF varies with wavelength (chromatic aberration)
-   Cone density and size vary across the retina
-   Pigment density varies

To make accurate predictions, we need **computational models** that account for these variations (e.g., ISETBio).

### Shift-Varying Point Spread

The PSF describes how a single point of light spreads on the retina. It varies with:

1.  **Eccentricity**: blur increases away from fovea
2.  **Wavelength**: different wavelengths focus at different depths (chromatic aberration)
3.  **Pupil size**: larger pupil = more aberrations

```{r psf-eccentricity}
#| fig-width: 7
#| fig-height: 3

# Simulate PSF at different eccentricities
# PSF width increases with eccentricity
eccentricities <- c(0, 5, 15)  # degrees from fovea
psf_widths <- c(1, 2, 4)       # blur increases

x <- seq(-10, 10, length.out = 100)

psf_data <- data.frame()
for (i in 1:length(eccentricities)) {
  psf <- exp(-0.5 * (x / psf_widths[i])^2)
  psf <- psf / max(psf)
  psf_data <- rbind(psf_data, data.frame(
    x = x,
    psf = psf,
    eccentricity = paste0(eccentricities[i], "° from fovea")
  ))
}

ggplot(psf_data, aes(x, psf, color = eccentricity)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Point Spread Function varies with eccentricity",
       subtitle = "More blur in peripheral vision",
       x = "Position (arcmin)", y = "Normalized intensity") +
  theme_minimal()
```

### Shift-Varying Sampling

Cone density is **highest in the fovea** and decreases dramatically with eccentricity:

```{r cone-density}
#| fig-width: 7
#| fig-height: 3

# Approximate cone density vs eccentricity (based on Curcio et al., 1990)
eccentricity <- seq(0, 30, by = 0.5)

# Cone density falls off approximately exponentially from fovea
# Peak ~200,000 cones/mm² at fovea, drops to ~5,000 at 30°
cone_density <- 5000 + 195000 * exp(-eccentricity / 3)

density_df <- data.frame(eccentricity, cone_density)

ggplot(density_df, aes(eccentricity, cone_density / 1000)) +
  geom_line(linewidth = 1.2, color = "darkgreen") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  annotate("text", x = 1, y = 180, label = "Fovea", hjust = 0) +
  labs(title = "Cone density decreases with eccentricity",
       subtitle = "Less spatial information in peripheral vision",
       x = "Eccentricity (degrees)", y = "Cone density (thousands/mm²)") +
  theme_minimal()
```

**Key facts about the cone mosaic:**

-   No S-cones in the very central fovea (specialized for high resolution)
-   Cones become larger (bigger apertures) in periphery
-   Cone spacing limits spatial resolution

```{r mosaic-schematic}
#| fig-width: 7
#| fig-height: 3

# Schematic: cone spacing at different eccentricities
set.seed(42)

# Fovea: dense, small cones
n_fovea <- 100
fovea_x <- runif(n_fovea, 0, 10)
fovea_y <- runif(n_fovea, 0, 10)

# Periphery: sparse, large cones
n_periph <- 20
periph_x <- runif(n_periph, 0, 10)
periph_y <- runif(n_periph, 0, 10)

par(mfrow = c(1, 2), mar = c(3, 1, 3, 1))

plot(fovea_x, fovea_y, pch = 19, cex = 0.8, col = "darkgreen",
     xlim = c(0, 10), ylim = c(0, 10), asp = 1,
     main = "Fovea (0°)", xlab = "", ylab = "", axes = FALSE)
box()

plot(periph_x, periph_y, pch = 19, cex = 2.5, col = "darkgreen",
     xlim = c(0, 10), ylim = c(0, 10), asp = 1,
     main = "Periphery (15°)", xlab = "", ylab = "", axes = FALSE)
box()
```

### Spatial Derivatives

Neurons in the visual system compute **local differences** - approximations to spatial derivatives:

-   **Spatial position**: edge detectors (luminance changes)
-   **Wavelength**: color opponent cells (L-M, S-(L+M))
-   **Time**: motion detectors
-   **Between eyes**: stereo disparity

Why derivatives? Local **contrast** (change), not absolute intensity, provides critical information.

```{r derivatives}
#| fig-width: 7
#| fig-height: 4

# Demo: edge detection as derivative
x <- 1:100
# Image with edge
image <- ifelse(x > 50, 1, 0.3)

# Derivative (difference)
derivative <- c(0, diff(image))

par(mfrow = c(2, 1), mar = c(3, 4, 2, 1))
plot(x, image, type = "l", lwd = 2, col = "black",
     main = "Original image (intensity)", ylab = "Intensity", xlab = "")
plot(x, derivative, type = "l", lwd = 2, col = "red",
     main = "Spatial derivative (edge detection)", ylab = "dI/dx", xlab = "Position")
abline(h = 0, lty = 2, col = "gray")
```

Derivatives are also **compressible** - natural images change slowly, so most derivatives are near zero.

## 1.5 Perceptual Inference

### The Ambiguity Problem

A critical insight: **many different physical scenes can produce the same cone response**.

Examples of many-to-one mappings:

1.  **Metamerism**: Different spectra → same L, M, S values
2.  **Depth reconstruction**: 3D world → 2D retinal image

```{r ambiguity-visual}
#| fig-width: 6
#| fig-height: 4

# Visual: multiple scenes → same cone response
par(mar = c(1, 1, 3, 1))
plot(1, type = "n", xlim = c(0, 10), ylim = c(0, 8),
     axes = FALSE, xlab = "", ylab = "",
     main = "Many scenes can produce the same neural response")

# Multiple scenes on left
rect(0.5, 5.5, 2.5, 7.5, border = "blue", lwd = 2)
text(1.5, 6.5, "Scene A", cex = 0.9)

rect(0.5, 3, 2.5, 5, border = "red", lwd = 2)
text(1.5, 4, "Scene B", cex = 0.9)

rect(0.5, 0.5, 2.5, 2.5, border = "darkgreen", lwd = 2)
text(1.5, 1.5, "Scene C", cex = 0.9)

# Arrows pointing to single response
arrows(2.7, 6.5, 5.3, 4, length = 0.1, lwd = 1.5)
arrows(2.7, 4, 5.3, 4, length = 0.1, lwd = 1.5)
arrows(2.7, 1.5, 5.3, 4, length = 0.1, lwd = 1.5)

# Single cone response box
rect(5.5, 2.5, 8.5, 5.5, border = "black", lwd = 3)
text(7, 4, "Same cone\nresponse\n(L, M, S)", cex = 1)

# Question mark
text(9.2, 4, "?", cex = 3, col = "red")
```

The brain must **infer** which scene actually produced the observed response. This is Helmholtz's "unconscious inference."

### Bayesian Framework

Let:

-   **x** = scene state (what's actually out there)
-   **y** = sensory measurements (noisy cone excitations)

We want to know: given what I measured (y), what scene (x) is most likely?

**Bayes' Rule:**

$$p(x|y) = \frac{p(y|x) \cdot p(x)}{p(y)}$$

Or simply: **posterior ∝ likelihood × prior**

```{r bayes-components}
#| fig-width: 8
#| fig-height: 3

# Visualize the three components of Bayesian inference
x <- seq(-1, 1, length.out = 200)

# Prior: what scenes are likely (before seeing data)
prior <- dnorm(x, mean = 0, sd = 0.1)

# Likelihood: how likely is data given each scene
likelihood <- dnorm(x, mean = 0.5, sd = 0.1)

# Posterior: prior × likelihood (normalized)
posterior_unnorm <- prior * likelihood
posterior <- posterior_unnorm / sum(posterior_unnorm) * length(x)

bayes_df <- data.frame(
  x = rep(x, 3),
  density = c(prior, likelihood * 2, posterior),  # scale for visibility
  component = rep(c("Prior p(x)", "Likelihood p(y|x)", "Posterior p(x|y)"), each = length(x))
)
bayes_df$component <- factor(bayes_df$component,
                              levels = c("Prior p(x)", "Likelihood p(y|x)", "Posterior p(x|y)"))

ggplot(bayes_df, aes(x, density, color = component)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Prior p(x)" = "blue",
                                 "Likelihood p(y|x)" = "red",
                                 "Posterior p(x|y)" = "purple")) +
  labs(title = "Bayesian Inference: Posterior = Prior × Likelihood",
       subtitle = "The posterior combines what's likely (prior) with what data says (likelihood)",
       x = "Scene parameter x", y = "Density") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**The key components:**

| Component | Meaning | Example |
|----|----|----|
| **Prior** p(x) | What scenes are likely before seeing data | "Natural images have smooth gradients" |
| **Likelihood** p(y\|x) | How likely is this data if scene is x | Forward model: scene → cone response + noise |
| **Posterior** p(x\|y) | What scene is most likely given data | Our best guess about the world |

### Simple Example: Estimating Light Intensity

Suppose a cone receives photons from a light source. The true intensity is unknown, but we:

1.  Have a **prior** belief about likely intensities
2.  Observe noisy **data** (photon count follows Poisson)

```{r bayes-example}
#| fig-width: 8
#| fig-height: 6

# Simple Bayesian estimation example
#set.seed(123)

# True intensity (unknown to observer)
true_intensity <- 25

# Observed data: noisy photon count (Poisson)
observed_count <- rpois(1, true_intensity)

# Possible intensities
intensities <- seq(0, 60, by = 0.5)

# Prior: broad belief about intensities (log-normal, typical for light)
prior <- dlnorm(intensities, meanlog = log(20), sdlog = 0.5)
prior <- prior / sum(prior)

# Likelihood: Poisson probability of observed count given each intensity
likelihood <- dpois(observed_count, lambda = intensities)
likelihood[1] <- 0  # avoid div by 0

# Posterior (unnormalized)
posterior_unnorm <- prior * likelihood
posterior <- posterior_unnorm / sum(posterior_unnorm)

# MAP estimate
map_estimate <- intensities[which.max(posterior)]

par(mfrow = c(3, 1), mar = c(4, 4, 3, 1))

plot(intensities, prior, type = "l", lwd = 2, col = "blue",
     main = "Step 1: Prior belief about intensity",
     xlab = "Intensity", ylab = "Probability")
abline(v = true_intensity, lty = 2, col = "gray50")
legend("topright", "True intensity", lty = 2, col = "gray50", bty = "n")

plot(intensities, likelihood, type = "l", lwd = 2, col = "red",
     main = paste0("Step 2: Likelihood (observed ", observed_count, " photons)"),
     xlab = "Intensity", ylab = "Likelihood")
abline(v = true_intensity, lty = 2, col = "gray50")

plot(intensities, posterior, type = "l", lwd = 2, col = "purple",
     main = paste0("Step 3: Posterior (MAP estimate = ", map_estimate, ")"),
     xlab = "Intensity", ylab = "Probability")
abline(v = true_intensity, lty = 2, col = "gray50")
abline(v = map_estimate, lty = 1, col = "purple")
legend("topright", c("True intensity", "MAP estimate"),
       lty = c(2, 1), col = c("gray50", "purple"), bty = "n")
```

The posterior is **more concentrated** than either prior or likelihood alone - we've combined both sources of information.

### Linear Model with Gaussian Noise

For the initial encoding, we can write:

$$y = Cx + \epsilon$$

where:

-   y = noisy cone excitations
-   C = cone sensitivity matrix
-   x = scene (spectral radiance)
-   ε = noise

If we assume Gaussian noise and a Gaussian prior, the posterior is also Gaussian (nice math!):

$$p(y|x) = \text{Normal}(Cx, \sigma^2 I)$$ $$p(x) = \text{Normal}(\mu_x, \Sigma_x)$$

```{r gaussian-posterior-setup}
#| echo: true

# =============================================================================
# 2D MATRIX EXAMPLE: 2-pixel display with 1 cone measurement
# Based on Figure 1.9 from Wandell & Brainard
# =============================================================================
#
# THE SETUP (Equation 1.15):
#   y = C·x + ε
#
# Variables:
#   x = [x₁, x₂]ᵀ     Scene: 2 pixel intensities (what we want to infer)
#   C = [0.5, 0.5]    Cone sensitivity: equally sensitive to both pixels
#   y = scalar        Measurement: single noisy cone excitation
#   ε ~ N(0, σ²)      Noise: Gaussian with σ = 0.1
#
# The forward model:
#   E[y] = 0.5·x₁ + 0.5·x₂  (weighted average of pixel values)
#
# THE PROBLEM:
#   Given observation y, what was x = [x₁, x₂]?
#   Many x values produce the same y! (ambiguity)

# Define the model parameters
C <- matrix(c(0.5, 0.5), nrow = 1)  # 1×2 cone sensitivity matrix
sigma_noise <- 0.1                   # measurement noise SD

cat("Cone sensitivity matrix C:\n")
print(C)
cat("\nForward model: y = C·x + noise = 0.5·x₁ + 0.5·x₂ + ε\n")
```

### The Prior: What scenes are likely?

Natural images have **correlated** nearby pixels - if one pixel is bright, its neighbor is likely bright too. We encode this as a **bivariate Gaussian prior**:

$$p(\mathbf{x}) = \mathcal{N}(\boldsymbol{\mu}_x, \boldsymbol{\Sigma}_x)$$

```{r gaussian-prior-params}
# Prior parameters (Equation 1.17)
mu_x <- c(0.5, 0.5)        # prior mean: expect mid-gray pixels
sigma_x <- 0.2             # marginal SD for each pixel
rho <- 0.89                # correlation: nearby pixels are similar

# Covariance matrix Σ_x
Sigma_x <- matrix(c(sigma_x^2,           rho*sigma_x^2,
                    rho*sigma_x^2,       sigma_x^2),
                  nrow = 2, byrow = TRUE)

cat("Prior mean μ_x:\n")
print(mu_x)
cat("\nPrior covariance Σ_x:\n")
print(round(Sigma_x, 4))
cat("\nCorrelation ρ =", rho, "(high correlation = natural images)\n")
```

::: callout-note
## Reminder: What is a Covariance Matrix?

The **covariance matrix** $\boldsymbol{\Sigma}$ captures how variables vary together:

$$\boldsymbol{\Sigma} = \begin{bmatrix} \text{Var}(X_1) & \text{Cov}(X_1, X_2) \\ \text{Cov}(X_2, X_1) & \text{Var}(X_2) \end{bmatrix} = \begin{bmatrix} \sigma_1^2 & \rho\sigma_1\sigma_2 \\ \rho\sigma_1\sigma_2 & \sigma_2^2 \end{bmatrix}$$

**Key properties:**

| Property | Meaning |
|----|----|
| **Diagonal elements** | Variances: how much each variable spreads |
| **Off-diagonal elements** | Covariances: how variables move together |
| **Always symmetric** | $\Sigma_{ij} = \Sigma_{ji}$ because $\text{Cov}(X_i, X_j) = \text{Cov}(X_j, X_i)$ |
| **Positive semi-definite** | $\mathbf{v}^T \Sigma \mathbf{v} \geq 0$ for all vectors $\mathbf{v}$ |

**Correlation vs Covariance:**

-   Covariance: $\text{Cov}(X_1, X_2) = \rho \cdot \sigma_1 \cdot \sigma_2$ (depends on scale)
-   Correlation: $\rho = \frac{\text{Cov}(X_1, X_2)}{\sigma_1 \sigma_2}$ (normalized, always between -1 and +1)

In our example: $\rho = 0.89$ means pixels are **strongly positively correlated** — when one is bright, the other likely is too.
:::

**Step 1: Create a grid of possible scenes**

We want to evaluate the prior probability for many possible $(x_1, x_2)$ combinations:

```{r prior-grid-setup}
# Create grid for 2D plotting
# Each pixel can range from 0 (black) to 1 (white)
x1_seq <- seq(0, 1, length.out = 100)  # 100 possible values for pixel 1
x2_seq <- seq(0, 1, length.out = 100)  # 100 possible values for pixel 2

# expand.grid creates ALL combinations: 100 × 100 = 10,000 possible scenes
grid <- expand.grid(x1 = x1_seq, x2 = x2_seq)

cat("Grid dimensions:", nrow(grid), "rows (one per possible scene)\n")
```

::: callout-note
## From 1D Gaussian to Multivariate Gaussian

**1D Gaussian (univariate):** One variable, bell curve

$$p(x) = \frac{1}{\sigma\sqrt{2\pi}} \exp\left(-\frac{(x - \mu)^2}{2\sigma^2}\right)$$

**2D Gaussian (bivariate):** Two variables, bell "hill"

$$p(x_1, x_2) = \frac{1}{2\pi|\boldsymbol{\Sigma}|^{1/2}} \exp\left(-\frac{1}{2}(\mathbf{x} - \boldsymbol{\mu})^T \boldsymbol{\Sigma}^{-1} (\mathbf{x} - \boldsymbol{\mu})\right)$$

| 1D Gaussian | 2D Gaussian |
|----|----|
| Mean: $\mu$ (a number) | Mean: $\boldsymbol{\mu} = [\mu_1, \mu_2]^T$ (a vector) |
| Variance: $\sigma^2$ (a number) | Covariance: $\boldsymbol{\Sigma}$ (a 2×2 matrix) |
| $(x - \mu)^2$ | $(\mathbf{x} - \boldsymbol{\mu})^T \boldsymbol{\Sigma}^{-1} (\mathbf{x} - \boldsymbol{\mu})$ |

The key insight: **the covariance matrix replaces the single variance**, and it can encode correlations between variables!
:::

**Step 1b: Visualizing 1D vs 2D Gaussian**

```{r gaussian-1d-2d-3d}
#| fig-width: 10
#| fig-height: 4

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

# LEFT: 1D Gaussian - Bell Curve
x_1d <- seq(-3, 3, length.out = 100)
y_1d <- dnorm(x_1d, mean = 0, sd = 1)
plot(x_1d, y_1d, type = "l", lwd = 3, col = "blue",
     main = "1D Gaussian\nBell Curve",
     xlab = "x", ylab = "p(x)")
polygon(c(x_1d, rev(x_1d)), c(y_1d, rep(0, length(y_1d))),
        col = rgb(0, 0, 1, 0.2), border = NA)
abline(v = 0, lty = 2, col = "red")

# MIDDLE: 2D Gaussian - Contours (top view)
x1_temp <- seq(-3, 3, length.out = 50)
x2_temp <- seq(-3, 3, length.out = 50)
z_2d <- outer(x1_temp, x2_temp, function(a, b) {
  dnorm(a, 0, 1) * dnorm(b, 0, 1)
})
contour(x1_temp, x2_temp, z_2d, nlevels = 10, col = "blue",
        main = "2D Gaussian\nContours (top view)",
        xlab = expression(x[1]), ylab = expression(x[2]))
points(0, 0, pch = 19, col = "red", cex = 1.5)

# RIGHT: 2D Gaussian - 3D Surface (the actual bell hill!)
persp(x1_temp, x2_temp, z_2d,
      theta = 30, phi = 25,        # viewing angles
      col = "lightblue",
      shade = 0.5,
      border = NA,
      main = "2D Gaussian\n3D Surface (bell hill)",
      xlab = "x₁", ylab = "x₂", zlab = "p(x)")
```

**The 3D view shows the actual "bell hill"** — contours are like elevation lines on a topographic map!

**Step 1b-2: 3D surfaces with different correlations**

```{r gaussian-3d-correlations}
#| fig-width: 10
#| fig-height: 8

# Create 3D surface for a given covariance matrix
make_gaussian_surface <- function(Sigma, n = 50) {
  x1 <- seq(-3, 3, length.out = n)
  x2 <- seq(-3, 3, length.out = n)
  inv_Sigma <- solve(Sigma)

  z <- outer(x1, x2, function(a, b) {
    sapply(1:length(a), function(i) {
      x <- c(a[i], b[i])
      mahal <- sum(x * (inv_Sigma %*% x))
      exp(-0.5 * mahal) / (2 * pi * sqrt(det(Sigma)))
    })
  })
  list(x1 = x1, x2 = x2, z = z)
}

par(mfrow = c(2, 2), mar = c(1, 1, 3, 1))

# Case 1: Independent (ρ = 0) - circular base
Sigma1 <- matrix(c(1, 0, 0, 1), 2, 2)
surf1 <- make_gaussian_surface(Sigma1)
persp(surf1$x1, surf1$x2, surf1$z,
      theta = 30, phi = 25, col = "lightblue", shade = 0.5, border = NA,
      main = "ρ = 0 (independent)\nCircular base",
      xlab = "", ylab = "", zlab = "", ticktype = "detailed")

# Case 2: Positive correlation (ρ = 0.8) - tilted ridge
Sigma2 <- matrix(c(1, 0.8, 0.8, 1), 2, 2)
surf2 <- make_gaussian_surface(Sigma2)
persp(surf2$x1, surf2$x2, surf2$z,
      theta = 30, phi = 25, col = "lightgreen", shade = 0.5, border = NA,
      main = "ρ = 0.8 (positive correlation)\nRidge along diagonal",
      xlab = "", ylab = "", zlab = "", ticktype = "detailed")

# Case 3: Negative correlation (ρ = -0.8) - tilted other way
Sigma3 <- matrix(c(1, -0.8, -0.8, 1), 2, 2)
surf3 <- make_gaussian_surface(Sigma3)
persp(surf3$x1, surf3$x2, surf3$z,
      theta = 30, phi = 25, col = "lightyellow", shade = 0.5, border = NA,
      main = "ρ = -0.8 (negative correlation)\nRidge along anti-diagonal",
      xlab = "", ylab = "", zlab = "", ticktype = "detailed")

# Case 4: High correlation (ρ = 0.95) - very narrow ridge
Sigma4 <- matrix(c(1, 0.95, 0.95, 1), 2, 2)
surf4 <- make_gaussian_surface(Sigma4)
persp(surf4$x1, surf4$x2, surf4$z,
      theta = 30, phi = 25, col = "pink", shade = 0.5, border = NA,
      main = "ρ = 0.95 (very high correlation)\nNarrow ridge - x1 ≈ x2",
      xlab = "", ylab = "", zlab = "", ticktype = "detailed")
```

**Key insight from 3D view:**

-   **ρ = 0**: Bell hill with circular base (variables independent)
-   **ρ \> 0**: Ridge stretched along diagonal (when $x_1$ ↑, $x_2$ ↑ too)
-   **ρ \< 0**: Ridge stretched along anti-diagonal (when $x_1$ ↑, $x_2$ ↓)
-   **ρ → 1**: Ridge becomes a narrow "blade" — knowing $x_1$ almost tells you $x_2$

Our prior uses **ρ = 0.89** — a fairly narrow ridge, meaning the two pixels are expected to be very similar!

**Step 1c: How the covariance matrix shapes the distribution**

The covariance matrix $\boldsymbol{\Sigma}$ controls the **shape and orientation** of the Gaussian:

```{r covariance-shapes}
#| fig-width: 10
#| fig-height: 8

# Function to plot 2D Gaussian contours
plot_2d_gaussian <- function(Sigma, title, show_matrix = TRUE) {
  # Create grid
  x1 <- seq(-3, 3, length.out = 100)
  x2 <- seq(-3, 3, length.out = 100)

  # Compute density
  z <- matrix(0, length(x1), length(x2))
  for (i in 1:length(x1)) {
    for (j in 1:length(x2)) {
      x <- c(x1[i], x2[j])
      inv_S <- solve(Sigma)
      mahal <- sum(x * (inv_S %*% x))
      z[i, j] <- exp(-0.5 * mahal)
    }
  }

  contour(x1, x2, z, nlevels = 8, col = "blue", lwd = 1.5,
          main = title, xlab = expression(x[1]), ylab = expression(x[2]),
          asp = 1)
  abline(h = 0, v = 0, lty = 3, col = "gray")
  points(0, 0, pch = 19, col = "red")

  if (show_matrix) {
    # Show matrix values in corner
    legend("topright",
           legend = c(paste0("Σ₁₁ = ", Sigma[1,1]),
                      paste0("Σ₁₂ = ", Sigma[1,2]),
                      paste0("Σ₂₂ = ", Sigma[2,2])),
           bty = "n", cex = 0.9)
  }
}

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# Case 1: Identity (independent, equal variance)
Sigma1 <- matrix(c(1, 0, 0, 1), 2, 2)
plot_2d_gaussian(Sigma1, "Independent, equal variance\nρ = 0, σ₁ = σ₂ = 1")

# Case 2: Different variances (still independent)
Sigma2 <- matrix(c(2, 0, 0, 0.5), 2, 2)
plot_2d_gaussian(Sigma2, "Independent, different variance\nρ = 0, σ₁ = 1.4, σ₂ = 0.7")

# Case 3: Positive correlation
Sigma3 <- matrix(c(1, 0.8, 0.8, 1), 2, 2)
plot_2d_gaussian(Sigma3, "Positive correlation\nρ = 0.8")

# Case 4: Negative correlation
Sigma4 <- matrix(c(1, -0.8, -0.8, 1), 2, 2)
plot_2d_gaussian(Sigma4, "Negative correlation\nρ = -0.8")
```

**Key observations:**

| Covariance Matrix | Shape | Meaning |
|----|----|----|
| $\begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix}$ | Circle | Variables are independent, equal spread |
| $\begin{bmatrix} 2 & 0 \\ 0 & 0.5 \end{bmatrix}$ | Horizontal ellipse | $x_1$ has more variance than $x_2$ |
| $\begin{bmatrix} 1 & 0.8 \\ 0.8 & 1 \end{bmatrix}$ | Tilted ellipse (↗) | Positive correlation: $x_1$ ↑ means $x_2$ ↑ |
| $\begin{bmatrix} 1 & -0.8 \\ -0.8 & 1 \end{bmatrix}$ | Tilted ellipse (↘) | Negative correlation: $x_1$ ↑ means $x_2$ ↓ |

**Step 1d: The matrix algebra - visualized step by step**

::: callout-note
## What is a Matrix Inverse? (Σ⁻¹)

**The problem:** In the Gaussian formula, we need to "divide" by the covariance matrix. But you can't divide by a matrix! Instead, we multiply by its **inverse**.

**Analogy with numbers:** - To compute $\frac{5}{2}$, you can write it as $5 \times \frac{1}{2} = 5 \times 2^{-1}$ - The inverse of 2 is $2^{-1} = 0.5$ because $2 \times 0.5 = 1$

**For matrices:** - The inverse of $\boldsymbol{\Sigma}$ is $\boldsymbol{\Sigma}^{-1}$ such that $\boldsymbol{\Sigma} \times \boldsymbol{\Sigma}^{-1} = \mathbf{I}$ (identity matrix) - Identity matrix $\mathbf{I} = \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix}$ is like "1" for matrices

**For a 2×2 matrix**, there's a formula:

$$\boldsymbol{\Sigma} = \begin{bmatrix} a & b \\ c & d \end{bmatrix} \quad \Rightarrow \quad \boldsymbol{\Sigma}^{-1} = \frac{1}{ad - bc} \begin{bmatrix} d & -b \\ -c & a \end{bmatrix}$$

The term $(ad - bc)$ is called the **determinant** (written $|\boldsymbol{\Sigma}|$ or $\det(\boldsymbol{\Sigma})$).

**In R:** `solve(Sigma)` computes the inverse.
:::

```{r matrix-inverse-example}
# Let's see this concretely with our covariance matrix
cat("Our covariance matrix Σ:\n")
print(Sigma_x)

cat("\nCompute inverse using solve():\n")
inv_Sigma_x <- solve(Sigma_x)
print(round(inv_Sigma_x, 4))

cat("\nVerify: Σ × Σ⁻¹ should equal Identity matrix I:\n")
print(round(Sigma_x %*% inv_Sigma_x, 10))

cat("\nYes! The off-diagonals are essentially 0, diagonals are 1.\n")
```

```{r matrix-inverse-visual}
#| fig-width: 10
#| fig-height: 4

# Visualize what the inverse does
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

# Function to draw an ellipse from covariance matrix
draw_cov_ellipse <- function(Sigma, mu = c(0, 0), col = "blue", ...) {
  angles <- seq(0, 2*pi, length.out = 100)
  circle <- cbind(cos(angles), sin(angles))
  # Transform circle by sqrt of covariance (Cholesky)
  L <- chol(Sigma)
  ellipse <- t(t(circle %*% L) + mu)
  lines(ellipse, col = col, lwd = 2, ...)
}

# LEFT: Original covariance Σ - wide ellipse along diagonal
plot(NULL, xlim = c(-4, 4), ylim = c(-4, 4), asp = 1,
     main = "Covariance Σ\n(spreads data along diagonal)",
     xlab = expression(x[1]), ylab = expression(x[2]))
abline(h = 0, v = 0, col = "gray", lty = 2)
draw_cov_ellipse(Sigma_x * 20, col = "blue")  # scaled for visibility
text(2, 2, "Wide along\ndiagonal", col = "blue")

# MIDDLE: Inverse covariance Σ⁻¹ - narrow ellipse perpendicular
plot(NULL, xlim = c(-4, 4), ylim = c(-4, 4), asp = 1,
     main = expression("Inverse " * Sigma^-1 * "\n(narrow perpendicular)"),
     xlab = expression(x[1]), ylab = expression(x[2]))
abline(h = 0, v = 0, col = "gray", lty = 2)
draw_cov_ellipse(inv_Sigma_x / 50, col = "red")  # scaled for visibility
text(-2, 2, "Narrow\nperpendicular", col = "red")

# RIGHT: What this means for distance
plot(NULL, xlim = c(-3, 3), ylim = c(-3, 3), asp = 1,
     main = "Effect on Distance",
     xlab = expression(x[1]), ylab = expression(x[2]))
abline(h = 0, v = 0, col = "gray", lty = 2)
abline(0, 1, col = "blue", lty = 2)  # diagonal
abline(0, -1, col = "red", lty = 2)  # anti-diagonal

# Two points same Euclidean distance from origin
p1 <- c(1.5, 1.5) / sqrt(2)  # on diagonal
p2 <- c(1.5, -1.5) / sqrt(2)  # perpendicular

points(p1[1], p1[2], pch = 19, col = "blue", cex = 2)
points(p2[1], p2[2], pch = 19, col = "red", cex = 2)

text(p1[1] + 0.3, p1[2], "Small Mahal.\ndistance", col = "blue", cex = 0.8)
text(p2[1] + 0.3, p2[2], "Large Mahal.\ndistance", col = "red", cex = 0.8)
legend("topleft", c("Along correlation", "Against correlation"),
       col = c("blue", "red"), pch = 19, bty = "n")
```

**Key insight:** The inverse Σ⁻¹ "flips" the covariance's shape: - Where Σ is wide (along diagonal), Σ⁻¹ is narrow → moving along correlation costs **less** - Where Σ is narrow (perpendicular), Σ⁻¹ is wide → moving against correlation costs **more**

This is why the Mahalanobis distance is smaller along the correlation direction!

::: callout-note
## Why Transpose? Understanding Matrix Dimensions

**The problem:** We need the Mahalanobis distance to be a **single number** (scalar), not a vector or matrix.

**Column vector vs Row vector:**

```         
Column vector (2×1):     Row vector (1×2):
    ┌───┐
    │ a │                [ a  b ]
    │ b │
    └───┘
```

By convention, vectors are **column vectors**. The transpose (ᵀ) flips a column into a row:

$$\mathbf{v} = \begin{bmatrix} a \\ b \end{bmatrix} \quad \Rightarrow \quad \mathbf{v}^T = \begin{bmatrix} a & b \end{bmatrix}$$

**Matrix multiplication dimension rules:**

To multiply matrices, inner dimensions must match:

```         
(m × n) × (n × p) = (m × p)
        ↑   ↑
      must match
```

**For Mahalanobis distance** $(x-\mu)^T \Sigma^{-1} (x-\mu)$:

```         
Step 1: Σ⁻¹ × (x - μ)
        (2×2) × (2×1) = (2×1)  ← result is column vector

Step 2: (x - μ)ᵀ × [result from Step 1]
        (1×2) × (2×1) = (1×1)  ← SCALAR! ✓
```

Without the transpose, we'd get:

```         
(2×1) × (2×1) = ERROR! Dimensions don't match!
```
:::

```{r transpose-visual}
#| fig-width: 8
#| fig-height: 3

# Visualize the matrix multiplication dimensions
par(mar = c(1, 1, 2, 1))
plot(NULL, xlim = c(0, 10), ylim = c(0, 3), axes = FALSE,
     main = "Matrix Multiplication: (x-μ)ᵀ Σ^-1 (x-μ) = scalar")

# (x - μ)ᵀ : 1×2
rect(0.5, 1.5, 1.5, 2.5, border = "blue", lwd = 2)
text(1, 2, "1×2", col = "blue", cex = 0.9)
text(1, 1.2, "(x-μ)ᵀ", col = "blue", cex = 0.8)

# × symbol
text(1.8, 2, "×", cex = 1.2)

# Σ⁻¹ : 2×2
rect(2.2, 1, 3.5, 2.5, border = "red", lwd = 2)
text(2.85, 1.75, "2×2", col = "red", cex = 0.9)
text(2.85, 0.7, "Σ^-1", col = "red", cex = 0.8)

# × symbol
text(3.8, 2, "×", cex = 1.2)

# (x - μ) : 2×1
rect(4.2, 1, 4.8, 2.5, border = "darkgreen", lwd = 2)
text(4.5, 1.75, "2×1", col = "darkgreen", cex = 0.9)
text(4.5, 0.7, "(x-μ)", col = "darkgreen", cex = 0.8)

# = symbol
text(5.3, 2, "=", cex = 1.2)

# Result: 1×1 (scalar)
rect(5.7, 1.8, 6.3, 2.3, border = "purple", lwd = 3)
text(6, 2.05, "1×1", col = "purple", cex = 0.9)
text(6, 1.5, "scalar!", col = "purple", cex = 0.9, font = 2)

# Show dimension matching
arrows(1.3, 2.5, 2.4, 2.5, length = 0.1, col = "gray50")
text(1.85, 2.7, "2=2", cex = 0.7, col = "gray50")

arrows(3.3, 2.5, 4.0, 2.5, length = 0.1, col = "gray50")
text(3.65, 2.7, "2=2", cex = 0.7, col = "gray50")
```

Let's compute the Mahalanobis distance for a specific point, showing each matrix operation:

```{r matrix-algebra-visual}
#| fig-width: 9
#| fig-height: 6

# Our covariance matrix (from the prior)
cat("=== STEP-BY-STEP MATRIX ALGEBRA ===\n\n")

# The point we're evaluating
x_point <- c(0.7, 0.5)
mu_point <- c(0.5, 0.5)

cat("Point x =", x_point[1], ",", x_point[2], "\n")
cat("Mean μ =", mu_point[1], ",", mu_point[2], "\n\n")

# Step A: Center the point
cat("STEP A: Center the point (x - μ)\n")
x_centered <- x_point - mu_point
cat("x - μ = [", x_point[1], "-", mu_point[1], ",", x_point[2], "-", mu_point[2], "]\n")
cat("     = [", x_centered[1], ",", x_centered[2], "]\n\n")

# Step B: Show the covariance matrix
cat("STEP B: The covariance matrix Σ\n")
print(Sigma_x)
cat("\nThis encodes: σ² =", sigma_x^2, ", ρ =", rho, "\n\n")

# Step C: Compute the inverse
cat("STEP C: Compute Σ⁻¹ (the inverse)\n")
inv_Sigma_x <- solve(Sigma_x)
cat("Σ⁻¹ =\n")
print(round(inv_Sigma_x, 3))
cat("\n")

# Step D: Matrix multiplication Σ⁻¹(x - μ)
cat("STEP D: Multiply Σ⁻¹ × (x - μ)\n")
step_d <- inv_Sigma_x %*% x_centered
cat("Σ⁻¹(x - μ) = [", round(step_d[1], 3), ",", round(step_d[2], 3), "]ᵀ\n\n")

# Step E: Dot product (x - μ)ᵀ × [Σ⁻¹(x - μ)]
cat("STEP E: Compute (x - μ)ᵀ Σ⁻¹ (x - μ)\n")
mahal_squared <- sum(x_centered * step_d)
cat("= [", x_centered[1], ",", x_centered[2], "] · [",
    round(step_d[1], 3), ",", round(step_d[2], 3), "]\n")
cat("=", x_centered[1], "×", round(step_d[1], 3), "+",
    x_centered[2], "×", round(step_d[2], 3), "\n")
cat("=", round(mahal_squared, 4), "(Mahalanobis distance squared)\n\n")

# Step F: Final density
cat("STEP F: Compute the density\n")
k <- 2
normalizer <- sqrt((2*pi)^k * det(Sigma_x))
density_value <- exp(-0.5 * mahal_squared) / normalizer
cat("p(x) = exp(-0.5 ×", round(mahal_squared, 4), ") / √(2π)² |Σ|\n")
cat("     = exp(", round(-0.5 * mahal_squared, 4), ") /", round(normalizer, 4), "\n")
cat("     =", round(density_value, 6), "\n")
```

```{r matrix-algebra-plot}
#| fig-width: 7
#| fig-height: 6

# Visualize this computation
par(mar = c(4, 4, 3, 1))

# Plot the prior contours
x1_plot <- seq(0, 1, length.out = 100)
x2_plot <- seq(0, 1, length.out = 100)
z_plot <- matrix(0, length(x1_plot), length(x2_plot))
for (i in 1:length(x1_plot)) {
  for (j in 1:length(x2_plot)) {
    x_temp <- c(x1_plot[i], x2_plot[j])
    x_c <- x_temp - mu_point
    m <- sum(x_c * (inv_Sigma_x %*% x_c))
    z_plot[i, j] <- exp(-0.5 * m)
  }
}

contour(x1_plot, x2_plot, z_plot, nlevels = 10, col = "purple",
        main = "Visualizing the Matrix Algebra",
        xlab = expression(x[1]), ylab = expression(x[2]), asp = 1)

# Mark the mean
points(mu_point[1], mu_point[2], pch = 19, col = "blue", cex = 2)
text(mu_point[1] + 0.05, mu_point[2] + 0.05, "μ = (0.5, 0.5)", pos = 4, col = "blue")

# Mark the point
points(x_point[1], x_point[2], pch = 17, col = "red", cex = 2)
text(x_point[1] + 0.02, x_point[2] - 0.05, "x = (0.7, 0.5)", pos = 4, col = "red")

# Draw the vector (x - μ)
arrows(mu_point[1], mu_point[2], x_point[1], x_point[2],
       col = "darkgreen", lwd = 2, length = 0.1)
text(0.6, 0.52, "x - μ", col = "darkgreen", pos = 3)

# Add legend
legend("bottomright",
       legend = c("Mean μ", "Point x", "Vector (x - μ)"),
       pch = c(19, 17, NA), lty = c(NA, NA, 1), lwd = c(NA, NA, 2),
       col = c("blue", "red", "darkgreen"), bty = "n")
```

**Step 2: Define the multivariate Gaussian density**

The multivariate Gaussian (normal) density function is:

$$p(\mathbf{x}) = \frac{1}{\sqrt{(2\pi)^k |\boldsymbol{\Sigma}|}} \exp\left(-\frac{1}{2}(\mathbf{x} - \boldsymbol{\mu})^T \boldsymbol{\Sigma}^{-1} (\mathbf{x} - \boldsymbol{\mu})\right)$$

Let's break this down:

```{r dmvnorm-explanation}
# Multivariate Gaussian density function - explained step by step
dmvnorm <- function(x, mu, Sigma) {

  # k = number of dimensions (2 for our 2-pixel case)
  k <- length(mu)

  # Step A: Center the data (subtract the mean)
  # x_centered = (x - μ)
  # This shifts the distribution to be centered at origin
  x_centered <- x - mu

  # Step B: Compute the "Mahalanobis distance" squared
  # This measures how far x is from μ, accounting for correlation
  # Formula: (x - μ)ᵀ Σ⁻¹ (x - μ)
  inv_Sigma <- solve(Sigma)  # Σ⁻¹ = inverse of covariance matrix
  mahal <- sum(x_centered * (inv_Sigma %*% x_centered))

  # Step C: Compute the density
  # Numerator: exp(-0.5 × mahalanobis²) — the "bell" shape
  # Denominator: √((2π)^k × |Σ|) — normalization constant
  density <- exp(-0.5 * mahal) / sqrt((2*pi)^k * det(Sigma))

  return(density)
}
```

::: callout-tip
## Understanding the Mahalanobis Distance

The **Mahalanobis distance** is like Euclidean distance, but stretched/rotated by the covariance:

-   **Euclidean:** $d = \sqrt{(x_1 - \mu_1)^2 + (x_2 - \mu_2)^2}$ — treats all directions equally
-   **Mahalanobis:** $d = \sqrt{(\mathbf{x} - \boldsymbol{\mu})^T \boldsymbol{\Sigma}^{-1} (\mathbf{x} - \boldsymbol{\mu})}$ — accounts for correlation

When variables are correlated, moving along the correlation direction is "shorter" than moving perpendicular to it.

```{r mahalanobis-example}
# Example: Two points at same Euclidean distance from mean
point_diagonal <- c(0.6, 0.6)      # Along correlation direction
point_perpendicular <- c(0.6, 0.4) # Perpendicular to correlation

euclidean_diag <- sqrt(sum((point_diagonal - mu_x)^2))
euclidean_perp <- sqrt(sum((point_perpendicular - mu_x)^2))

inv_Sigma <- solve(Sigma_x)
mahal_diag <- sqrt(sum((point_diagonal - mu_x) * (inv_Sigma %*% (point_diagonal - mu_x))))
mahal_perp <- sqrt(sum((point_perpendicular - mu_x) * (inv_Sigma %*% (point_perpendicular - mu_x))))

data.frame(
  point = c("(0.6, 0.6) - along diagonal", "(0.6, 0.4) - off diagonal"),
  euclidean_distance = round(c(euclidean_diag, euclidean_perp), 3),
  mahalanobis_distance = round(c(mahal_diag, mahal_perp), 3),
  prior_density = round(c(dmvnorm(point_diagonal, mu_x, Sigma_x),
                          dmvnorm(point_perpendicular, mu_x, Sigma_x)), 4)
)
```

The point along the diagonal has **smaller Mahalanobis distance** → **higher probability** under our correlated prior!
:::

**Step 3: Compute and plot the prior**

```{r gaussian-prior-2d}
#| fig-width: 5
#| fig-height: 5

# Compute prior density for every point in the grid
# Using a for loop (easier to understand than apply)

grid$prior <- numeric(nrow(grid))  # Create empty vector to store results

for (i in 1:nrow(grid)) {
  # Get the i-th point (x1, x2)
  x1 <- grid$x1[i]
  x2 <- grid$x2[i]
  point <- c(x1, x2)

  # Compute the Gaussian density at this point
  density <- dmvnorm(point, mu_x, Sigma_x)

  # Store the result
  grid$prior[i] <- density
}

# Plot as contour map (top view)
contour_plot <- ggplot(grid, aes(x1, x2)) +
  geom_contour_filled(aes(z = prior), bins = 12) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "white", linewidth = 0.8) +
  scale_fill_viridis_d(option = "plasma") +
  coord_fixed() +
  labs(title = "Contour (top view)",
       x = expression(x[1] ~ "(pixel 1)"),
       y = expression(x[2] ~ "(pixel 2)")) +
  theme_minimal() +
  theme(legend.position = "none")

print(contour_plot)
```

Now let's see the same prior as a **3D surface** (the actual bell hill):

```{r gaussian-prior-3d}
#| fig-width: 8
#| fig-height: 6

# Convert grid to matrix format for 3D plotting
# We need a matrix where z[i,j] = density at (x1[i], x2[j])
z_matrix <- matrix(grid$prior, nrow = 100, ncol = 100, byrow = FALSE)

# Create color gradient based on height
n_colors <- 50
colors <- hcl.colors(n_colors, palette = "Purple-Yellow")
z_normalized <- (z_matrix - min(z_matrix)) / (max(z_matrix) - min(z_matrix))
color_indices <- pmax(1, ceiling(z_normalized * n_colors))
facet_colors <- colors[color_indices]

# 3D surface plot
persp(x1_seq, x2_seq, z_matrix,
      theta = 35,              # rotation angle (left-right)
      phi = 25,                # tilt angle (up-down)
      col = facet_colors,      # color by height
      shade = 0.3,             # shading for depth
      border = NA,             # no grid lines
      xlab = "X1 (pixel 1)",
      ylab = "x2 (pixel 2)",
      zlab = "p(x)",
      main = "Prior p(x) as 3D Surface\nρ = 0.89 creates a ridge along the diagonal",
      ticktype = "detailed")
```

**The 3D view shows:**

-   The **peak** is at μ = (0.5, 0.5)
-   The hill is stretched into a **ridge along the diagonal** (because ρ = 0.89)
-   Points where x₁ ≈ x₂ have high probability (on the ridge)
-   Points where x₁ ≠ x₂ have low probability (off the ridge)

The prior is elongated along the diagonal because **correlated pixels** (x₁ ≈ x₂) are more probable than anti-correlated ones.

### The Likelihood: What does the data say?

We observe **y = 0.5** (a single cone measurement). The likelihood tells us: for each possible scene x, how probable is this measurement?

$$p(y|\mathbf{x}) = \mathcal{N}(y; \mathbf{C}\mathbf{x}, \sigma_y^2)$$

```{r gaussian-likelihood-2d}
#| fig-width: 5
#| fig-height: 5

# Observed measurement
y_obs <- 0.2

# Likelihood: p(y|x) = N(y; Cx, σ²)
# For each x, compute the Gaussian density of observing y_obs
grid$likelihood <- dnorm(y_obs,
                          mean = 0.5 * grid$x1 + 0.5 * grid$x2,  # Cx
                          sd = sigma_noise)

# Plot likelihood
ggplot(grid, aes(x1, x2)) +
  geom_contour_filled(aes(z = likelihood), bins = 12) +
  # Show the constraint line: 0.5*x1 + 0.5*x2 = y_obs
  geom_abline(slope = -1, intercept = 2*y_obs,
              linetype = "dashed", color = "white", linewidth = 1) +
  scale_fill_viridis_d(option = "viridis") +
  coord_fixed() +
  labs(title = "Likelihood p(y|x): What does data say?",
       subtitle = paste0("Observation y = ", y_obs, " constrains x to a RIDGE"),
       x = expression(x[1] ~ "(pixel 1)"),
       y = expression(x[2] ~ "(pixel 2)")) +
  theme_minimal() +
  theme(legend.position = "none")
```

**The ambiguity is visible!** The likelihood is a **ridge** along the constraint line $0.5x_1 + 0.5x_2 = 0.5$. Many different scenes produce the same measurement: - (0, 1), (0.5, 0.5), (1, 0) all give y = 0.5

### The Posterior: Combining prior and likelihood

Bayes' rule combines both:

$$p(\mathbf{x}|y) \propto p(y|\mathbf{x}) \cdot p(\mathbf{x})$$

```{r gaussian-posterior-2d}
#| fig-width: 5
#| fig-height: 5

# Posterior = Prior × Likelihood (unnormalized)
grid$posterior_unnorm <- grid$prior * grid$likelihood
grid$posterior <- grid$posterior_unnorm / sum(grid$posterior_unnorm)

# Find MAP estimate (maximum of posterior)
map_idx <- which.max(grid$posterior)
x_map <- c(grid$x1[map_idx], grid$x2[map_idx])

# Plot posterior
ggplot(grid, aes(x1, x2)) +
  geom_contour_filled(aes(z = posterior), bins = 12) +
  # Show constraint line from likelihood
  geom_abline(slope = -1, intercept = 2*y_obs,
              linetype = "dashed", color = "white", linewidth = 0.5, alpha = 0.5) +
  # Show diagonal from prior
  geom_abline(slope = 1, intercept = 0,
              linetype = "dotted", color = "white", linewidth = 0.5, alpha = 0.5) +
  # Mark MAP estimate
  geom_point(aes(x = x_map[1], y = x_map[2]),
             color = "red", size = 3, shape = 4, stroke = 2) +
  annotate("text", x = x_map[1] + 0.08, y = x_map[2] - 0.05,
           label = paste0("MAP: (", round(x_map[1], 2), ", ", round(x_map[2], 2), ")"),
           color = "white", size = 3.5, hjust = 0) +
  scale_fill_viridis_d(option = "magma") +
  coord_fixed() +
  labs(title = "Posterior p(x|y): Best guess about scene",
       subtitle = "Prior resolves the ambiguity in the likelihood",
       x = expression(x[1] ~ "(pixel 1)"),
       y = expression(x[2] ~ "(pixel 2)")) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Key insight:** The posterior is concentrated near (0.5, 0.5) because:

1\. The **likelihood** says x must lie on the ridge where 0.5x₁ + 0.5x₂ = 0.5

2\. The **prior** says x₁ ≈ x₂ is most probable

3\. The intersection picks out (0.5, 0.5) as the best estimate

### All three together

```{r gaussian-posterior-combined}
#| fig-width: 10
#| fig-height: 4

library(gridExtra)

# Normalize for comparable visualization
grid$prior_norm <- grid$prior / max(grid$prior)
grid$likelihood_norm <- grid$likelihood / max(grid$likelihood)
grid$posterior_norm <- grid$posterior / max(grid$posterior)

p1 <- ggplot(grid, aes(x1, x2)) +
  geom_contour_filled(aes(z = prior_norm), bins = 10) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "white") +
  scale_fill_viridis_d(option = "plasma") +
  coord_fixed() +
  labs(title = "Prior p(x)",
       subtitle = "x1 ≈ x2 expected",
       x = expression(x[1]), y = expression(x[2])) +
  theme_minimal() + theme(legend.position = "none")

p2 <- ggplot(grid, aes(x1, x2)) +
  geom_contour_filled(aes(z = likelihood_norm), bins = 10) +
  geom_abline(slope = -1, intercept = 2*y_obs, linetype = "dashed", color = "white") +
  scale_fill_viridis_d(option = "viridis") +
  coord_fixed() +
  labs(title = paste0("Likelihood p(y=", y_obs, "|x)"),
       subtitle = "Ridge of ambiguity",
       x = expression(x[1]), y = expression(x[2])) +
  theme_minimal() + theme(legend.position = "none")

p3 <- ggplot(grid, aes(x1, x2)) +
  geom_contour_filled(aes(z = posterior_norm), bins = 10) +
  geom_point(aes(x = x_map[1], y = x_map[2]), color = "red", size = 3, shape = 4, stroke = 2) +
  scale_fill_viridis_d(option = "magma") +
  coord_fixed() +
  labs(title = "Posterior p(x|y)",
       subtitle = paste0("MAP = (", round(x_map[1],2), ", ", round(x_map[2],2), ")"),
       x = expression(x[1]), y = expression(x[2])) +
  theme_minimal() + theme(legend.position = "none")

grid.arrange(p1, p2, p3, ncol = 3)
```

**The math in action:**

| Symbol | Meaning | Value |
|----|----|----|
| $\mathbf{x} = [x_1, x_2]^T$ | Scene (2 pixels) | Unknown |
| $\mathbf{C} = [0.5, 0.5]$ | Cone sensitivity matrix | $1 \times 2$ |
| $y$ | Measurement | 0.5 |
| $\boldsymbol{\mu}_x$ | Prior mean | \[0.5, 0.5\] |
| $\boldsymbol{\Sigma}_x$ | Prior covariance | Correlated (ρ=0.89) |
| $\sigma_y$ | Noise SD | 0.1 |
| $\hat{\mathbf{x}}_{MAP}$ | Best estimate | (`r round(x_map[1],2)`, `r round(x_map[2],2)`) |

### Thresholds and Ideal Observer Theory

**Threshold** = minimum difference needed to discriminate two stimuli.

::: callout-note
## What is a 2IFC Task?

**2IFC = Two-Interval Forced Choice**

A classic psychophysics experiment:

1.  **Interval 1:** Show stimulus A (e.g., blank gray screen)
2.  **Interval 2:** Show stimulus B (e.g., faint striped pattern)
3.  **Question:** "Which interval had the pattern?"

The observer **must choose** one (forced choice), even if unsure. By measuring how often they're correct at different pattern strengths, we find their **threshold** — the minimum detectable difference.
:::

**Step 1: Create the two stimuli**

```{r threshold-stimuli}
set.seed(42)  # For reproducibility

# We have 50 cones in a row
n_cones <- 50

# REFERENCE stimulus: uniform gray field
# Every cone gets the same mean excitation (100 photons)
ref_mean <- rep(100, n_cones)

cat("Reference stimulus (uniform):\n")
cat("All cones have mean =", ref_mean[1], "excitations\n")
```

```{r threshold-grating}
# COMPARISON stimulus: faint grating (striped pattern)
# A sine wave adds light/dark stripes

contrast <- 0.4  # 10% contrast (pretty faint!)

# Create sine wave pattern (2 full cycles across 50 cones)
grating <- sin(seq(0, 4*pi, length.out = n_cones))

# Apply contrast to the mean
# Where grating = +1: excitation = 100 × (1 + 0.1) = 110
# Where grating = -1: excitation = 100 × (1 - 0.1) = 90
# Where grating =  0: excitation = 100 × (1 + 0)   = 100
comp_mean <- ref_mean * (1 + contrast * grating)

cat("Comparison stimulus (grating):\n")
cat("Cone excitations vary from", round(min(comp_mean)), "to", round(max(comp_mean)), "\n")
cat("Contrast =", contrast * 100, "%\n")
```

```{r threshold-plot-means}
#| fig-width: 7
#| fig-height: 3

# Plot the MEAN responses (no noise yet)
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot(1:n_cones, ref_mean, type = "l", lwd = 2, col = "blue",
     ylim = c(80, 120), main = "Reference (uniform)",
     xlab = "Cone position", ylab = "Mean excitations")

plot(1:n_cones, comp_mean, type = "l", lwd = 2, col = "red",
     ylim = c(80, 120), main = "Comparison (grating)",
     xlab = "Cone position", ylab = "Mean excitations")
```

**Step 2: Add Poisson noise (what the cones actually report)**

The cones don't report the true mean — they report **noisy samples** from a Poisson distribution:

```{r threshold-add-noise}
# Simulate what the cones actually report (with Poisson noise)
# Each cone draws a random number from Poisson(mean)

ref_response <- rpois(n_cones, ref_mean)    # Noisy reference
comp_response <- rpois(n_cones, comp_mean)  # Noisy comparison

cat("Example: Cone 1\n")
cat("  Reference: mean =", ref_mean[1], ", actual response =", ref_response[1], "\n")
cat("  Comparison: mean =", round(comp_mean[1], 1), ", actual response =", comp_response[1], "\n")
```

**Step 3: The discrimination challenge**

Now look at what the observer actually sees (the noisy responses):

```{r threshold-task-plot}
#| fig-width: 7
#| fig-height: 4

response_df <- data.frame(
  cone = rep(1:n_cones, 2),
  response = c(ref_response, comp_response),
  stimulus = rep(c("Reference (uniform)", "Comparison (grating)"), each = n_cones)
)

ggplot(response_df, aes(cone, response, color = stimulus)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1) +
  labs(title = "2IFC Task: Can YOU tell which has the grating?",
       subtitle = "Both have Poisson noise — the pattern is hidden in the noise!",
       x = "Cone position", y = "Excitations") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**The problem:** The grating pattern is **buried in noise**! The brain must figure out which stimulus has the pattern based on these noisy measurements.

This is why **thresholds exist** — if the contrast is too low, the pattern disappears into the noise and becomes undetectable.

### Ideal Observer Decision Rule

The **ideal observer** uses Bayes' rule to make optimal decisions.

::: callout-note
## What is an Ideal Observer?

An **ideal observer** is a theoretical "perfect" decision-maker that:

1.  Knows the true statistics of the stimuli (means, noise distributions)
2.  Uses Bayes' rule to make the mathematically optimal decision
3.  Represents the **best possible performance** given the available information

Comparing humans to ideal observers tells us how much information is "lost" in neural processing.
:::

**Step 1: Setup the simple discrimination task**

Let's simplify to a single cone comparing two intensities:

```{r ideal-setup}
set.seed(123)

# Two stimuli with slightly different means
mean_ref <- 100   # Reference: dimmer
mean_comp <- 105  # Comparison: brighter (5% difference)

cat("Reference stimulus: mean =", mean_ref, "excitations\n")
cat("Comparison stimulus: mean =", mean_comp, "excitations\n")
cat("Difference:", mean_comp - mean_ref, "excitations (",
    round((mean_comp - mean_ref) / mean_ref * 100, 1), "% )\n")
```

**Step 2: Walk through ONE trial**

```{r ideal-one-trial}
# On each trial, stimuli are shown in random order
# The observer must guess which order they were shown

# Let's say the TRUE order is: reference first, comparison second
true_order <- 1  # 1 = ref first, 2 = comp first

# What the cone actually reports (noisy Poisson samples)
obs_1 <- rpois(1, mean_ref)   # First interval: drew from reference
obs_2 <- rpois(1, mean_comp)  # Second interval: drew from comparison

cat("TRUE order: Reference first, Comparison second\n\n")
cat("Interval 1: observed", obs_1, "excitations (true mean was", mean_ref, ")\n")
cat("Interval 2: observed", obs_2, "excitations (true mean was", mean_comp, ")\n")
```

**Step 3: The ideal observer's decision rule**

The ideal observer asks: "Which ordering is more likely to have produced these observations?"

```{r ideal-likelihood}
# Hypothesis 1: Reference was first (obs_1 ~ Poisson(100), obs_2 ~ Poisson(105))
# Hypothesis 2: Comparison was first (obs_1 ~ Poisson(105), obs_2 ~ Poisson(100))

# Compute log-likelihood for each hypothesis
# P(obs_1 | mean) × P(obs_2 | mean) ... use log for numerical stability

# If reference was first:
ll_ref_first <- dpois(obs_1, mean_ref, log = TRUE) +
                dpois(obs_2, mean_comp, log = TRUE)

# If comparison was first:
ll_comp_first <- dpois(obs_1, mean_comp, log = TRUE) +
                 dpois(obs_2, mean_ref, log = TRUE)

cat("Log-likelihood if REFERENCE first:", round(ll_ref_first, 3), "\n")
cat("Log-likelihood if COMPARISON first:", round(ll_comp_first, 3), "\n")
```

```{r ideal-decision}
# Decision variable: difference in log-likelihoods
delta <- ll_ref_first - ll_comp_first

cat("\nDecision variable δ =", round(delta, 3), "\n\n")

# Decision rule:
# If δ ≥ 0: more likely that reference was first → choose "ref first"
# If δ < 0: more likely that comparison was first → choose "comp first"

if (delta >= 0) {
  decision <- 1
  cat("δ ≥ 0, so decide: REFERENCE was first\n")
} else {
  decision <- 2
  cat("δ < 0, so decide: COMPARISON was first\n")
}

# Was the decision correct?
if (decision == true_order) {
  cat("✓ CORRECT!\n")
} else {
  cat("✗ WRONG!\n")
}
```

**Step 4: Run many trials to measure performance**

```{r ideal-simulation}
n_trials <- 500
correct <- 0
delta_values <- c()

for (trial in 1:n_trials) {

  # Random order: 1 = ref first, 2 = comp first
  true_order <- sample(1:2, 1)

  # Generate noisy observations based on true order
  if (true_order == 1) {
    # Reference first, comparison second
    obs_1 <- rpois(1, mean_ref)
    obs_2 <- rpois(1, mean_comp)
  } else {
    # Comparison first, reference second
    obs_1 <- rpois(1, mean_comp)
    obs_2 <- rpois(1, mean_ref)
  }

  # Compute log-likelihoods
  ll_ref_first <- dpois(obs_1, mean_ref, log = TRUE) +
                  dpois(obs_2, mean_comp, log = TRUE)
  ll_comp_first <- dpois(obs_1, mean_comp, log = TRUE) +
                   dpois(obs_2, mean_ref, log = TRUE)

  # Decision variable
  delta <- ll_ref_first - ll_comp_first
  delta_values <- c(delta_values, delta)

  # Make decision
  decision <- ifelse(delta >= 0, 1, 2)

  # Track correctness
  if (decision == true_order) {
    correct <- correct + 1
  }
}

percent_correct <- correct / n_trials * 100
cat("Results after", n_trials, "trials:\n")
cat("Correct:", correct, "/", n_trials, "\n")
cat("Percent correct:", round(percent_correct, 1), "%\n")
```

**Step 5: Visualize the decision variable distribution**

```{r ideal-plot}
#| fig-width: 7
#| fig-height: 4

delta_df <- data.frame(delta = delta_values)

ggplot(delta_df, aes(delta)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = 0.3, y = 35, label = "Decision\nboundary\n(δ = 0)",
           color = "red", hjust = 0) +
  annotate("text", x = -1.5, y = 30, label = "Choose\n'comp first'", color = "gray40") +
  annotate("text", x = 1.5, y = 30, label = "Choose\n'ref first'", color = "gray40") +
  labs(title = paste0("Ideal Observer Performance: ", round(percent_correct, 1), "% correct"),
       subtitle = paste0("Discriminating ", mean_ref, " vs ", mean_comp, " mean excitations"),
       x = "Decision variable (δ)", y = "Count") +
  theme_minimal()
```

**Key insight:** The decision variable δ is sometimes positive, sometimes negative (due to noise). When it crosses the wrong side of 0, the observer makes an error. The closer the two means, the more overlap → more errors → lower percent correct.

### Contrast Sensitivity Function

::: callout-note
## What is Spatial Frequency?

**Spatial frequency** = how many light/dark cycles fit in one degree of visual angle

```         
Low frequency (1 cpd):     High frequency (10 cpd):
█████░░░░░█████░░░░░       █░█░█░█░█░█░█░█░█░█░
     1 cycle                    10 cycles
←───── 1 degree ─────→     ←───── 1 degree ─────→
```

-   **Low frequency**: Wide stripes (easy to see even when blurry)
-   **High frequency**: Narrow stripes (need sharp vision)
-   **Units**: cycles per degree (cpd)
:::

**Step 1: What do different spatial frequencies look like?**

```{r csf-gratings}
#| fig-width: 10
#| fig-height: 3

# Create gratings at different spatial frequencies
par(mfrow = c(1, 4), mar = c(2, 1, 3, 1))

for (freq in c(1, 3, 10, 30)) {
  # Create a sinusoidal grating
  x <- seq(0, 1, length.out = 200)
  grating <- matrix(rep(sin(2 * pi * freq * x), 100), nrow = 100, byrow = TRUE)

  image(grating, col = gray.colors(100), axes = FALSE,
        main = paste0(freq, " cycles/degree"))
}
```

**Step 2: What is contrast threshold?**

```{r csf-threshold-concept}
#| fig-width: 8
#| fig-height: 3

# Show gratings at decreasing contrast
par(mfrow = c(1, 4), mar = c(2, 1, 3, 1))

freq <- 5  # Fixed frequency
contrasts <- c(1.0, 0.3, 0.1, 0.03)

for (c in contrasts) {
  x <- seq(0, 1, length.out = 200)
  # Grating with contrast c: values range from 0.5-c/2 to 0.5+c/2
  grating <- 0.5 + c/2 * sin(2 * pi * freq * x)
  grating_mat <- matrix(rep(grating, 100), nrow = 100, byrow = TRUE)

  image(grating_mat, col = gray.colors(100), axes = FALSE, zlim = c(0, 1),
        main = paste0("Contrast = ", c * 100, "%"))
}
```

-   **High contrast (100%)**: Easy to see
-   **Low contrast (3%)**: Hard to see — near threshold!

**Threshold** = the minimum contrast needed to detect the grating (e.g., 75% correct in 2IFC)

**Step 3: What is contrast sensitivity?**

```{r csf-sensitivity-concept}
# Sensitivity is the INVERSE of threshold

# Example:
threshold_low_freq <- 0.01   # 1% contrast needed → easy to detect
threshold_high_freq <- 0.10  # 10% contrast needed → harder to detect

sensitivity_low_freq <- 1 / threshold_low_freq    # = 100
sensitivity_high_freq <- 1 / threshold_high_freq  # = 10

cat("Low frequency (2 cpd):\n")
cat("  Threshold =", threshold_low_freq * 100, "% → Sensitivity =", sensitivity_low_freq, "\n\n")

cat("High frequency (20 cpd):\n")
cat("  Threshold =", threshold_high_freq * 100, "% → Sensitivity =", sensitivity_high_freq, "\n\n")

cat("Higher sensitivity = lower threshold = easier to see!\n")
```

**Step 4: The human CSF curve**

```{r csf-curve}
#| fig-width: 7
#| fig-height: 5

# Spatial frequencies to test (0.5 to 50 cycles per degree)
spatial_freq <- seq(0.5, 50, length.out = 100)

# Human CSF model (approximate)
# - Rises at low frequencies (up to peak at ~3-6 cpd)
# - Falls at high frequencies (optical blur, cone spacing)
csf <- 300 * (spatial_freq / 4)^0.8 * exp(-(spatial_freq / 20)^1.5)

# Find the peak
peak_freq <- spatial_freq[which.max(csf)]
peak_sens <- max(csf)

cat("Peak sensitivity at", round(peak_freq, 1), "cycles/degree\n")
cat("Peak sensitivity =", round(peak_sens), "(threshold =", round(1/peak_sens * 100, 2), "%)\n")
```

```{r csf-plot}
#| fig-width: 8
#| fig-height: 5

ggplot(data.frame(sf = spatial_freq, sensitivity = csf), aes(sf, sensitivity)) +
  geom_line(linewidth = 1.5, color = "darkblue") +
  geom_vline(xintercept = peak_freq, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = peak_freq * 1.3, y = peak_sens * 0.9,
           label = paste0("Peak at ~", round(peak_freq), " cpd"), color = "red") +
  annotate("text", x = 1, y = 50, label = "Low freq:\nneural limits", hjust = 0, size = 3) +
  annotate("text", x = 25, y = 20, label = "High freq:\noptical blur +\ncone spacing", hjust = 0, size = 3) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Human Contrast Sensitivity Function (CSF)",
       subtitle = "We're most sensitive to medium spatial frequencies (3-6 cpd)",
       x = "Spatial frequency (cycles/degree)",
       y = "Contrast sensitivity (1/threshold)") +
  theme_minimal()
```

**Step 5: Why this shape?**

```{r csf-explanation}
#| fig-width: 9
#| fig-height: 4

# Visual explanation of the CSF shape
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

# Low frequencies - neural limits
x <- seq(0, 1, length.out = 200)
grating_low <- 0.5 + 0.3 * sin(2 * pi * 1 * x)
plot(x * 100, grating_low, type = "l", lwd = 2, ylim = c(0, 1),
     main = "Low Frequency (1 cpd)\nLIMITED BY: Neural processing",
     xlab = "Position", ylab = "Intensity", col = "blue")
text(50, 0.15, "Brain computes DIFFERENCES\nUniform regions give weak signal", cex = 0.8)

# Medium frequencies - optimal
grating_mid <- 0.5 + 0.3 * sin(2 * pi * 5 * x)
plot(x * 100, grating_mid, type = "l", lwd = 2, ylim = c(0, 1),
     main = "Medium Frequency (5 cpd)\nOPTIMAL: Best match to cones",
     xlab = "Position", ylab = "Intensity", col = "darkgreen")
text(50, 0.15, "Good balance of\ncontrast and sharpness", cex = 0.8)

# High frequencies - optical limits
grating_high <- 0.5 + 0.3 * sin(2 * pi * 20 * x)
# Blur it to simulate optics
blurred <- stats::filter(grating_high, rep(1/15, 15), sides = 2)
blurred[is.na(blurred)] <- 0.5
plot(x * 100, blurred, type = "l", lwd = 2, ylim = c(0, 1),
     main = "High Frequency (20 cpd)\nLIMITED BY: Optical blur",
     xlab = "Position", ylab = "Intensity", col = "red")
lines(x * 100, grating_high, lwd = 1, col = "pink", lty = 2)
legend("bottomright", c("After blur", "Original"), col = c("red", "pink"),
       lty = c(1, 2), lwd = c(2, 1), bty = "n", cex = 0.8)
text(50, 0.15, "Optics blur fine details\nCones too sparse to sample", cex = 0.8)
```

**Summary: Why the inverted-U shape?**

| Region | Spatial Freq | Limiting Factor |
|----|----|----|
| **Left (low freq)** | \< 2 cpd | Neural: brain computes derivatives, ignores uniform regions |
| **Peak (mid freq)** | 3-6 cpd | Optimal: matches cone spacing, survives blur |
| **Right (high freq)** | \> 10 cpd | Optical: blur + cone spacing can't resolve fine stripes |

**Ideal observer analysis** helps explain the CSF:

-   **High frequency rolloff**: optical blur + cone sampling limits
-   **Low frequency rolloff**: neural factors (lateral inhibition, computes derivatives)
-   **Gap from ideal**: efficiency loss from additional noise, stimulus uncertainty

### Fitting the CSF Model to Real Data

How do we find the best parameters for real experimental data? **Nonlinear least squares!**

**Step 1: The CSF model with parameters**

Our model has 4 free parameters:

$$\text{Sensitivity}(f) = A \cdot \left(\frac{f}{f_{peak}}\right)^{s_1} \cdot \exp\left(-\left(\frac{f}{f_{cut}}\right)^{s_2}\right)$$

```{r csf-model-function}
# Define the CSF model as a function with parameters
csf_model <- function(freq, A, f_peak, s1, f_cut, s2) {
  # A      = amplitude (peak sensitivity)
  # f_peak = frequency scaling for rise
  # s1     = low-frequency slope
  # f_cut  = high-frequency cutoff
  # s2     = high-frequency slope

  sensitivity <- A * (freq / f_peak)^s1 * exp(-(freq / f_cut)^s2)
  return(sensitivity)
}

cat("Parameters:\n")
cat("  A      = overall amplitude (peak height)\n")
cat("  f_peak = controls where sensitivity rises\n")
cat("  s1     = low-frequency slope (how fast it rises)\n")
cat("  f_cut  = high-frequency cutoff\n")
cat("  s2     = high-frequency slope (how fast it falls)\n")
```

::: {.callout-note title="Understanding the CSF Equation"}
The CSF model has two main parts multiplied together:

$$\text{Sensitivity}(f) = \underbrace{A \cdot \left(\frac{f}{f_{peak}}\right)^{s_1}}_{\text{Low-frequency rise}} \cdot \underbrace{\exp\left(-\left(\frac{f}{f_{cut}}\right)^{s_2}\right)}_{\text{High-frequency fall}}$$

**Part 1: Rising part** $(f/f_{peak})^{s_1}$

\- When $f < f_{peak}$: this is less than 1, so sensitivity is lower

\- When $f = f_{peak}$: this equals 1

\- $s_1$ controls how steep the rise is

**Part 2: Falling part** $\exp(-(f/f_{cut})^{s_2})$

\- When $f \ll f_{cut}$: exponent is near 0, so this equals \~1

\- When $f = f_{cut}$: this equals $e^{-1} \approx 0.37$

\- When $f > f_{cut}$: this drops rapidly toward 0

\- $s_2$ controls how steep the fall is
:::

**Let's visualize what each parameter does:**

```{r csf-param-A}
#| fig-cap: "Effect of A (amplitude): Scales the entire curve up or down"
#| fig-width: 8
#| fig-height: 4

# Fixed parameters
f_peak <- 4
s1 <- 0.9
f_cut <- 18
s2 <- 1.4

# Frequency range
freq <- seq(0.5, 50, length.out = 200)

# Vary A
A_values <- c(100, 200, 300, 400)

par(mfrow = c(1, 2))

# Linear scale
plot(NULL, xlim = c(0.5, 50), ylim = c(0, 600),
     xlab = "Spatial Frequency (cycles/degree)", ylab = "Sensitivity",
     main = "Effect of A (Amplitude) - Linear Scale")

colors <- c("blue", "green", "orange", "red")
for (i in seq_along(A_values)) {
  A <- A_values[i]
  sensitivity <- A * (freq / f_peak)^s1 * exp(-(freq / f_cut)^s2)
  lines(freq, sensitivity, col = colors[i], lwd = 2)
}
legend("topright", legend = paste("A =", A_values), col = colors, lwd = 2)

# Log scale
plot(NULL, xlim = c(0.5, 50), ylim = c(1, 500), log = "xy",
     xlab = "Spatial Frequency (cycles/degree)", ylab = "Sensitivity",
     main = "Effect of A (Amplitude) - Log Scale")

for (i in seq_along(A_values)) {
  A <- A_values[i]
  sensitivity <- A * (freq / f_peak)^s1 * exp(-(freq / f_cut)^s2)
  lines(freq, sensitivity, col = colors[i], lwd = 2)
}
legend("bottomright", legend = paste("A =", A_values), col = colors, lwd = 2)

cat("A simply scales the height of the curve.\n")
cat("Higher A = more sensitive overall (lower thresholds).\n")
```

```{r csf-param-fpeak}
#| fig-cap: "Effect of f_peak: Controls where the low-frequency rise reaches maximum"
#| fig-width: 8
#| fig-height: 4

# Fixed parameters
A <- 280
s1 <- 0.9
f_cut <- 18
s2 <- 1.4

# Vary f_peak
f_peak_values <- c(2, 4, 6, 8)

par(mfrow = c(1, 2))

# Linear scale
plot(NULL, xlim = c(0.5, 50), ylim = c(0, 850),
     xlab = "Spatial Frequency (cycles/degree)", ylab = "Sensitivity",
     main = "Effect of f_peak - Linear Scale")

colors <- c("blue", "green", "orange", "red")
for (i in seq_along(f_peak_values)) {
  f_peak <- f_peak_values[i]
  sensitivity <- A * (freq / f_peak)^s1 * exp(-(freq / f_cut)^s2)
  lines(freq, sensitivity, col = colors[i], lwd = 2)
}
legend("topright", legend = paste("f_peak =", f_peak_values), col = colors, lwd = 2)
abline(v = f_peak_values, col = colors, lty = 2, lwd = 1)

# Log scale
plot(NULL, xlim = c(0.5, 50), ylim = c(1, 600), log = "xy",
     xlab = "Spatial Frequency (cycles/degree)", ylab = "Sensitivity",
     main = "Effect of f_peak - Log Scale")

for (i in seq_along(f_peak_values)) {
  f_peak <- f_peak_values[i]
  sensitivity <- A * (freq / f_peak)^s1 * exp(-(freq / f_cut)^s2)
  lines(freq, sensitivity, col = colors[i], lwd = 2)
}
legend("bottomright", legend = paste("f_peak =", f_peak_values), col = colors, lwd = 2)

cat("f_peak controls where the rising part reaches its contribution.\n")
cat("Larger f_peak shifts the curve to the right.\n")
```

```{r csf-param-s1}
#| fig-cap: "Effect of s1: Controls the steepness of the low-frequency rise"
#| fig-width: 8
#| fig-height: 4

# Fixed parameters
A <- 280
f_peak <- 4
f_cut <- 18
s2 <- 1.4

# Vary s1 (low-frequency slope)
s1_values <- c(0.3, 0.6, 0.9, 1.2)

par(mfrow = c(1, 2))

# Linear scale
plot(NULL, xlim = c(0.5, 50), ylim = c(0, 800),
     xlab = "Spatial Frequency (cycles/degree)", ylab = "Sensitivity",
     main = "Effect of s1 (low-freq slope) - Linear")

colors <- c("blue", "green", "orange", "red")
for (i in seq_along(s1_values)) {
  s1 <- s1_values[i]
  sensitivity <- A * (freq / f_peak)^s1 * exp(-(freq / f_cut)^s2)
  lines(freq, sensitivity, col = colors[i], lwd = 2)
}
legend("topright", legend = paste("s1 =", s1_values), col = colors, lwd = 2)

# Log scale - this shows the slope more clearly!
plot(NULL, xlim = c(0.5, 50), ylim = c(1, 700), log = "xy",
     xlab = "Spatial Frequency (cycles/degree)", ylab = "Sensitivity",
     main = "Effect of s1 - Log Scale (slope visible!)")

for (i in seq_along(s1_values)) {
  s1 <- s1_values[i]
  sensitivity <- A * (freq / f_peak)^s1 * exp(-(freq / f_cut)^s2)
  lines(freq, sensitivity, col = colors[i], lwd = 2)
}
legend("bottomright", legend = paste("s1 =", s1_values), col = colors, lwd = 2)

cat("s1 controls how steep the left side of the curve is.\n")
cat("On a log-log plot, s1 IS the slope at low frequencies!\n")
cat("s1 = 0 would be flat, s1 = 1 would be proportional to frequency.\n")
```

```{r csf-param-fcut}
#| fig-cap: "Effect of f_cut: Controls where the high-frequency fall-off begins"
#| fig-width: 8
#| fig-height: 4

# Fixed parameters
A <- 280
f_peak <- 4
s1 <- 0.9
s2 <- 1.4

# Vary f_cut (high-frequency cutoff)
f_cut_values <- c(10, 15, 20, 25)

par(mfrow = c(1, 2))

# Linear scale
plot(NULL, xlim = c(0.5, 50), ylim = c(0, 750),
     xlab = "Spatial Frequency (cycles/degree)", ylab = "Sensitivity",
     main = "Effect of f_cut - Linear Scale")

colors <- c("blue", "green", "orange", "red")
for (i in seq_along(f_cut_values)) {
  f_cut <- f_cut_values[i]
  sensitivity <- A * (freq / f_peak)^s1 * exp(-(freq / f_cut)^s2)
  lines(freq, sensitivity, col = colors[i], lwd = 2)
}
legend("topright", legend = paste("f_cut =", f_cut_values), col = colors, lwd = 2)
abline(v = f_cut_values, col = colors, lty = 2, lwd = 1)

# Log scale
plot(NULL, xlim = c(0.5, 50), ylim = c(1, 700), log = "xy",
     xlab = "Spatial Frequency (cycles/degree)", ylab = "Sensitivity",
     main = "Effect of f_cut - Log Scale")

for (i in seq_along(f_cut_values)) {
  f_cut <- f_cut_values[i]
  sensitivity <- A * (freq / f_peak)^s1 * exp(-(freq / f_cut)^s2)
  lines(freq, sensitivity, col = colors[i], lwd = 2)
}
legend("bottomleft", legend = paste("f_cut =", f_cut_values), col = colors, lwd = 2)

cat("f_cut controls WHERE the curve starts falling sharply.\n")
cat("Larger f_cut = sensitivity extends to higher frequencies.\n")
cat("Dashed lines show the f_cut values.\n")
```

```{r csf-param-s2}
#| fig-cap: "Effect of s2: Controls the steepness of the high-frequency fall"
#| fig-width: 8
#| fig-height: 4

# Fixed parameters
A <- 280
f_peak <- 4
s1 <- 0.9
f_cut <- 18

# Vary s2 (high-frequency slope)
s2_values <- c(0.8, 1.2, 1.6, 2.0)

par(mfrow = c(1, 2))

# Linear scale
plot(NULL, xlim = c(0.5, 50), ylim = c(0, 850),
     xlab = "Spatial Frequency (cycles/degree)", ylab = "Sensitivity",
     main = "Effect of s2 (high-freq slope) - Linear")

colors <- c("blue", "green", "orange", "red")
for (i in seq_along(s2_values)) {
  s2 <- s2_values[i]
  sensitivity <- A * (freq / f_peak)^s1 * exp(-(freq / f_cut)^s2)
  lines(freq, sensitivity, col = colors[i], lwd = 2)
}
legend("topright", legend = paste("s2 =", s2_values), col = colors, lwd = 2)

# Log scale
plot(NULL, xlim = c(0.5, 50), ylim = c(0.01, 700), log = "xy",
     xlab = "Spatial Frequency (cycles/degree)", ylab = "Sensitivity",
     main = "Effect of s2 - Log Scale")

for (i in seq_along(s2_values)) {
  s2 <- s2_values[i]
  sensitivity <- A * (freq / f_peak)^s1 * exp(-(freq / f_cut)^s2)
  lines(freq, sensitivity, col = colors[i], lwd = 2)
}
legend("bottomleft", legend = paste("s2 =", s2_values), col = colors, lwd = 2)

cat("s2 controls how SHARPLY sensitivity drops at high frequencies.\n")
cat("Larger s2 = steeper drop-off = sharper cutoff.\n")
cat("This relates to optical blur and cone spacing limits.\n")
```

::: {.callout-tip title="Summary: What Each Parameter Controls"}
| Parameter | Name | What it controls | Physical interpretation |
|----|----|----|----|
| **A** | Amplitude | Overall height | Overall neural sensitivity |
| **f_peak** | Peak frequency | Left-right position of rise | Related to neural tuning |
| **s1** | Low-freq slope | Steepness at low frequencies | Lateral inhibition effects |
| **f_cut** | Cutoff frequency | Where drop-off begins | Optical blur + cone spacing |
| **s2** | High-freq slope | Steepness of drop-off | Sharpness of resolution limit |

**Fitting strategy:** 1. Start with **A** to get the overall scale right 2. Adjust **f_cut** to match where sensitivity drops 3. Adjust **s2** to get the steepness of the drop 4. Adjust **f_peak** and **s1** to fit the low-frequency side
:::

```{r csf-param-combined}
#| fig-cap: "How the two parts combine to create the CSF"
#| fig-width: 10
#| fig-height: 4

# Parameters
A <- 280
f_peak <- 4
s1 <- 0.9
f_cut <- 18
s2 <- 1.4

freq <- seq(0.5, 50, length.out = 200)

# Compute each part separately
rising_part <- A * (freq / f_peak)^s1
falling_part <- exp(-(freq / f_cut)^s2)
combined <- rising_part * falling_part

par(mfrow = c(1, 3))

# Part 1: Rising
plot(freq, rising_part, type = "l", col = "blue", lwd = 2,
     xlab = "Frequency", ylab = "Value",
     main = "Part 1: Rising Part\nA × (f/f_peak)^s1",
     xlim = c(0, 50), ylim = c(0, max(rising_part)))
abline(h = A, lty = 2, col = "gray")
text(40, A + 20, "A", col = "gray")

# Part 2: Falling
plot(freq, falling_part, type = "l", col = "red", lwd = 2,
     xlab = "Frequency", ylab = "Value",
     main = "Part 2: Falling Part\nexp(-(f/f_cut)^s2)",
     xlim = c(0, 50), ylim = c(0, 1))
abline(v = f_cut, lty = 2, col = "gray")
abline(h = exp(-1), lty = 2, col = "gray")
text(f_cut + 3, 0.1, "f_cut", col = "gray")
text(5, exp(-1) + 0.05, "1/e", col = "gray")

# Combined
plot(freq, combined, type = "l", col = "purple", lwd = 3,
     xlab = "Frequency", ylab = "Sensitivity",
     main = "Combined CSF\n(Part 1) × (Part 2)",
     xlim = c(0, 50), ylim = c(0, max(combined) * 1.1))

# Find and mark the peak
peak_idx <- which.max(combined)
peak_freq <- freq[peak_idx]
peak_sens <- combined[peak_idx]
points(peak_freq, peak_sens, pch = 19, col = "purple", cex = 2)
text(peak_freq + 5, peak_sens, sprintf("Peak at %.1f cpd", peak_freq))

cat("The CSF is the PRODUCT of:\n")
cat("  1. A rising power function (blue)\n")
cat("  2. A falling exponential (red)\n")
cat("The product creates the inverted-U shape!\n")
```

### How Real Behavioral Data Becomes Sensitivity

::: {.callout-important title="From Behavior to Sensitivity"}
In a real experiment, you don't measure sensitivity directly. You measure **threshold contrast** through a behavioral task:

**Step 1: Run 2IFC trials at one spatial frequency** - Present grating vs. uniform field (random order) - Subject says which interval had the grating - Vary contrast using a staircase procedure

**Step 2: Find threshold** - Threshold = contrast where subject is \~75% correct - Example: At 4 cpd, threshold might be 0.5% contrast

**Step 3: Convert to sensitivity** $$\text{Sensitivity} = \frac{1}{\text{Threshold Contrast}}$$ - If threshold = 0.005 (0.5%), then sensitivity = 200 - If threshold = 0.02 (2%), then sensitivity = 50

**Repeat for each spatial frequency!**
:::

```{r csf-threshold-to-sensitivity}
#| fig-cap: "How thresholds become sensitivity values"
#| fig-width: 10
#| fig-height: 4

# Simulate a psychometric function for ONE spatial frequency
# This is what you'd get from many trials at different contrasts

set.seed(123)

# Contrasts tested (log-spaced is typical)
contrasts <- c(0.001, 0.002, 0.004, 0.008, 0.016, 0.032, 0.064)

# True threshold for this frequency (4 cpd)
true_threshold <- 0.02  # 0.5% contrast

# Psychometric function (Weibull is common)
# Probability correct as function of contrast
weibull_psychometric <- function(contrast, threshold, slope = 2) {
  # gamma = 0.5 (chance level for 2IFC)
  # lambda = 0.02 (lapse rate)
  gamma <- 0.5
  lambda <- 0.02
  p_correct <- gamma + (1 - gamma - lambda) * (1 - exp(-(contrast/threshold)^slope))
  return(p_correct)
}

# True probability correct at each contrast
p_correct <- weibull_psychometric(contrasts, true_threshold)

# Simulate 50 trials at each contrast
n_trials <- 50
n_correct <- rbinom(length(contrasts), n_trials, p_correct)
observed_p <- n_correct / n_trials

par(mfrow = c(1, 3))

# Panel 1: Psychometric function
plot(contrasts, observed_p, pch = 19, cex = 1.5,
     xlab = "Contrast", ylab = "Proportion Correct",
     main = "Step 1: Psychometric Function\n(one spatial frequency)",
     xlim = c(0, 0.07), ylim = c(0.4, 1))

# Smooth curve
contrast_smooth <- seq(0.0005, 0.07, length.out = 100)
lines(contrast_smooth, weibull_psychometric(contrast_smooth, true_threshold),
      col = "blue", lwd = 2)

# Mark threshold (75% correct for 2IFC)
abline(h = 0.75, lty = 2, col = "red")
abline(v = true_threshold, lty = 2, col = "red")
points(true_threshold, 0.75, pch = 4, cex = 2, col = "red", lwd = 3)
text(true_threshold + 0.015, 0.75, "Threshold\n(75% correct)", col = "red")
text(true_threshold, 0.45, sprintf("%.1f%%", true_threshold * 100), col = "red")

# Panel 2: Thresholds at multiple frequencies
# Simulated threshold data at different spatial frequencies
frequencies <- c(0.5, 1, 2, 4, 8, 16, 32)
# Thresholds follow inverse of CSF (with some noise)
true_sensitivities <- 280 * (frequencies / 4)^0.9 * exp(-(frequencies / 18)^1.4)
true_thresholds <- 1 / true_sensitivities
# Add measurement noise
measured_thresholds <- true_thresholds * exp(rnorm(length(frequencies), 0, 0.1))

plot(frequencies, measured_thresholds * 100, pch = 19, cex = 1.5,
     xlab = "Spatial Frequency (cpd)", ylab = "Threshold Contrast (%)",
     main = "Step 2: Measure Threshold\nat Each Frequency",
     log = "xy")
lines(frequencies, measured_thresholds * 100, col = "blue", lwd = 1)

cat("Lower threshold = easier to see = MORE sensitive\n")

# Panel 3: Convert to sensitivity
measured_sensitivity <- 1 / measured_thresholds

plot(frequencies, measured_sensitivity, pch = 19, cex = 1.5,
     xlab = "Spatial Frequency (cpd)", ylab = "Sensitivity (1/threshold)",
     main = "Step 3: Sensitivity = 1/Threshold\n(This is the CSF!)",
     log = "xy")
lines(frequencies, measured_sensitivity, col = "blue", lwd = 1)

cat("\nSensitivity = 1 / Threshold Contrast\n")
cat("High sensitivity = low threshold = easy to detect\n")
```

```{r csf-real-data-example}
# Show what real experimental data looks like
cat("Example: Real data from one subject\n")
cat("=========================================\n\n")

example_data <- data.frame(
  Frequency_cpd = c(0.5, 1, 2, 4, 8, 16, 32),
  Threshold_pct = c(2.1, 0.8, 0.4, 0.35, 0.5, 1.8, 12.5),
  Trials_per_threshold = rep(100, 7)
)

example_data$Sensitivity <- 100 / example_data$Threshold_pct

print(example_data)

cat("\nKey insight: Each 'sensitivity' value represents ~100 trials!\n")
cat("The CSF curve is fitted to these ~7-10 data points.\n")
```

**Step 2: Simulate "real" experimental data**

In a real experiment, you would measure thresholds at several spatial frequencies. Let's simulate this:

```{r csf-simulate-data}
set.seed(42)

# "True" parameters (unknown to the fitting procedure)
true_A <- 280
true_f_peak <- 4
true_s1 <- 0.9
true_f_cut <- 18
true_s2 <- 1.4

# Spatial frequencies tested in experiment (typically 8-12 values)
test_frequencies <- c(0.5, 1, 2, 4, 6, 8, 12, 16, 24, 32)

# "True" sensitivity at each frequency
true_sensitivity <- csf_model(test_frequencies, true_A, true_f_peak, true_s1, true_f_cut, true_s2)

# Add measurement noise (real data is noisy!)
# Use multiplicative noise (log-normal) since sensitivity spans orders of magnitude
noise <- exp(rnorm(length(test_frequencies), mean = 0, sd = 0.15))
observed_sensitivity <- true_sensitivity * noise

# Create data frame
exp_data <- data.frame(
  frequency = test_frequencies,
  sensitivity = observed_sensitivity
)

cat("Simulated experimental data:\n")
print(round(exp_data, 1))
```

```{r csf-plot-data}
#| fig-width: 7
#| fig-height: 4

ggplot(exp_data, aes(frequency, sensitivity)) +
  geom_point(size = 3, color = "red") +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Simulated Experimental Data",
       subtitle = "10 threshold measurements (with noise)",
       x = "Spatial frequency (cpd)",
       y = "Contrast sensitivity") +
  theme_minimal()
```

**Step 3: Fit the model using optimization**

### How does `optim()` work?

::: {.callout-note title="The Idea Behind Optimization"}
**Goal:** Find parameter values that make our model best match the data.

**How:** 1. Start with initial guesses for parameters 2. Compute how "wrong" the model is (the **loss**) 3. Adjust parameters to reduce the loss 4. Repeat until loss stops decreasing

It's like finding the lowest point in a valley while blindfolded - you feel which direction goes downhill and step that way!
:::

```{r optim-explanation}
#| fig-cap: "How optimization finds the best parameters"
#| fig-width: 10
#| fig-height: 4

# Let's visualize optimization with a SIMPLE 1-parameter example first
# Question: What value of 'A' best fits our data (holding other params fixed)?

# Fix all parameters except A
fixed_f_peak <- 4
fixed_s1 <- 0.9
fixed_f_cut <- 18
fixed_s2 <- 1.4

# Try many values of A and compute the loss for each
A_values <- seq(100, 500, by = 5)

# Compute loss (SSE) for each A value
losses <- numeric(length(A_values))
for (i in seq_along(A_values)) {
  A <- A_values[i]
  predicted <- A * (exp_data$frequency / fixed_f_peak)^fixed_s1 *
               exp(-(exp_data$frequency / fixed_f_cut)^fixed_s2)
  losses[i] <- sum((log(exp_data$sensitivity) - log(predicted))^2)
}

par(mfrow = c(1, 3))

# Panel 1: The loss landscape
plot(A_values, losses, type = "l", lwd = 2, col = "blue",
     xlab = "Parameter A", ylab = "Loss (SSE)",
     main = "Panel 1: Loss Landscape\n(for one parameter)")
best_A <- A_values[which.min(losses)]
abline(v = best_A, col = "red", lty = 2, lwd = 2)
points(best_A, min(losses), pch = 19, col = "red", cex = 2)
text(best_A + 30, min(losses) + 0.5, sprintf("Best A = %.0f", best_A), col = "red")

# Panel 2: Show the optimization steps
# Simulate gradient descent
set.seed(42)
A_current <- 150  # Start far from optimum
learning_rate <- 50
A_history <- A_current
loss_history <- sum((log(exp_data$sensitivity) -
                    log(A_current * (exp_data$frequency / fixed_f_peak)^fixed_s1 *
                    exp(-(exp_data$frequency / fixed_f_cut)^fixed_s2)))^2)

for (step in 1:10) {
  # Compute gradient (numerical approximation)
  eps <- 1
  loss_plus <- sum((log(exp_data$sensitivity) -
                   log((A_current + eps) * (exp_data$frequency / fixed_f_peak)^fixed_s1 *
                   exp(-(exp_data$frequency / fixed_f_cut)^fixed_s2)))^2)
  loss_minus <- sum((log(exp_data$sensitivity) -
                    log((A_current - eps) * (exp_data$frequency / fixed_f_peak)^fixed_s1 *
                    exp(-(exp_data$frequency / fixed_f_cut)^fixed_s2)))^2)
  gradient <- (loss_plus - loss_minus) / (2 * eps)

  # Update A (move in direction that decreases loss)
  A_current <- A_current - learning_rate * sign(gradient)
  A_current <- max(100, min(500, A_current))  # Stay in bounds

  # Record history
  A_history <- c(A_history, A_current)
  current_loss <- sum((log(exp_data$sensitivity) -
                      log(A_current * (exp_data$frequency / fixed_f_peak)^fixed_s1 *
                      exp(-(exp_data$frequency / fixed_f_cut)^fixed_s2)))^2)
  loss_history <- c(loss_history, current_loss)
}

plot(A_values, losses, type = "l", lwd = 2, col = "lightblue",
     xlab = "Parameter A", ylab = "Loss (SSE)",
     main = "Panel 2: Optimization Steps\n(walking downhill)")
# Draw the path
for (i in 1:(length(A_history)-1)) {
  arrows(A_history[i], loss_history[i],
         A_history[i+1], loss_history[i+1],
         col = "red", lwd = 2, length = 0.1)
}
points(A_history, loss_history, pch = 19, col = "red", cex = 1.5)
text(A_history[1], loss_history[1] + 1, "Start", col = "red")
text(A_history[length(A_history)], loss_history[length(loss_history)] - 0.5,
     "End", col = "red")

# Panel 3: Loss decreasing over iterations
plot(0:(length(loss_history)-1), loss_history, type = "b", pch = 19, col = "red",
     xlab = "Iteration", ylab = "Loss (SSE)",
     main = "Panel 3: Loss vs Iteration\n(should decrease!)")

cat("Optimization is like walking downhill:\n")
cat("  1. Look around to see which direction is 'downhill' (compute gradient)\n")
cat("  2. Take a step in that direction (update parameters)\n")
cat("  3. Repeat until you reach the bottom (loss stops decreasing)\n")
```

```{r optim-2d-landscape}
#| fig-cap: "With 2 parameters, we search a 2D landscape"
#| fig-width: 8
#| fig-height: 6

# Now show 2-parameter optimization: A and f_cut
# This is a 2D loss landscape (like a terrain map)

A_range <- seq(150, 400, length.out = 30)
fcut_range <- seq(10, 30, length.out = 30)

# Compute loss for each combination
loss_grid <- matrix(NA, nrow = length(A_range), ncol = length(fcut_range))
for (i in seq_along(A_range)) {
  for (j in seq_along(fcut_range)) {
    A <- A_range[i]
    f_cut <- fcut_range[j]
    predicted <- A * (exp_data$frequency / fixed_f_peak)^fixed_s1 *
                 exp(-(exp_data$frequency / f_cut)^fixed_s2)
    loss_grid[i, j] <- sum((log(exp_data$sensitivity) - log(predicted))^2)
  }
}

# Find the minimum
min_idx <- which(loss_grid == min(loss_grid), arr.ind = TRUE)
best_A_2d <- A_range[min_idx[1]]
best_fcut_2d <- fcut_range[min_idx[2]]

par(mfrow = c(1, 2))

# Contour plot (top view of the landscape)
contour(A_range, fcut_range, loss_grid, nlevels = 20,
        xlab = "A (amplitude)", ylab = "f_cut (cutoff frequency)",
        main = "2D Loss Landscape (Top View)\nContour = same loss")
points(best_A_2d, best_fcut_2d, pch = 4, col = "red", cex = 3, lwd = 3)
text(best_A_2d + 20, best_fcut_2d, "Minimum!", col = "red")

# Simulate 2D optimization path
A_curr <- 180
fcut_curr <- 25
path_A <- A_curr
path_fcut <- fcut_curr

for (step in 1:15) {
  # Numerical gradient for A
  eps <- 1
  pred_plus <- A_curr + eps
  pred_base <- A_curr
  loss_plus_A <- sum((log(exp_data$sensitivity) -
                     log(pred_plus * (exp_data$frequency / fixed_f_peak)^fixed_s1 *
                     exp(-(exp_data$frequency / fcut_curr)^fixed_s2)))^2)
  loss_base <- sum((log(exp_data$sensitivity) -
                   log(pred_base * (exp_data$frequency / fixed_f_peak)^fixed_s1 *
                   exp(-(exp_data$frequency / fcut_curr)^fixed_s2)))^2)
  grad_A <- (loss_plus_A - loss_base) / eps

  # Numerical gradient for f_cut
  loss_plus_fcut <- sum((log(exp_data$sensitivity) -
                        log(A_curr * (exp_data$frequency / fixed_f_peak)^fixed_s1 *
                        exp(-(exp_data$frequency / (fcut_curr + eps))^fixed_s2)))^2)
  grad_fcut <- (loss_plus_fcut - loss_base) / eps

  # Update
  A_curr <- A_curr - 5 * sign(grad_A)
  fcut_curr <- fcut_curr - 0.5 * sign(grad_fcut)

  path_A <- c(path_A, A_curr)
  path_fcut <- c(path_fcut, fcut_curr)
}

# Add path to contour plot
lines(path_A, path_fcut, col = "red", lwd = 2)
points(path_A, path_fcut, pch = 19, col = "red", cex = 0.8)
points(path_A[1], path_fcut[1], pch = 19, col = "blue", cex = 2)
text(path_A[1] - 15, path_fcut[1], "Start", col = "blue")

# 3D surface plot
persp(A_range, fcut_range, loss_grid,
      theta = 35, phi = 25,
      xlab = "A", ylab = "f_cut", zlab = "Loss",
      main = "3D Loss Landscape\n(find the valley!)",
      col = "lightblue", shade = 0.5)

cat("\nWith 5 parameters, optim() searches a 5-dimensional landscape!\n")
cat("Same principle: walk downhill until you reach the bottom.\n")
```

::: {.callout-tip title="What optim() Returns"}
``` r
fit <- optim(par, fn, ...)
```

| Component         | Description                              |
|-------------------|------------------------------------------|
| `fit$par`         | Best parameter values found              |
| `fit$value`       | Final loss (should be small)             |
| `fit$convergence` | 0 = success, other = problem             |
| `fit$counts`      | How many times it evaluated the function |
:::

```{r csf-fit-model}
# We minimize sum of squared errors in LOG space

# Define the loss function
csf_loss <- function(params, data) {
  A <- params[1]
  f_peak <- params[2]
  s1 <- params[3]
  f_cut <- params[4]
  s2 <- params[5]

  # Predicted sensitivity
  predicted <- A * (data$frequency / f_peak)^s1 * exp(-(data$frequency / f_cut)^s2)

  # Sum of squared errors in log space
  # (log space because sensitivity spans orders of magnitude)
  sse <- sum((log(data$sensitivity) - log(predicted))^2)
  return(sse)
}

# Starting values (initial guesses) - as a vector
start_params <- c(A = 200, f_peak = 3, s1 = 0.8, f_cut = 15, s2 = 1.5)

# Fit using optim() with L-BFGS-B method (allows parameter bounds)
fit <- optim(
  par = start_params,
  fn = csf_loss,
  data = exp_data,
  method = "L-BFGS-B",
  lower = c(10, 0.5, 0.1, 5, 0.5),   # Lower bounds (sensible constraints)
  upper = c(1000, 20, 3, 50, 5)      # Upper bounds
)

# Extract fitted parameters
fitted_params <- fit$par
names(fitted_params) <- c("A", "f_peak", "s1", "f_cut", "s2")

cat("=== FITTING RESULTS ===\n\n")

# Check convergence (0 = success)
if (fit$convergence == 0) {
  cat("Optimization converged successfully!\n\n")
} else {
  cat("Warning: Optimization may not have converged (code:", fit$convergence, ")\n\n")
}

cat("Fitted parameters:\n")
for (name in names(fitted_params)) {
  cat("  ", name, "=", round(fitted_params[name], 2), "\n")
}

cat("\nTrue parameters (for comparison):\n")
cat("   A =", true_A, "\n")
cat("   f_peak =", true_f_peak, "\n")
cat("   s1 =", true_s1, "\n")
cat("   f_cut =", true_f_cut, "\n")
cat("   s2 =", true_s2, "\n")
```

**Step 4: Evaluate the fit**

```{r csf-evaluate-fit}
# Compute R-squared
predicted <- csf_model(exp_data$frequency,
                       fitted_params["A"],
                       fitted_params["f_peak"],
                       fitted_params["s1"],
                       fitted_params["f_cut"],
                       fitted_params["s2"])

SS_res <- sum((log(exp_data$sensitivity) - log(predicted))^2)
SS_tot <- sum((log(exp_data$sensitivity) - mean(log(exp_data$sensitivity)))^2)
R_squared <- 1 - SS_res / SS_tot

cat("Model fit quality:\n")
cat("  R² =", round(R_squared, 4), "\n")
cat("  (1.0 = perfect fit, >0.95 = excellent)\n")
```

**Step 5: Plot data with fitted curve**

```{r csf-plot-fit}
#| fig-width: 8
#| fig-height: 5

# Generate smooth curve from fitted model
freq_smooth <- seq(0.5, 50, length.out = 200)
fitted_curve <- csf_model(freq_smooth,
                          fitted_params["A"],
                          fitted_params["f_peak"],
                          fitted_params["s1"],
                          fitted_params["f_cut"],
                          fitted_params["s2"])

# Also show the true curve for comparison
true_curve <- csf_model(freq_smooth, true_A, true_f_peak, true_s1, true_f_cut, true_s2)

# Create plot data
plot_data <- data.frame(
  frequency = rep(freq_smooth, 2),
  sensitivity = c(fitted_curve, true_curve),
  curve = rep(c("Fitted", "True"), each = length(freq_smooth))
)

ggplot() +
  # Data points
  geom_point(data = exp_data, aes(frequency, sensitivity),
             size = 4, color = "red", shape = 19) +
  # Fitted curve
  geom_line(data = subset(plot_data, curve == "Fitted"),
            aes(frequency, sensitivity, color = "Fitted"),
            linewidth = 1.5) +
  # True curve (dashed)
  geom_line(data = subset(plot_data, curve == "True"),
            aes(frequency, sensitivity, color = "True"),
            linewidth = 1, linetype = "dashed") +
  scale_color_manual(values = c("Fitted" = "blue", "True" = "gray50")) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = paste0("CSF Model Fit (R² = ", round(R_squared, 3), ")"),
       subtitle = "Red points = data, Blue = fitted model, Gray dashed = true curve",
       x = "Spatial frequency (cpd)",
       y = "Contrast sensitivity",
       color = "Curve") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

::: callout-tip
## Fitting Tips for Real Data

1.  **Starting values matter**: Bad initial guesses can lead to poor fits. Use visual inspection to estimate starting values.

2.  **Fit in log space**: Sensitivity spans orders of magnitude. Fitting `log(sensitivity)` gives equal weight to all data points.

3.  **Check residuals**: Plot residuals vs. frequency — they should be random, not systematic.

4.  **Consider simpler models**: Sometimes 3-4 parameters are enough. More parameters = risk of overfitting.

5.  **Bootstrap for confidence intervals**: Resample your data and refit to get parameter uncertainty.
:::

**Good fit indicators:** - Residuals are **randomly scattered** around 0 (no pattern) - Residuals are **approximately normal** (bell-shaped histogram)

------------------------------------------------------------------------

### 1.5.4 Computational Observers

The ideal observer we built uses a simplified model. What happens when we add more realistic details?

```{r computational-observer}
#| fig-cap: "Building up from ideal to computational observer"
#| fig-width: 10
#| fig-height: 5

# Simulate how different factors affect the CSF
freq <- seq(0.5, 50, length.out = 100)

# Ideal observer (Poisson-limited)
ideal_csf <- 400 * (freq / 4)^0.9 * exp(-(freq / 25)^1.5)

# Add realistic factors one by one:
# 1. More accurate optics (Thibos et al., 2002) - modest effect
optics_csf <- 380 * (freq / 4)^0.9 * exp(-(freq / 22)^1.5)

# 2. Trained linear classifier (not optimal) - reduces sensitivity
classifier_csf <- 300 * (freq / 4)^0.9 * exp(-(freq / 22)^1.5)

# 3. Fixational eye movements - further reduction
fixation_csf <- 250 * (freq / 4)^0.8 * exp(-(freq / 20)^1.4)

# 4. Photocurrent transformation - brings high-freq in line with human
photocurrent_csf <- 280 * (freq / 4)^0.7 * exp(-(freq / 18)^1.4)

# Human data (approximate)
human_csf <- 280 * (freq / 4)^0.9 * exp(-(freq / 18)^1.4)

par(mfrow = c(1, 2))

# Panel 1: Building up the model
plot(freq, ideal_csf, type = "l", log = "xy", col = "gray70", lwd = 2,
     xlab = "Spatial Frequency (cpd)", ylab = "Sensitivity",
     main = "Adding Realistic Factors", ylim = c(1, 700))
lines(freq, optics_csf, col = "red", lwd = 2)
lines(freq, classifier_csf, col = "orange", lwd = 2)
lines(freq, fixation_csf, col = "green", lwd = 2)
lines(freq, photocurrent_csf, col = "blue", lwd = 2)
lines(freq, human_csf, col = "black", lwd = 3, lty = 2)

legend("bottomright",
       c("Ideal (Poisson)", "+ Better optics", "+ Linear classifier",
         "+ Eye movements", "+ Photocurrent", "Human data"),
       col = c("gray70", "red", "orange", "green", "blue", "black"),
       lwd = c(2,2,2,2,2,3), lty = c(1,1,1,1,1,2), cex = 0.7)

# Panel 2: Summary of effects
effects <- data.frame(
  Factor = c("Ideal\n(Poisson)", "Realistic\noptics", "Linear\nclassifier",
             "Eye\nmovements", "Photo-\ncurrent"),
  Sensitivity = c(400, 380, 300, 250, 280),
  x = 1:5
)

barplot(effects$Sensitivity, names.arg = effects$Factor, col = c("gray70", "red", "orange", "green", "blue"),
        main = "Peak Sensitivity After Each Factor",
        ylab = "Peak Sensitivity", ylim = c(0, 450))
abline(h = 280, lty = 2, col = "black", lwd = 2)
text(5.5, 290, "Human", col = "black")

cat("Key insight: Combining principles + detailed computation\n")
cat("accounts for both SHAPE and ABSOLUTE LEVEL of human CSF.\n")
```

::: {.callout-note title="What Each Factor Does"}
| Factor | Effect on CSF |
|----|----|
| **Better optics** (Thibos et al.) | Modest effect - more accurate but similar shape |
| **Linear classifier** | Reduces absolute sensitivity (not optimal like ideal) |
| **Fixational eye movements** | Further reduces performance (image motion) |
| **Photocurrent transformation** | Gain control + noise brings high-freq in line with human |
:::

------------------------------------------------------------------------

### 1.5.5 Image Reconstruction

So far we've focused on **thresholds** (can you detect it?). But what about **suprathreshold vision** - what do clearly visible things look like?

```{r image-reconstruction-concept}
#| fig-cap: "Image reconstruction: From cone excitations back to image"
#| fig-width: 10
#| fig-height: 4

par(mfrow = c(1, 3), mar = c(2, 2, 3, 1))

# Create a simple "image" (2D Gaussian blob)
set.seed(42)
n <- 32
x <- seq(-2, 2, length.out = n)
y <- seq(-2, 2, length.out = n)
grid <- expand.grid(x = x, y = y)

# Original image: Gaussian blob
original <- matrix(exp(-0.5 * (grid$x^2 + grid$y^2)), n, n)

image(x, y, original, col = heat.colors(100), main = "Original Image\n(Scene)",
      xlab = "", ylab = "", axes = FALSE)
box()

# Cone excitations: blurred + sampled + noisy
# Blur (simple convolution approximation)
blur_kernel <- matrix(c(1,2,1,2,4,2,1,2,1)/16, 3, 3)

blurred <- original
for (i in 2:(n-1)) {
  for (j in 2:(n-1)) {
    blurred[i,j] <- sum(original[(i-1):(i+1), (j-1):(j+1)] * blur_kernel)
  }
}

# Add Poisson-like noise (scaled)
cone_excitations <- blurred + matrix(rnorm(n*n, 0, 0.05), n, n)
cone_excitations[cone_excitations < 0] <- 0

image(x, y, cone_excitations, col = heat.colors(100),
      main = "Cone Excitations\n(Blurred + Noisy)",
      xlab = "", ylab = "", axes = FALSE)
box()

# Reconstruction (simplified - just smoothed version)
# In reality this uses Bayesian inference with natural image prior
reconstructed <- cone_excitations
for (iter in 1:3) {
  for (i in 2:(n-1)) {
    for (j in 2:(n-1)) {
      # Simple smoothing (like applying a prior that nearby pixels are similar)
      reconstructed[i,j] <- 0.5 * cone_excitations[i,j] +
                           0.5 * mean(reconstructed[(i-1):(i+1), (j-1):(j+1)])
    }
  }
}

image(x, y, reconstructed, col = heat.colors(100),
      main = "Reconstruction\n(Bayesian estimate)",
      xlab = "", ylab = "", axes = FALSE)
box()

cat("Forward: Image → Blur → Sample → Add noise → Cone excitations\n")
cat("Inverse: Cone excitations → Bayesian inference → Reconstructed image\n")
```

::: {.callout-important title="The Reconstruction Equation"}
**Forward model:** $$\bar{\mathbf{y}} = \mathbf{R}\mathbf{x}$$

Where $\mathbf{R}$ encodes optics + cone sampling.

**Likelihood:** Poisson noise $$p(\mathbf{y}|\mathbf{x}) = \prod_i \text{Poisson}(y_i | [\mathbf{R}\mathbf{x}]_i)$$

**Prior:** Natural images have correlations - Nearby pixels tend to be similar (spatial correlation) - Nearby wavelengths tend to be similar (spectral correlation)

**Posterior:** Combine likelihood × prior to get best estimate of original image!
:::

```{r reconstruction-eccentricity}
#| fig-cap: "Reconstruction quality depends on retinal location"
#| fig-width: 10
#| fig-height: 4

# Simulate reconstruction at different eccentricities
# Key factor: cone density decreases with eccentricity

par(mfrow = c(1, 4), mar = c(2, 1, 3, 1))

# Original
set.seed(123)
n <- 32
original <- matrix(0, n, n)
# Simple pattern: vertical stripes
for (i in 1:n) {
  original[, i] <- sin(2 * pi * i / 8) * 0.5 + 0.5
}

image(original, col = gray.colors(100), main = "Original",
      axes = FALSE, asp = 1)
box()

# Function to simulate reconstruction at different eccentricities
simulate_recon <- function(original, cone_spacing, noise_level) {
  n <- nrow(original)
  # Subsample (fewer cones)
  sampled <- original
  if (cone_spacing > 1) {
    sampled <- matrix(0, n, n)
    for (i in seq(1, n, by = cone_spacing)) {
      for (j in seq(1, n, by = cone_spacing)) {
        val <- original[min(i, n), min(j, n)]
        # Fill in the region
        i_end <- min(i + cone_spacing - 1, n)
        j_end <- min(j + cone_spacing - 1, n)
        sampled[i:i_end, j:j_end] <- val
      }
    }
  }
  # Add noise
  sampled + matrix(rnorm(n*n, 0, noise_level), n, n)
}

# 1° eccentricity (high cone density)
recon_1deg <- simulate_recon(original, cone_spacing = 1, noise_level = 0.05)
image(recon_1deg, col = gray.colors(100), main = "1° eccentricity\n(fovea - dense cones)",
      axes = FALSE, asp = 1)
box()

# 10° eccentricity
recon_10deg <- simulate_recon(original, cone_spacing = 2, noise_level = 0.1)
image(recon_10deg, col = gray.colors(100), main = "10° eccentricity\n(fewer cones)",
      axes = FALSE, asp = 1)
box()

# 18° eccentricity
recon_18deg <- simulate_recon(original, cone_spacing = 3, noise_level = 0.15)
image(recon_18deg, col = gray.colors(100), main = "18° eccentricity\n(sparse cones)",
      axes = FALSE, asp = 1)
box()

cat("Fovea (1°): Dense cones → good reconstruction\n")
cat("Periphery (18°): Sparse cones → more distortion, but still recognizable!\n")
cat("\nThe PRIOR helps fill in missing information from sparse sampling.\n")
```

------------------------------------------------------------------------

### 1.5.6 Optimizing Sensory Measurements

A deeper question: **Why does the visual system measure what it measures?**

```{r optimal-measurements}
#| fig-cap: "Why local derivatives might be optimal measurements"
#| fig-width: 10
#| fig-height: 4

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

# Natural images have 1/f power spectrum
# This means most information is in LOW frequencies
# But we also need edges (HIGH frequencies)

freq <- seq(0.1, 50, length.out = 100)

# Natural image power spectrum (1/f²)
natural_power <- 1 / freq^2
natural_power <- natural_power / max(natural_power)

plot(freq, natural_power, type = "l", log = "xy", lwd = 2, col = "blue",
     xlab = "Spatial Frequency", ylab = "Power",
     main = "Natural Image Statistics\n(1/f² power spectrum)")
text(10, 0.1, "Most power at\nlow frequencies", col = "blue", cex = 0.8)

# Derivative filter boosts high frequencies
# d/dx in frequency domain = multiplication by frequency
derivative_response <- freq
derivative_response <- derivative_response / max(derivative_response)

# Combined: what derivatives "see" in natural images
combined <- natural_power * derivative_response^2

plot(freq, derivative_response, type = "l", log = "xy", lwd = 2, col = "red",
     xlab = "Spatial Frequency", ylab = "Response",
     main = "Derivative Filter\n(edge detector)")
text(5, 0.05, "Boosts high\nfrequencies", col = "red", cex = 0.8)

# The result is roughly flat - uniform information across frequencies!
plot(freq, combined, type = "l", log = "xy", lwd = 2, col = "purple",
     xlab = "Spatial Frequency", ylab = "Power × Response²",
     main = "Natural Images × Derivative\n= roughly uniform!",
     ylim = c(0.001, 1))
abline(h = 0.01, lty = 2, col = "gray")
text(5, 0.02, "Whitened!\nUniform information", col = "purple", cex = 0.8)

cat("Natural images: most power at low frequencies (1/f²)\n")
cat("Derivative filters: boost high frequencies\n")
cat("Combined: roughly uniform information across frequencies!\n")
cat("\nThis is called 'whitening' - it's efficient coding.\n")
```

::: {.callout-tip title="Why Derivatives Are Optimal"}
**The Bayesian optimization framework asks:** 1. What is the prior distribution of scenes? → Natural images (1/f² spectrum) 2. What measurements are feasible? → Local filters at each location 3. What is the cost of errors? → Reconstruction error 4. What are the energetic costs? → Metabolic limits on neurons

**Answer:** Local derivatives (edges, gradients) are close to optimal because they: - "Whiten" the natural image statistics - Extract maximum information per neuron - Explain why neurons are edge-selective, orientation-selective, motion-selective
:::

------------------------------------------------------------------------

## 1.6 Summary

::: {.callout-important title="Key Takeaways"}
**The Forward Calculation (§1.2-1.4):** $$\text{Scene} \xrightarrow{\text{optics}} \text{Retinal Image} \xrightarrow{\text{cones}} \text{Excitations} \xrightarrow{\text{noise}} \text{Neural Signal}$$

**The Inverse Problem (§1.5):** $$p(\text{scene}|\text{signal}) \propto p(\text{signal}|\text{scene}) \times p(\text{scene})$$

**Core insight:** Perception is Bayesian inference - we perceive the most likely scene given our noisy sensory signals.
:::

```{r summary-forward}
#| fig-cap: "Summary: The forward calculation pipeline"
#| fig-width: 10
#| fig-height: 3

par(mar = c(1, 1, 2, 1))
plot(NULL, xlim = c(0, 12), ylim = c(0, 4), axes = FALSE, xlab = "", ylab = "",
     main = "Forward Calculation: Scene → Cone Excitations")

# Boxes
box_y <- 2
box_h <- 1.5
colors <- c("lightyellow", "lightblue", "lightgreen", "lavender")
labels <- c("Light Field\nLF(x,y,z,α,β,λ)", "Retinal Image\nE(rx,ry,λ)",
            "Cone Excitations\nρk = ∫ E(λ)Sk(λ)dλ", "Neural Signal\n+ Poisson noise")
x_positions <- c(0.5, 3.5, 6.5, 9.5)

for (i in 1:4) {
  rect(x_positions[i], box_y - box_h/2, x_positions[i] + 2, box_y + box_h/2,
       col = colors[i], border = "gray40", lwd = 2)
  text(x_positions[i] + 1, box_y, labels[i], cex = 0.75)
}

# Arrows with labels
arrows(2.5, box_y, 3.5, box_y, lwd = 2)
arrows(5.5, box_y, 6.5, box_y, lwd = 2)
arrows(8.5, box_y, 9.5, box_y, lwd = 2)

text(3, box_y + 0.9, "Optics\n(PSF)", cex = 0.7, col = "blue")
text(6, box_y + 0.9, "Cones\n(Linear)", cex = 0.7, col = "blue")
text(9, box_y + 0.9, "Noise\n(Poisson)", cex = 0.7, col = "blue")

# Section references
text(3, box_y - 1.1, "§1.2", col = "gray50", cex = 0.8)
text(6, box_y - 1.1, "§1.3", col = "gray50", cex = 0.8)
text(9, box_y - 1.1, "§1.4", col = "gray50", cex = 0.8)
```

| Principle          | Application                          |
|--------------------|--------------------------------------|
| **Linear algebra** | Spectral radiance → cone excitations |
| **Convolution**    | Optical blur (PSF)                   |
| **Poisson noise**  | Photon counting variability          |
| **Bayes' rule**    | Prior × Likelihood = Posterior       |

**PRINCIPLES + DATA + COMPUTATION = UNDERSTANDING**

------------------------------------------------------------------------

## References

Wandell, B. & Brainard, D. (2024). Principles and consequences of the initial visual encoding. In J. Batchelder et al. (Eds.), *New handbook of mathematical psychology* (Vol. 3). Cambridge University Press.
