---
title: "Initial Visual Encoding"
subtitle: "Handbook Ch.1 - Wandell & Brainard"
date: 2026-01-23
format: html
execute:
  echo: true
---

```{r setup}
#| echo: false
#| message: false
library(ggplot2)
library(tidyr)
```

Vision science uses mathematics to communicate across disciplines. The initial visual encoding describes how light from a scene becomes neural signals - from physics (light) through optics (eye) to biology (photoreceptors).

## 1.2 Scene to Retinal Image

### Light Field

Leonardo da Vinci noticed that a pinhole camera forms an image wherever you place it - meaning light rays must be present everywhere, traveling in all directions. Gershun (1939) formalized this as the **light field**:

$$LF(x, y, z, \alpha, \beta, \lambda, \rho)$$

Each ray is described by:

-   Position: $(x, y, z)$
-   Direction: $(\alpha, \beta)$
-   Wavelength: $\lambda$
-   Polarization: $\rho$

This is a 7-dimensional function!

### Incident Light Field

The eye only captures rays entering the pupil - the **incident light field**:

$$ILF(u, v, \alpha, \beta, \lambda, \rho, t)$$

where $(u, v)$ is position at the pupil aperture and $t$ is time.

### Plenoptic Function

Adelson & Bergen (1991) simplified further for human vision. The eye focuses rays onto the retina, so direction is determined by retinal position $(r_x, r_y)$. Ignoring polarization (unimportant for perception):

$$E(r_x, r_y, \lambda, t)$$

This is the **retinal spectral irradiance** - power per unit area as a function of wavelength at each retinal location.

```{r light-field-reduction}
# The eye dramatically reduces information dimensionality
dimensions <- data.frame(
  representation = c("Light Field", "Incident Light Field", "Retinal Image"),
  dimensions = c(7, 7, 4),
  parameters = c("x,y,z,α,β,λ,ρ", "u,v,α,β,λ,ρ,t", "rx,ry,λ,t")
)
dimensions
```

The key insight: many different light fields produce the **same** retinal image. Information is inevitably lost.

### Simulating a Light Field (Toy Model)

> This is a simplified **toy model** for learning - not from a published source.

Let's make the abstract light field concrete by building a simple 2D scene and querying it. We'll work in 2D (dropping z) with one direction angle α (dropping β) to keep things visualizable.

**Our scene**: Three colored rectangles at different positions along a wall.

```{r define-scene}
# Define a simple scene: colored rectangles on a wall at y = 10
# Each object has: x_min, x_max, and a spectrum (simplified to peak wavelength)
scene <- list(
  list(x_min = -3, x_max = -1, peak_wavelength = 620),  # red object
  list(x_min = 1, x_max = 3, peak_wavelength = 530),    # green object
  list(x_min = 5, x_max = 7, peak_wavelength = 460)     # blue object
)
wall_y <- 10
```

```{r plot-scene}
#| fig-width: 7
#| fig-height: 5

# Visualize the scene from bird's eye view
scene_plot <- ggplot() +
  # Draw the wall
  geom_segment(aes(x = -5, xend = 9, y = wall_y, yend = wall_y),
               color = "gray50", linewidth = 1) +
  # Draw the colored objects on the wall
  geom_segment(aes(x = -3, xend = -1, y = wall_y, yend = wall_y),
               color = "red", linewidth = 4) +
  geom_segment(aes(x = 1, xend = 3, y = wall_y, yend = wall_y),
               color = "green3", linewidth = 4) +
  geom_segment(aes(x = 5, xend = 7, y = wall_y, yend = wall_y),
               color = "blue", linewidth = 4) +
  # Draw the observer
  geom_point(aes(x = 0, y = 0), size = 4, color = "black") +
  annotate("text", x = 0, y = -0.8, label = "Observer (0, 0)", size = 3.5) +
  # Draw rays from observer to center of each object
  geom_segment(aes(x = 0, xend = -2, y = 0, yend = wall_y),
               color = "red", linetype = "dashed", alpha = 0.6) +
  geom_segment(aes(x = 0, xend = 2, y = 0, yend = wall_y),
               color = "green3", linetype = "dashed", alpha = 0.6) +
  geom_segment(aes(x = 0, xend = 6, y = 0, yend = wall_y),
               color = "blue", linetype = "dashed", alpha = 0.6) +
  # Draw the "straight ahead" reference line
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = wall_y),
               color = "gray30", linetype = "dotted") +
  annotate("text", x = 0.3, y = 5, label = "α = 0°", size = 3, angle = 90) +
  # Label the angles
  annotate("text", x = -2, y = wall_y + 0.6, label = "α ≈ -11°",
           size = 3, color = "red") +
  annotate("text", x = 2, y = wall_y + 0.6, label = "α ≈ +11°",
           size = 3, color = "green4") +
  annotate("text", x = 6, y = wall_y + 0.6, label = "α ≈ +31°",
           size = 3, color = "blue") +
  # Formatting
  coord_fixed(ratio = 1) +
  xlim(-5, 9) + ylim(-2, 12) +
  labs(title = "Bird's Eye View of the Scene",
       subtitle = "Observer at origin, wall at y = 10, three colored objects",
       x = "x position", y = "y position") +
  theme_minimal()

scene_plot
```

**Understanding the angles:**

The angle α tells you which direction you're looking:

-   **α = 0** means looking straight ahead (along the y-axis)
-   **α \< 0** (negative) means looking LEFT
-   **α \> 0** (positive) means looking RIGHT

The angle to each object depends on geometry. From (0, 0), to hit the center of the green object at (2, 10):

$$\alpha = \arctan\left(\frac{x_{object}}{y_{wall}}\right) = \arctan\left(\frac{2}{10}\right) \approx 11.3°$$

```{r angle-calculation}
# Calculate the viewing angle to each object's center
object_centers <- c(-2, 2, 6)  # x-coordinates of object centers
angles_to_objects <- atan(object_centers / wall_y) * 180 / pi  # convert to degrees

data.frame(
  object = c("Red", "Green", "Blue"),
  center_x = object_centers,
  angle_degrees = round(angles_to_objects, 1)
)
```

**The light field function:**

```{r light-field-function}
light_field <- function(x, y, alpha, lambda) {
  if (y >= wall_y) return(0)

  # Ray trace: where does ray hit the wall?
  hit_x <- x + (wall_y - y) * tan(alpha)

  # Check if we hit an object
  for (obj in scene) {
    if (hit_x >= obj$x_min && hit_x <= obj$x_max) {
      # Return spectrum (Gaussian centered at object's peak wavelength)
      return(exp(-0.5 * ((lambda - obj$peak_wavelength) / 30)^2))
    }
  }
  return(0.05)  # empty wall
}
```

> **Reality vs Toy Model:** Real radiance varies due to light source distance, surface angle, texture, and shadows. Our toy model only uses spectrum match (Gaussian).

**Scan across directions** from (0,0):

```{r query-directions}
#| fig-width: 6
#| fig-height: 3

angles <- seq(-0.6, 0.6, length.out = 200)
results <- data.frame(
  angle = rep(angles * 180/pi, 3),
  radiance = c(
    sapply(angles, function(a) light_field(0, 0, a, 620)),
    sapply(angles, function(a) light_field(0, 0, a, 530)),
    sapply(angles, function(a) light_field(0, 0, a, 460))
  ),
  color = rep(c("red", "green", "blue"), each = 200)
)

ggplot(results, aes(angle, radiance, color = color)) +
  geom_line(linewidth = 1) +
  scale_color_identity() +
  labs(title = "Light field: scanning left to right",
       x = "Angle (degrees)", y = "Radiance") +
  theme_minimal()
```

Each peak = one object. The function answers: "Looking in direction α, how much light of wavelength λ do I see?"

### From Light Field to Cone Response

The eye compresses the light field in two steps:

1.  **Direction → Retinal position**: The lens maps viewing angle to a spot on the retina
2.  **Many wavelengths → 3 cone values**: Each cone type sums up light weighted by its sensitivity

```{r compression-step1}
#| fig-width: 5
#| fig-height: 3

directions <- seq(-0.4, 0.7, length.out = 30)
wavelengths <- seq(400, 700, by = 10)

# Pick a direction (change this to see different objects!)
alpha <- -0.1

# What spectrum arrives at this direction?
spectrum_at_alpha <- sapply(wavelengths, function(w) light_field(0, 0, alpha, w))

# Determine color based on which object we're looking at
hit_x <- (wall_y - 0) * tan(alpha)
plot_color <- if (hit_x >= -3 && hit_x <= -1) "red" else
              if (hit_x >= 1 && hit_x <= 3) "green3" else
              if (hit_x >= 5 && hit_x <= 7) "blue" else "gray50"
object_name <- if (hit_x >= -3 && hit_x <= -1) "red" else
               if (hit_x >= 1 && hit_x <= 3) "green" else
               if (hit_x >= 5 && hit_x <= 7) "blue" else "wall"

ggplot(data.frame(w = wavelengths, r = spectrum_at_alpha), aes(w, r)) +
  geom_line(color = plot_color, linewidth = 1) +
  geom_area(fill = plot_color, alpha = 0.3) +
  labs(title = paste0("Spectrum at α = ", round(alpha * 180/pi, 1), "°"),
       subtitle = paste0("This spot sees the ", object_name, " object"),
       x = "Wavelength (nm)", y = "Radiance") +
  theme_minimal()
```

```{r compression-step2}
#| fig-width: 5
#| fig-height: 3

# Cone sensitivities (different widths reflect real biology!)
# S-cone: narrow (SD=35), M-cone: medium (SD=45), L-cone: wide (SD=50)
L_sens <- exp(-0.5 * ((wavelengths - 565) / 50)^2)
M_sens <- exp(-0.5 * ((wavelengths - 535) / 45)^2)
S_sens <- exp(-0.5 * ((wavelengths - 445) / 35)^2)

# Cone response = sum(spectrum × sensitivity)
L_resp <- sum(spectrum_at_alpha * L_sens)
M_resp <- sum(spectrum_at_alpha * M_sens)
S_resp <- sum(spectrum_at_alpha * S_sens)

ggplot(data.frame(cone = c("L", "M", "S"), response = c(L_resp, M_resp, S_resp)),
       aes(cone, response, fill = cone)) +
  geom_col() +
  scale_fill_manual(values = c("L" = "red", "M" = "green", "S" = "blue")) +
  labs(title = "Cone responses at this position",
       subtitle = "31 wavelengths → 3 numbers",
       x = "", y = "Response") +
  theme_minimal() + theme(legend.position = "none")
```

M-cone responds most because green object matches M-cone sensitivity!

> **Note:** The different bandwidths (SD = 35, 45, 50) approximate real cone biology. S-cones are narrowest (most selective), L-cones are widest (respond to broader range). This comes from the molecular structure of each cone's photopigment (opsin).

```{r compression-full}
#| fig-width: 6
#| fig-height: 3

# Do this for ALL retinal positions
L_responses <- c()
M_responses <- c()
S_responses <- c()

for (a in directions) {
  # Step 1: Get spectrum at this direction
  spectrum <- c()
  for (w in wavelengths) {
    spectrum <- c(spectrum, light_field(0, 0, a, w))
  }

  # Step 2: Compute cone responses (weighted sum)
  L_responses <- c(L_responses, sum(spectrum * L_sens))
  M_responses <- c(M_responses, sum(spectrum * M_sens))
  S_responses <- c(S_responses, sum(spectrum * S_sens))
}

# Combine into data frame for plotting
cone_responses <- data.frame(
  position = rep(directions * 180/pi, 3),
  response = c(L_responses, M_responses, S_responses),
  cone = rep(c("L", "M", "S"), each = length(directions))
)

ggplot(cone_responses, aes(position, response, color = cone)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("L" = "red", "M" = "green", "S" = "blue")) +
  labs(title = "Cone responses across retina",
       x = "Retinal position (degrees)", y = "Response") +
  theme_minimal()
```

**The compression:** Light field (position × direction × wavelength) → just 3 numbers (L, M, S) per retinal spot. This is the "initial visual encoding."

## 1.3 Mathematical Principles

### Linear Systems

A system $L$ is **linear** if it follows the superposition rule:

$$L(x + y) = L(x) + L(y)$$

This implies **homogeneity**: $L(mx) = mL(x)$ for any scalar $m$.

```{r linear-system}
# Example: a linear transformation (like cone absorption)
L <- function(input) {
  0.6 * input[1] + 0.3 * input[2] + 0.1 * input[3]
}

x <- c(1, 2, 3)
y <- c(0.5, 1.5, 2.5)

# Test superposition: L(x + y) should equal L(x) + L(y)
L_sum <- L(x + y)
sum_L <- L(x) + L(y)

data.frame(
  test = c("L(x + y)", "L(x) + L(y)"),
  value = c(L_sum, sum_L)
)
```

Why is linearity so powerful? If we know how a system responds to a set of **basis inputs**, we can predict its response to **any** input (as a weighted sum of those basis responses).

### Cone Excitations

The transformation from spectral irradiance to cone excitations is linear. For a cone with fundamental $C(\lambda)$:

$$N(r_x, r_y) = \int C(\lambda) \cdot E(r_x, r_y, \lambda) \, d\lambda$$

In discrete form (how we compute it):

$$N(r_x, r_y) = \sum_{\lambda_i} C(\lambda_i) \cdot E(r_x, r_y, \lambda_i) \cdot \Delta\lambda$$

This is just a weighted sum - the inner product of the cone fundamental with the spectrum.

```{r cone-fundamentals}
# Simplified cone fundamentals (Gaussian approximations)
wavelengths <- seq(400, 700, by = 5)

# L-cone peaks ~565nm, M-cone ~535nm, S-cone ~445nm
L_cone <- exp(-0.5 * ((wavelengths - 565) / 50)^2)
M_cone <- exp(-0.5 * ((wavelengths - 535) / 45)^2)
S_cone <- exp(-0.5 * ((wavelengths - 445) / 35)^2)

cone_data <- data.frame(
  wavelength = rep(wavelengths, 3),
  sensitivity = c(L_cone, M_cone, S_cone),
  cone = rep(c("L (long)", "M (medium)", "S (short)"), each = length(wavelengths))
)

ggplot(cone_data, aes(x = wavelength, y = sensitivity, color = cone)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("L (long)" = "red", "M (medium)" = "green", "S (short)" = "blue")) +
  labs(title = "Human Cone Fundamentals (simplified)",
       x = "Wavelength (nm)", y = "Relative Sensitivity") +
  theme_minimal()
```

### Matrix Formulation

With three cone types (L, M, S), we can write cone excitations as matrix multiplication:

$$\mathbf{n}(r_x, r_y) = \mathbf{C} \cdot \mathbf{e}(r_x, r_y)$$

where:

-   $\mathbf{C}$ is a $3 \times w$ matrix (3 cone types, $w$ wavelengths)
-   $\mathbf{e}$ is a $w \times 1$ vector (spectral irradiance at one location)
-   $\mathbf{n}$ is a $3 \times 1$ vector (L, M, S cone excitations)

```{r matrix-formulation}
# Create cone matrix C (3 x w)
C <- rbind(L_cone, M_cone, S_cone)

# Example spectrum: "white" light (equal energy)
e_white <- rep(1, length(wavelengths))

# Compute cone excitations: n = C * e
n_white <- C %*% e_white

data.frame(
  cone = c("L", "M", "S"),
  excitation = round(as.vector(n_white), 2)
)
```

### Color Matching and Metamers

A profound consequence: **different spectra can produce identical cone excitations**. These pairs are called **metamers**.

For two spectra $\mathbf{e}_1$ and $\mathbf{e}_2$ to be metamers:

$$\mathbf{C} \cdot \mathbf{e}_1 = \mathbf{C} \cdot \mathbf{e}_2$$

This means their difference is in the **null space** of $\mathbf{C}$:

$$\mathbf{C} \cdot (\mathbf{e}_2 - \mathbf{e}_1) = \mathbf{0}$$

Wyszecki called vectors in this null space "**metameric blacks**" - adding them to any spectrum creates a metamer.

```{r metamers}
# Create two different spectra that produce the same cone response

# Spectrum 1: narrow yellow (580nm)
e_yellow <- exp(-0.5 * ((wavelengths - 580) / 15)^2)

# Spectrum 2: mixture of red + green
e_red <- exp(-0.5 * ((wavelengths - 620) / 15)^2)
e_green <- exp(-0.5 * ((wavelengths - 540) / 15)^2)

# Find weights to match yellow's cone excitations
n_yellow <- C %*% e_yellow

# We want: w_r * C*e_red + w_g * C*e_green = C*e_yellow
# Solve for weights using least squares
A <- cbind(C %*% e_red, C %*% e_green)
weights <- solve(t(A) %*% A) %*% t(A) %*% n_yellow

e_mixture <- as.numeric(weights[1]) * e_red + as.numeric(weights[2]) * e_green
n_mixture <- C %*% e_mixture

# Compare
data.frame(
  cone = c("L", "M", "S"),
  yellow = round(as.vector(n_yellow), 3),
  red_green_mix = round(as.vector(n_mixture), 3),
  difference = round(as.vector(n_yellow - n_mixture), 6)
)
```

The spectra are physically different, but perceptually identical!

```{r plot-metamers}
#| fig-width: 6
#| fig-height: 3

metamer_data <- data.frame(
  wavelength = rep(wavelengths, 2),
  intensity = c(e_yellow / max(e_yellow), e_mixture / max(e_mixture)),
  spectrum = rep(c("Yellow (narrow)", "Red+Green mix"), each = length(wavelengths))
)

ggplot(metamer_data, aes(x = wavelength, y = intensity, color = spectrum)) +
  geom_line(linewidth = 1) +
  labs(title = "Metameric Spectra",
       subtitle = "Different physics, same perception",
       x = "Wavelength (nm)", y = "Relative Intensity") +
  theme_minimal()
```

This is why displays work! They don't reproduce the original spectrum - they produce a **metamer** using just RGB primaries.

### Noise in Sensory Measurements

Cone excitations are inherently noisy. Two distributions matter:

**Gaussian**: $N(x; \mu, \sigma) = \frac{1}{\sigma\sqrt{2\pi}} \exp\left(-\frac{(x-\mu)^2}{2\sigma^2}\right)$

-   Two parameters: mean $\mu$, variance $\sigma^2$
-   Often used for "signal-independent" noise (same variance everywhere)

**Poisson**: $P(M; \mu) = e^{-\mu} \frac{\mu^M}{M!}$

-   One parameter: $\mu$ (mean = variance)
-   "Signal-dependent" noise: variance scales with signal

Cone excitations follow **Poisson** statistics - this is fundamental.

```{r noise-comparison}
#| fig-width: 7
#| fig-height: 3

# Compare Poisson and Gaussian for different means
mu_values <- c(5, 20, 100)

noise_data <- data.frame()

for (mu in mu_values) {
  x_range <- seq(max(0, mu - 4*sqrt(mu)), mu + 4*sqrt(mu), by = 1)

  noise_data <- rbind(noise_data,
    data.frame(x = x_range,
               prob = dpois(x_range, mu),
               dist = "Poisson",
               mu = paste0("mean = ", mu)),
    data.frame(x = x_range,
               prob = dnorm(x_range, mu, sqrt(mu)),
               dist = "Gaussian",
               mu = paste0("mean = ", mu))
  )
}

ggplot(noise_data, aes(x = x, y = prob, color = dist)) +
  geom_line(linewidth = 0.8) +
  geom_point(data = subset(noise_data, dist == "Poisson"), size = 1) +
  facet_wrap(~mu, scales = "free") +
  labs(title = "Gaussian approximation to Poisson",
       subtitle = "Works well when mean > 10",
       x = "Count", y = "Probability") +
  theme_minimal()
```

The critical difference: for **signal-independent** noise, variance is constant. For **Poisson** (signal-dependent), variance scales with the mean. This matters for dark vs bright regions!

### Image Formation and Shift-Invariance

For a planar image, the retinal image $E$ is formed from the displayed image $I$ through the optics:

$$E(r_x, r_y, \lambda) = \alpha E_1(r_x, r_y, \lambda) + \beta E_2(r_x, r_y, \lambda)$$

This is linear! But what about **shift-invariance**? A system is shift-invariant if shifting the input just shifts the output (without changing its form).

For shift-invariant systems, we can use **convolution**:

$$E(r_x, r_y) = \int \int I(u, v) \cdot P(r_x - u, r_y - v) \, du \, dv$$

where $P(r_x, r_y)$ is the **point spread function** - the retinal image of a single point.

```{r convolution}
#| fig-width: 6
#| fig-height: 3

# Simple 1D demonstration: sharp edge blurred by PSF
x <- 1:100

# Sharp edge (step function)
edge <- ifelse(x > 50, 1, 0)

# Point spread function (Gaussian blur)
psf_width <- 5
psf <- dnorm(seq(-15, 15), 0, psf_width)
psf <- psf / sum(psf)

# Convolve (using R's built-in filter)
blurred <- stats::filter(edge, psf, sides = 2)
blurred[is.na(blurred)] <- edge[is.na(blurred)]

conv_data <- data.frame(
  position = rep(x, 2),
  intensity = c(edge, blurred),
  image = rep(c("Original (display)", "Blurred (retina)"), each = length(x))
)

ggplot(conv_data, aes(x = position, y = intensity, color = image)) +
  geom_line(linewidth = 1) +
  labs(title = "Image formation by convolution",
       subtitle = "E(r) = I(x) * P(r-x)",
       x = "Position", y = "Intensity") +
  theme_minimal()
```

**Important caveat**: The human eye is NOT perfectly shift-invariant:

-   PSF varies with retinal eccentricity
-   PSF varies with wavelength (chromatic aberration)
-   Cone density varies across the retina

This is why computational models (like ISETBio) are essential for accurate predictions.

## Summary of Key Equations

| Equation | Description |
|------------------------------------|------------------------------------|
| $LF(x,y,z,\alpha,\beta,\lambda,\rho)$ | Light field (Eq. 1.1) |
| $E(r_x, r_y, \lambda, t)$ | Retinal spectral irradiance (Eq. 1.3) |
| $L(x+y) = L(x) + L(y)$ | Superposition (linearity) (Eq. 1.4) |
| $N = \int C(\lambda) E(\lambda) d\lambda$ | Cone excitation (Eq. 1.7) |
| $\mathbf{n} = \mathbf{C} \mathbf{e}$ | Matrix form (Eq. 1.9) |
| $\mathbf{C}(\mathbf{e}_2 - \mathbf{e}_1) = \mathbf{0}$ | Metamerism condition (Eq. 1.10) |
| $E = \int I(u,v) P(r-u) du dv$ | Convolution (Eq. 1.13) |

## References

Wandell, B. & Brainard, D. (2024). Principles and consequences of the initial visual encoding. In J. Batchelder et al. (Eds.), *New handbook of mathematical psychology* (Vol. 3). Cambridge University Press.

Adelson, E. H., & Bergen, J. (1991). The plenoptic function and the elements of early vision. In M. Landy & J. Movshon (Eds.), *Computational models of visual processing* (pp. 3-20). MIT Press.

Gershun, A. (1939). The light field. *Journal of Mathematical Physics, 18*(1-4), 51-151.
